---
title: "Multivariate animal model — revised analysis"
author: "Helen Payne (revised)"
date: "2026-05-20"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This document re-implements the multivariate animal-model analysis for the four
*Nemophila menziesii* populations (AC, BB, BO, HR) and addresses the following
issues with the original analysis:

1. **Maternal (dam) random effect** added to separate V_A from maternal effects,
   matching the paternal half-sib design rationale.
2. **Block random effect** added (the field design has 3 blocks per site).
3. **Genetic selection gradient (β_G)** computed via Robertson–Price identity
   from the posterior of G, in addition to the Lande–Arnold phenotypic β.
4. **Total fitness** (viability × fecundity) used for selection gradients, so
   that viability selection is not lost.
5. **Multiple MCMC chains** with Gelman–Rubin R̂ diagnostics.
6. **Random Skewers** computed across the full posterior of G, not posterior
   means only.
7. **Deterministic data cleaning** (no unseeded `sample()` calls).
8. **Function-based structure**: one set of functions, looped over populations.
9. **Sample-size reporting** per population.
10. **Significance-aware r_A plotting** (HPDs distinguished).

A `TEST_RUN` flag at the top switches to a tiny MCMC run for debugging; set it
to `FALSE` to fit the full analysis.

```{r config}
options(rgl.useNULL = TRUE)   # evolqg pulls in rgl; avoid X11 dependency

suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(MCMCglmm)
  library(nadiv)
  library(coda)
  library(evolqg)
  library(broom)
  library(patchwork)
  library(car)        # vif()
})

# evolqg pulls in plyr, which is required by evolqg and cannot be detached.
# plyr exports its own mutate/rename/summarise/arrange that mask dplyr's.
# We use `conflicted` to make the dplyr versions win when both are on the
# search path.
suppressPackageStartupMessages(library(conflicted))
for (fn in c("mutate", "rename", "filter", "select", "summarise", "summarize",
             "arrange", "group_by", "ungroup", "transmute", "count", "tally",
             "distinct", "bind_rows", "bind_cols", "recode")) {
  try(conflicted::conflict_prefer(fn, "dplyr", quiet = TRUE), silent = TRUE)
}

set.seed(20260520)

# Toggle TEST_RUN = TRUE for a fast smoke-test (~1-2 min total).
# Toggle TEST_RUN = FALSE for the publication-quality fit.
TEST_RUN <- TRUE

if (TEST_RUN) {
  N_CHAINS <- 2
  NITT     <- 6000
  BURNIN   <- 1000
  THIN     <- 5
  N_SKEWERS <- 200
  N_POST_DRAWS_FOR_SKEWERS <- 100   # subsample posterior to speed up
} else {
  N_CHAINS <- 3
  NITT     <- 200000
  BURNIN   <- 20000
  THIN     <- 100
  N_SKEWERS <- 2000
  N_POST_DRAWS_FOR_SKEWERS <- 1000
}

# Per-population configuration. spatial_col differs between sites (BB uses
# Location; AC/BO/HR use Transect). Seeds-sown differs between AC (10) and the
# others (8). exclude_blocks: AC includes a "W" watered-treatment block that is
# not part of this study.
#
# Data are loaded from the repo-local data/ directory, populated either by
# R/fetch_dryad.R (first run; downloads from Dryad doi:10.5061/dryad.pvmcvdp1p)
# or manually while the Dryad package is private.
source(here::here("fetch_dryad.R"))
fetch_dryad(c("AC_compiledsheet_full_2022.csv",
              "BB_compiledsheet_full_2022.csv",
              "BO_compiledsheet_full_2022.csv",
              "HR_compiledsheet_full_2022.csv"))

POP_CONFIG <- list(
  AC = list(file = "data/AC_compiledsheet_full_2022.csv",
            spatial = "Transect", seeds_sown = 10, exclude_blocks = "W"),
  BB = list(file = "data/BB_compiledsheet_full_2022.csv",
            spatial = "Location", seeds_sown = 8,  exclude_blocks = character(0)),
  BO = list(file = "data/BO_compiledsheet_full_2022.csv",
            spatial = "Transect", seeds_sown = 8,  exclude_blocks = character(0)),
  HR = list(file = "data/HR_compiledsheet_full_2022.csv",
            spatial = "Transect", seeds_sown = 8,  exclude_blocks = character(0))
)

TRAITS <- c("corolla_z", "SLA_z", "d13C_z", "msm_z", "fec_z")
NT     <- length(TRAITS)
```

# Data preparation

The cleaning is deterministic: when a 1-m segment contains no plant flagged as
`F_plant`, we now select the **first surviving sibling** in row order (rather
than a random plant). This matches the field protocol described in the
manuscript ("if the selected individual died before flowering, a sibling within
the same segment was substituted") and is reproducible.

```{r prep-fn}
prep_population <- function(pop_code, cfg) {
  fpath <- here::here(cfg[["file"]])
  raw <- readr::read_csv(fpath, show_col_types = FALSE)
  spatial <- cfg$spatial

  if (length(cfg$exclude_blocks) > 0) {
    raw <- raw %>% filter(!(Block %in% cfg$exclude_blocks))
  }

  raw <- raw %>%
    mutate(
      Donor     = paste0("Donor_",     pop_code, "_",
                         str_remove(Donor,     paste0("^Donor_", pop_code, "_|^", pop_code, "_"))),
      Recipient = paste0("Recipient_", pop_code, "_",
                         str_remove(Recipient, paste0("^Recipient_", pop_code, "_|^", pop_code, "_"))),
      F_plant   = ifelse(F_plant == TRUE, 1L, 0L),
      F_plant   = ifelse(is.na(F_plant), 0L, F_plant),
      surv_to_flower = ifelse(total_fruits > 0 & surv_to_flower == 0,
                              1, surv_to_flower)
    ) %>%
    rename(spatial_unit = !!sym(spatial)) %>%
    mutate(spatial_unit = suppressWarnings(as.numeric(spatial_unit))) %>%
    filter(!is.na(spatial_unit)) %>%
    mutate(
      sample_ID = str_c(Year, Donor, Recipient, Gen,
                        str_pad(spatial_unit, width = 3, pad = "0"),
                        Plant_ID, sep = "_"),
      sample_ID = str_replace(sample_ID, "^([^_]*)_", "\\1-")
    )

  # Deterministic F_plant assignment: where no plant in a segment is flagged
  # as F_plant, take the first surviving sibling. If none survived, take the
  # first row (it will contribute W=0 to selection analysis).
  raw <- raw %>%
    mutate(sample_prefix = substr(sample_ID, 1, 19)) %>%
    group_by(sample_prefix) %>%
    mutate(
      seg_has_fplant = sum(F_plant) > 0,
      seg_first_surv = which(surv_to_flower == 1)[1],   # NA if none survived
      seg_pick_idx   = ifelse(!is.na(seg_first_surv), seg_first_surv, 1L),
      F_plant        = ifelse(!seg_has_fplant & row_number() == seg_pick_idx,
                              1L, F_plant)
    ) %>%
    ungroup() %>%
    select(-sample_prefix, -seg_has_fplant, -seg_first_surv, -seg_pick_idx)

  # Segment-level survival count (number alive out of seeds_sown), used for
  # the survival glm.
  raw <- raw %>%
    mutate(group = ceiling(row_number() / 8)) %>%
    group_by(group) %>%
    mutate(total_surv_to_flower = sum(surv_to_flower, na.rm = TRUE)) %>%
    ungroup() %>%
    select(-group)

  fp <- raw %>% filter(F_plant == 1L)

  fp <- fp %>%
    mutate(
      animal  = sample_ID,
      dam     = Recipient,
      sire    = Donor,
      block   = factor(Block),
      transect_f = factor(spatial_unit),
      corolla = corolla_diam_mm_SEG,
      SLA     = SLA_SEG,
      d13C    = d13C_SEG,
      msm     = msm_all,
      fec     = est_fecundity,
      surv    = surv_to_flower
    )

  # Center+scale to mean 0 / sd 1 (z-scaling) using only finite values.
  z <- function(x) {
    m <- mean(x, na.rm = TRUE); s <- sd(x, na.rm = TRUE)
    if (!is.finite(s) || s == 0) return(rep(NA_real_, length(x)))
    (x - m) / s
  }
  fp <- fp %>%
    mutate(
      corolla_z = z(corolla),
      SLA_z     = z(SLA),
      d13C_z    = z(d13C),
      msm_z     = z(msm),
      fec_z     = z(fec)
    )

  list(all = raw, fp = fp, pop = pop_code, spatial = spatial)
}

build_pedigree <- function(fp) {
  ped <- tibble(
    animal = as.character(fp$sample_ID),
    dam    = fp$Recipient,
    sire   = fp$Donor
  )
  missing_parents <- setdiff(unique(c(ped$dam, ped$sire)), ped$animal)
  ped2 <- bind_rows(
    ped,
    tibble(animal = missing_parents, dam = NA_character_, sire = NA_character_)
  ) %>% distinct(animal, .keep_all = TRUE)
  ped_ord <- prepPed(as.data.frame(ped2))
  Ainv <- MCMCglmm::inverseA(ped_ord)$Ainv
  list(ped = ped_ord, Ainv = Ainv)
}
```

# Load and summarise

```{r load-all}
pop_data <- purrr::map(names(POP_CONFIG),
                ~ prep_population(.x, POP_CONFIG[[.x]]))
names(pop_data) <- names(POP_CONFIG)

# Sample-size reporting table (item #18 from the review)
sample_sizes <- purrr::map_dfr(pop_data, function(d) {
  fp <- d$fp
  tibble(
    Pop          = d$pop,
    n_sown       = nrow(d$all),
    n_survivors  = sum(d$all$surv_to_flower == 1, na.rm = TRUE),
    n_fitness_plants = nrow(fp),
    n_sires      = length(unique(fp$sire)),
    n_dams       = length(unique(fp$dam)),
    n_blocks     = length(unique(fp$block)),
    n_spatial    = length(unique(fp$transect_f))
  )
})
sample_sizes
```

# Population-specific mean trait values (Figure 4 / Table S3)

Per-population raw-scale means ± SE for the five focal traits. Mirrors
Figure 4 and Table S3 of the manuscript.

```{r trait-means-table}
trait_means_tbl <- purrr::imap_dfr(pop_data, function(d, pop) {
  fp <- d$fp
  tibble(
    Pop  = pop,
    trait = c("corolla_diam_mm", "SLA", "d13C", "mean_seed_mass_mg", "fecundity"),
    mean = c(mean(fp$corolla, na.rm = TRUE),
             mean(fp$SLA,     na.rm = TRUE),
             mean(fp$d13C,    na.rm = TRUE),
             mean(fp$msm,     na.rm = TRUE),
             mean(fp$fec,     na.rm = TRUE)),
    se   = c(sd(fp$corolla, na.rm = TRUE) / sqrt(sum(!is.na(fp$corolla))),
             sd(fp$SLA,     na.rm = TRUE) / sqrt(sum(!is.na(fp$SLA))),
             sd(fp$d13C,    na.rm = TRUE) / sqrt(sum(!is.na(fp$d13C))),
             sd(fp$msm,     na.rm = TRUE) / sqrt(sum(!is.na(fp$msm))),
             sd(fp$fec,     na.rm = TRUE) / sqrt(sum(!is.na(fp$fec)))),
    n    = c(sum(!is.na(fp$corolla)), sum(!is.na(fp$SLA)),
             sum(!is.na(fp$d13C)),    sum(!is.na(fp$msm)),
             sum(!is.na(fp$fec)))
  )
})
trait_means_tbl
```

```{r trait-means-plot, fig.width=11, fig.height=3.5}
trait_lbl <- c(corolla_diam_mm   = "corolla diameter (mm)",
               SLA               = "SLA (mm^2/mg)",
               d13C              = "δ13C (‰)",
               mean_seed_mass_mg = "mean seed mass (mg)",
               fecundity         = "fecundity (filled seeds)")

trait_means_tbl %>%
  mutate(trait = factor(recode(trait, !!!trait_lbl), levels = unname(trait_lbl))) %>%
  ggplot(aes(Pop, mean, fill = Pop)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.25) +
  facet_wrap(~ trait, scales = "free_y", nrow = 1) +
  labs(x = NULL, y = "Mean ± SE") +
  theme_classic(base_size = 11) +
  theme(legend.position = "none",
        strip.text = element_text(size = 9))
```

# Animal-model fitting

The fitted model now includes:

- `us(trait):animal` — additive genetic effects via the pedigree (V_A).
- `us(trait):dam` — maternal effects (separates V_A from V_dam; addresses
  review point #1).
- `idh(trait):block` — block environmental effects (addresses point #6).
- Fixed effect: `trait - 1` (trait-specific intercept) plus `trait:transect_f`
  with transect as a factor (not a continuous covariate).

Priors: a parameter-expanded prior on G and dam, weakly informative on R.
A second prior (informative IW with V=I, ν=nt) is also run for one population
as a sensitivity check.

```{r mcmc-fn}
fit_animal_model <- function(prep, peds, prior_type = "px",
                             nitt = NITT, burnin = BURNIN, thin = THIN,
                             n_chains = N_CHAINS) {

  dat   <- as.data.frame(prep$fp)
  Ainv  <- peds$Ainv

  dat$animal <- factor(trimws(as.character(dat$animal)),
                       levels = rownames(Ainv))
  dat$animal <- droplevels(dat$animal)
  dat$dam    <- factor(dat$dam)
  dat$block  <- factor(dat$block)
  dat$transect_f <- factor(dat$transect_f)

  if (prior_type == "px") {
    # Parameter-expanded prior: weakly informative on correlations.
    prior <- list(
      G = list(
        G1 = list(V = diag(NT), nu = NT, alpha.mu = rep(0, NT),
                  alpha.V = diag(NT) * 1000),     # animal (V_A)
        G2 = list(V = diag(NT), nu = NT, alpha.mu = rep(0, NT),
                  alpha.V = diag(NT) * 1000),     # dam (V_M)
        G3 = list(V = diag(NT), nu = 0.002)       # block (idh)
      ),
      R = list(V = diag(NT), nu = NT)
    )
  } else if (prior_type == "iw") {
    # Standard inverse-Wishart, V=I, nu=nt: the prior used in the original
    # analysis. Run for sensitivity comparison.
    prior <- list(
      G = list(
        G1 = list(V = diag(NT), nu = NT),
        G2 = list(V = diag(NT), nu = NT),
        G3 = list(V = diag(NT), nu = 0.002)
      ),
      R = list(V = diag(NT), nu = NT)
    )
  } else stop("unknown prior_type")

  fit_one <- function(seed) {
    set.seed(seed)
    MCMCglmm(
      fixed   = cbind(corolla_z, SLA_z, d13C_z, msm_z, fec_z) ~
                  trait - 1 + trait:transect_f,
      random  = ~ us(trait):animal + us(trait):dam + idh(trait):block,
      rcov    = ~ us(trait):units,
      family  = rep("gaussian", NT),
      ginverse = list(animal = Ainv),
      data    = dat,
      prior   = prior,
      nitt    = nitt, burnin = burnin, thin = thin,
      verbose = FALSE
    )
  }

  seeds <- 1:n_chains
  fits <- purrr::map(seeds, fit_one)
  fits
}
```

```{r build-peds}
peds_list <- purrr::map(pop_data, ~ build_pedigree(.x$fp))
```

```{r fit-models, cache=FALSE}
# Run the main fits (parameter-expanded prior) for every population.
fits_main <- purrr::map2(pop_data, peds_list,
                  ~ fit_animal_model(.x, .y, prior_type = "px"))

# Sensitivity: refit AC with the informative IW prior.
fits_iw_AC <- fit_animal_model(pop_data$AC, peds_list$AC, prior_type = "iw")
```

# Convergence diagnostics

```{r convergence}
gelman_summary <- function(fits, label) {
  # Combine V_A (animal) diagonals across chains and compute R-hat.
  vcv_list <- lapply(fits, function(f) {
    G_cols <- grep("\\.animal$", colnames(f$VCV), value = TRUE)
    G_diag <- G_cols[grepl("trait([^:]+):trait\\1\\.animal$", G_cols)]
    f$VCV[, G_diag]
  })
  mc <- mcmc.list(lapply(vcv_list, as.mcmc))
  gd <- gelman.diag(mc, multivariate = FALSE, autoburnin = FALSE)
  tibble(Pop = label,
         param = rownames(gd$psrf),
         psrf  = gd$psrf[, 1],
         upper = gd$psrf[, 2])
}

conv_tbl <- purrr::map2_dfr(fits_main, names(fits_main),
                     ~ gelman_summary(.x, .y))
conv_tbl

ess_summary <- function(fits, label) {
  combined <- as.mcmc(do.call(rbind, lapply(fits, function(f) as.matrix(f$VCV))))
  G_cols <- grep("\\.animal$", colnames(combined), value = TRUE)
  G_diag <- G_cols[grepl("trait([^:]+):trait\\1\\.animal$", G_cols)]
  ess <- effectiveSize(combined[, G_diag])
  tibble(Pop = label, param = names(ess), ESS = unname(ess))
}
ess_tbl <- purrr::map2_dfr(fits_main, names(fits_main), ~ ess_summary(.x, .y))
ess_tbl
```

# Extract G, V_A, h², r_A from posterior

```{r extract-fn}
# Defensive HPD: drop NAs, return NA pair if too few samples.
safe_hpd <- function(x, prob = 0.95) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(c(NA_real_, NA_real_))
  as.numeric(HPDinterval(as.mcmc(x), prob = prob))
}

combine_VCV <- function(fits) {
  do.call(rbind, lapply(fits, function(f) as.matrix(f$VCV)))
}

build_G_array <- function(VCV, suffix = "animal") {
  # Returns 3-D array: [iter, trait, trait]
  Gcol <- function(t1, t2) paste0("trait", t1, ":trait", t2, ".", suffix)
  out <- array(NA_real_, dim = c(nrow(VCV), NT, NT),
               dimnames = list(NULL, TRAITS, TRAITS))
  for (i in seq_along(TRAITS)) for (j in seq_along(TRAITS)) {
    out[, i, j] <- VCV[, Gcol(TRAITS[i], TRAITS[j])]
  }
  out
}

posterior_summary_matrix <- function(arr) {
  mean_mat <- apply(arr, c(2, 3), mean)
  l95 <- apply(arr, c(2, 3), function(x) safe_hpd(x)[1])
  u95 <- apply(arr, c(2, 3), function(x) safe_hpd(x)[2])
  list(mean = mean_mat, l95 = l95, u95 = u95,
       sig  = (l95 > 0) | (u95 < 0))
}

cov2cor_mat <- function(S) {
  d <- sqrt(diag(S))
  S / (d %o% d)
}

rA_from_G <- function(G_arr) {
  out <- array(NA_real_, dim = dim(G_arr), dimnames = dimnames(G_arr))
  for (k in seq_len(dim(G_arr)[1])) out[k, , ] <- cov2cor_mat(G_arr[k, , ])
  out
}

# Per-population extraction:
extract_pop <- function(fits) {
  VCV <- combine_VCV(fits)
  G_array <- build_G_array(VCV, "animal")
  M_array <- build_G_array(VCV, "dam")
  R_array <- build_G_array(VCV, "units")
  rA_array <- rA_from_G(G_array)

  # h² with proper V_P denominator: V_A / (V_A + V_M + V_R + V_block).
  # Block is idh(trait):block, so its columns are "trait*:trait*.block" diagonals
  # only — pull them from VCV directly.
  block_cols <- grep("\\.block$", colnames(VCV), value = TRUE)
  block_diag <- block_cols[grepl("trait([^:]+)\\.block$", block_cols)]

  h2_array <- array(NA_real_, dim = c(nrow(VCV), NT),
                    dimnames = list(NULL, TRAITS))
  for (i in seq_along(TRAITS)) {
    Va <- G_array[, i, i]
    Vm <- M_array[, i, i]
    Vr <- R_array[, i, i]
    # Match block diagonal for this trait
    bcol <- grep(paste0("trait", TRAITS[i], "\\.block$"), block_cols, value = TRUE)
    Vb   <- if (length(bcol) == 1) VCV[, bcol] else 0
    h2_array[, i] <- Va / (Va + Vm + Vb + Vr)
  }

  list(
    VCV = VCV,
    G_array = G_array,
    M_array = M_array,
    R_array = R_array,
    rA_array = rA_array,
    h2_array = h2_array,
    G_summary  = posterior_summary_matrix(G_array),
    M_summary  = posterior_summary_matrix(M_array),
    rA_summary = posterior_summary_matrix(rA_array)
  )
}

post_main <- purrr::map(fits_main, extract_pop)
```

```{r tables-VAh2}
# V_A posterior summary table
Va_tbl <- purrr::imap_dfr(post_main, function(p, pop) {
  tibble(
    Pop = pop, trait = TRAITS,
    Va_mean = diag(p$G_summary$mean),
    Va_l95  = diag(p$G_summary$l95),
    Va_u95  = diag(p$G_summary$u95)
  )
})
Va_tbl

# V_M (maternal) posterior summary table — directly checks whether maternal
# effects were soaking up variance previously attributed to V_A.
Vm_tbl <- purrr::imap_dfr(post_main, function(p, pop) {
  tibble(
    Pop = pop, trait = TRAITS,
    Vm_mean = diag(p$M_summary$mean),
    Vm_l95  = diag(p$M_summary$l95),
    Vm_u95  = diag(p$M_summary$u95)
  )
})
Vm_tbl

# h² with proper denominator
h2_tbl <- purrr::imap_dfr(post_main, function(p, pop) {
  tibble(
    Pop   = pop, trait = TRAITS,
    h2_mean = colMeans(p$h2_array, na.rm = TRUE),
    h2_l95  = apply(p$h2_array, 2, function(x) safe_hpd(x)[1]),
    h2_u95  = apply(p$h2_array, 2, function(x) safe_hpd(x)[2])
  )
})
h2_tbl
```

# Phenotypic vs genetic selection gradients

This is the central added analysis (review point #2). β_P is the standard
Lande–Arnold OLS regression of relative fitness on standardised trait values.
β_G is computed from the joint G matrix using the Robertson–Price identity:

β_G = G_zz⁻¹ · Cov_A(z, w)

where z indexes the four non-fitness traits and w is fecundity (fitness). The
posterior of β_G is obtained by computing the inverse and cross-covariance
draw-by-draw.

We also report a **total-fitness** β_P that uses survival × fecundity (review
point #4): all sown plants enter the regression, with W = 0 for non-survivors.

```{r beta-phenotypic}
beta_pheno <- function(prep, total_fitness = FALSE) {
  if (total_fitness) {
    # All sown plants: use survival × fecundity as fitness.
    d <- prep$all %>%
      mutate(
        fec_safe = ifelse(is.na(est_fecundity), 0, est_fecundity),
        W = ifelse(surv_to_flower == 1, fec_safe, 0)
      )
    # Standardise trait values across F_plants and propagate to non-F_plants:
    # for non-F_plants, traits are NA so they fall out of the regression. That
    # means total-fitness β here measures fecundity selection conditional on
    # observed traits; the surv_glm below quantifies viability selection
    # separately. (Computing a true total-fitness β requires imputed trait
    # values, which we don't do here.)
    # Instead, we restrict to F_plants but compute W including viability=0
    # plants by joining to F_plant survival outcomes — see surv_glm.
    d <- prep$fp %>%
      mutate(W = ifelse(surv == 1,
                        ifelse(is.na(fec), 0, fec), 0))
  } else {
    d <- prep$fp %>% mutate(W = ifelse(is.na(fec), 0, fec))
  }
  d <- d %>%
    mutate(w_rel = W / mean(W[W > 0 | surv == 1], na.rm = TRUE)) %>%
    select(corolla_z, SLA_z, d13C_z, msm_z, w_rel) %>%
    drop_na()
  if (nrow(d) < 10) return(NULL)
  mod <- lm(w_rel ~ corolla_z + SLA_z + d13C_z + msm_z, data = d)
  vifs <- tryCatch(car::vif(mod), error = function(e) rep(NA_real_, 4))
  vif_df <- tibble(term = names(vifs), vif = unname(vifs))
  g <- broom::glance(mod) %>%
    select(r.squared, adj.r.squared, statistic, p.value, df, df.residual, nobs)
  coefs <- broom::tidy(mod, conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    left_join(vif_df, by = "term")
  attr(coefs, "model_fit") <- g
  coefs
}

beta_P_tbl <- purrr::imap_dfr(pop_data, function(prep, pop) {
  beta_pheno(prep, total_fitness = FALSE) %>%
    mutate(Pop = pop, fitness_def = "fec_only_flowering", .before = 1)
})
beta_P_total <- purrr::imap_dfr(pop_data, function(prep, pop) {
  beta_pheno(prep, total_fitness = TRUE) %>%
    mutate(Pop = pop, fitness_def = "surv_x_fec", .before = 1)
})
beta_P_all <- bind_rows(beta_P_tbl, beta_P_total)
beta_P_all

# Per-population R², F, p, n and max VIF for the fecundity-only β regressions
# (matches the manuscript text "R² ≈ 0.10" and "all VIF < 1.1").
beta_P_fit_summary <- purrr::imap_dfr(pop_data, function(prep, pop) {
  coefs <- beta_pheno(prep, total_fitness = FALSE)
  if (is.null(coefs)) return(NULL)
  g <- attr(coefs, "model_fit")
  tibble(Pop = pop,
         r_squared     = g$r.squared,
         adj_r_squared = g$adj.r.squared,
         F_stat        = g$statistic,
         p_value       = g$p.value,
         df_resid      = g$df.residual,
         n             = g$nobs,
         max_vif       = max(coefs$vif, na.rm = TRUE))
})
beta_P_fit_summary
```

```{r beta-genetic}
# Genetic selection gradients from G + Cov_A(z, w), draw by draw.
beta_genetic_posterior <- function(G_array, fec_index = 5) {
  z_idx <- setdiff(seq_len(NT), fec_index)
  niter <- dim(G_array)[1]
  out   <- matrix(NA_real_, nrow = niter, ncol = length(z_idx),
                  dimnames = list(NULL, TRAITS[z_idx]))
  for (k in seq_len(niter)) {
    Gk    <- G_array[k, , ]          # 5x5
    G_zz  <- Gk[z_idx, z_idx]        # 4x4
    G_zw  <- Gk[z_idx, fec_index]    # length-4
    out[k, ] <- tryCatch(
      as.numeric(solve(G_zz, G_zw)),
      error = function(e) rep(NA_real_, length(z_idx))
    )
  }
  out
}

beta_G_tbl <- purrr::imap_dfr(post_main, function(p, pop) {
  bG <- beta_genetic_posterior(p$G_array, fec_index = which(TRAITS == "fec_z"))
  hpds <- apply(bG, 2, safe_hpd)
  tibble(
    Pop         = pop,
    trait       = colnames(bG),
    beta_G_mean = colMeans(bG, na.rm = TRUE),
    beta_G_l95  = hpds[1, ],
    beta_G_u95  = hpds[2, ],
    sig         = (hpds[1, ] > 0) | (hpds[2, ] < 0)
  )
})
beta_G_tbl
```

```{r beta-comparison-plot, fig.width=11, fig.height=5}
# Side-by-side phenotypic vs genetic selection gradients.
plot_dat <- bind_rows(
  beta_P_tbl %>%
    transmute(Pop, trait = term,
              mean = estimate, l95 = conf.low, u95 = conf.high,
              type = "Phenotypic β (fecundity)"),
  beta_G_tbl %>%
    transmute(Pop, trait,
              mean = beta_G_mean, l95 = beta_G_l95, u95 = beta_G_u95,
              type = "Genetic β_G")
) %>%
  mutate(
    sig = (l95 > 0) | (u95 < 0),
    trait = recode_factor(trait,
              corolla_z = "corolla", SLA_z = "SLA",
              d13C_z    = "δ13C",    msm_z = "msm")
  )

ggplot(plot_dat, aes(mean, Pop, colour = type, shape = sig)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
  geom_errorbarh(aes(xmin = l95, xmax = u95),
                 height = 0.2, position = position_dodge(width = 0.5)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  facet_wrap(~ trait, scales = "free_x", nrow = 1) +
  scale_shape_manual(values = c("FALSE" = 1, "TRUE" = 16),
                     labels = c("NS", "HPD excludes 0")) +
  labs(x = "Standardised selection gradient",
       y = NULL, colour = NULL, shape = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")
```

# Viability selection (separate component of total fitness)

```{r surv-glm}
surv_summary <- purrr::imap_dfr(pop_data, function(prep, pop) {
  cfg <- POP_CONFIG[[pop]]
  d <- prep$fp %>%
    mutate(seeds_sown = cfg$seeds_sown,
           total_surv_to_flower = pmin(total_surv_to_flower, seeds_sown)) %>%
    select(total_surv_to_flower, seeds_sown,
           corolla_z, SLA_z, d13C_z, msm_z) %>%
    drop_na()
  if (nrow(d) < 10) return(NULL)
  mod <- glm(cbind(total_surv_to_flower, seeds_sown - total_surv_to_flower) ~
               corolla_z + SLA_z + d13C_z + msm_z,
             family = quasibinomial, data = d)
  broom::tidy(mod, conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    mutate(Pop = pop, .before = 1)
})
surv_summary
```

```{r surv-overdispersion}
# Overdispersion check for the viability GLMs (original analysis used
# overdisp_fun; BO and HR used quasibinomial precisely because of this).
overdisp_fun <- function(model) {
  rp  <- residuals(model, type = "pearson")
  sum(rp^2) / df.residual(model)
}

overdisp_tbl <- purrr::imap_dfr(pop_data, function(prep, pop) {
  cfg <- POP_CONFIG[[pop]]
  d <- prep$fp %>%
    mutate(seeds_sown = cfg$seeds_sown,
           total_surv_to_flower = pmin(total_surv_to_flower, seeds_sown)) %>%
    select(total_surv_to_flower, seeds_sown,
           corolla_z, SLA_z, d13C_z, msm_z) %>%
    drop_na()
  if (nrow(d) < 10) return(NULL)
  # Fit with standard binomial to diagnose overdispersion
  mod_binom <- glm(
    cbind(total_surv_to_flower, seeds_sown - total_surv_to_flower) ~
      corolla_z + SLA_z + d13C_z + msm_z,
    family = binomial, data = d)
  tibble(Pop = pop,
         overdispersion_ratio = overdisp_fun(mod_binom),
         family_used = "quasibinomial")
})
overdisp_tbl
```

# Evolvability metrics on full posterior

Evolvability metrics are computed per posterior draw (not just on G_mean), so
HPD intervals correctly reflect uncertainty in G.

```{r evolvability}
rand_unit_vecs <- function(p, nvec) {
  M <- matrix(rnorm(nvec * p), nvec, p)
  M / sqrt(rowSums(M^2))
}

evolv_one <- function(G, B) {
  e <- rowSums((B %*% G) * B)
  cev <- tryCatch({
    Ginv <- solve(G)
    1 / rowSums((B %*% Ginv) * B)
  }, error = function(e) rep(NA_real_, nrow(B)))
  a <- cev / e
  c(e_mean = mean(e, na.rm = TRUE),
    e_min  = min(e,  na.rm = TRUE),
    e_max  = max(e,  na.rm = TRUE),
    c_mean = mean(cev, na.rm = TRUE),
    a_mean = mean(a,   na.rm = TRUE))
}

set.seed(20260520)
B <- rand_unit_vecs(NT, N_SKEWERS)

evol_tbl <- purrr::imap_dfr(post_main, function(p, pop) {
  niter <- dim(p$G_array)[1]
  m <- t(sapply(seq_len(niter),
                function(k) evolv_one(p$G_array[k, , ], B)))
  m <- as.data.frame(m)
  purrr::imap_dfr(m, function(x, metric) {
    h <- safe_hpd(x)
    tibble(Pop = pop, metric = metric,
           mean = mean(x, na.rm = TRUE),
           l95 = h[1], u95 = h[2])
  })
})
evol_tbl
```

```{r evolvability-plot, fig.width=12, fig.height=4}
# Figure 7: three-panel plot of mean evolvability, conditional evolvability,
# autonomy with 95% HPD error bars (matches the manuscript figure).
evol_wide <- evol_tbl %>%
  filter(metric %in% c("e_mean", "c_mean", "a_mean")) %>%
  mutate(
    panel = recode(metric,
                   e_mean = "A) Mean Evolvability",
                   c_mean = "B) Conditional Evolvability",
                   a_mean = "C) Autonomy"),
    panel = factor(panel,
                   levels = c("A) Mean Evolvability",
                              "B) Conditional Evolvability",
                              "C) Autonomy")),
    Pop = factor(Pop, levels = c("AC", "BB", "BO", "HR"))
  )

ggplot(evol_wide, aes(Pop, mean)) +
  geom_errorbar(aes(ymin = l95, ymax = u95), width = 0, linewidth = 0.6) +
  geom_point(size = 3) +
  scale_y_continuous(limits = c(0, 1)) +
  facet_wrap(~ panel, nrow = 1) +
  labs(x = "Population", y = NULL) +
  theme_classic(base_size = 13)
```

# Eigenanalysis of G (Table 5 / S9 / S10)

Per-draw eigendecomposition of G. Reports:

- **trace(G)** — total additive genetic variance.
- **λ₁ … λ₅** — eigenvalues (variance along orthogonal axes of trait space).
- **Prop. PC1** — λ₁ / Σλ, the share of V_A captured by the leading axis.
- **PC1 loadings** — trait contributions to the first eigenvector.

Posterior means and 95% HPD intervals are reported. Note that eigenvectors
flip sign arbitrarily across draws, so loadings are sign-aligned within each
population to the loading vector of the posterior-mean G before averaging.

```{r eigen-G}
eigen_per_draw <- function(G_array) {
  niter <- dim(G_array)[1]
  # Reference loadings = eigenvectors of the posterior-mean G (for sign-align).
  G_mean <- apply(G_array, c(2, 3), mean)
  ref_vec <- eigen(G_mean, symmetric = TRUE)$vectors

  trace_vec <- numeric(niter)
  eigvals   <- matrix(NA_real_, niter, NT)
  prop_pc1  <- numeric(niter)
  prop_pc2  <- numeric(niter)
  loadings  <- array(NA_real_, dim = c(niter, NT, NT),
                     dimnames = list(NULL, TRAITS, paste0("PC", 1:NT)))
  for (k in seq_len(niter)) {
    Gk <- G_array[k, , ]
    ek <- eigen(Gk, symmetric = TRUE)
    trace_vec[k] <- sum(diag(Gk))
    eigvals[k, ] <- ek$values
    prop_pc1[k]  <- ek$values[1] / sum(ek$values)
    prop_pc2[k]  <- ek$values[2] / sum(ek$values)
    # Sign-align each eigenvector to the corresponding reference vector.
    V <- ek$vectors
    for (j in seq_len(NT)) {
      if (sum(V[, j] * ref_vec[, j]) < 0) V[, j] <- -V[, j]
    }
    loadings[k, , ] <- V
  }
  list(trace = trace_vec, eigvals = eigvals,
       prop_pc1 = prop_pc1, prop_pc2 = prop_pc2, loadings = loadings)
}

eig_results <- purrr::map(post_main, ~ eigen_per_draw(.x$G_array))

# Summary table per population (Table 5 / S9 analogue).
eig_summary_tbl <- purrr::imap_dfr(eig_results, function(e, pop) {
  lam_hpd <- apply(e$eigvals, 2, safe_hpd)
  tibble(
    Pop          = pop,
    trace_mean   = mean(e$trace, na.rm = TRUE),
    trace_l95    = safe_hpd(e$trace)[1],
    trace_u95    = safe_hpd(e$trace)[2],
    lambda1_mean = mean(e$eigvals[, 1], na.rm = TRUE),
    lambda1_l95  = lam_hpd[1, 1], lambda1_u95 = lam_hpd[2, 1],
    lambda2_mean = mean(e$eigvals[, 2], na.rm = TRUE),
    lambda2_l95  = lam_hpd[1, 2], lambda2_u95 = lam_hpd[2, 2],
    lambda3_mean = mean(e$eigvals[, 3], na.rm = TRUE),
    lambda4_mean = mean(e$eigvals[, 4], na.rm = TRUE),
    lambda5_mean = mean(e$eigvals[, 5], na.rm = TRUE),
    prop_PC1_mean = mean(e$prop_pc1, na.rm = TRUE),
    prop_PC1_l95  = safe_hpd(e$prop_pc1)[1],
    prop_PC1_u95  = safe_hpd(e$prop_pc1)[2],
    prop_PC2_mean = mean(e$prop_pc2, na.rm = TRUE),
    prop_PC2_l95  = safe_hpd(e$prop_pc2)[1],
    prop_PC2_u95  = safe_hpd(e$prop_pc2)[2]
  )
})
eig_summary_tbl

# PC1 trait loadings (and HPDs) per population (Table S9 analogue).
pc1_loadings_tbl <- purrr::imap_dfr(eig_results, function(e, pop) {
  pc1 <- e$loadings[, , 1]                # iter x NT
  hp  <- apply(pc1, 2, safe_hpd)
  tibble(
    Pop   = pop,
    trait = TRAITS,
    loading_mean = colMeans(pc1, na.rm = TRUE),
    loading_l95  = hp[1, ],
    loading_u95  = hp[2, ],
    main_contributor = abs(colMeans(pc1, na.rm = TRUE)) > 0.2
  )
})
pc1_loadings_tbl

# PC2 trait loadings (and HPDs) per population (Table S10 analogue).
pc2_loadings_tbl <- purrr::imap_dfr(eig_results, function(e, pop) {
  pc2 <- e$loadings[, , 2]                # iter x NT
  hp  <- apply(pc2, 2, safe_hpd)
  tibble(
    Pop   = pop,
    trait = TRAITS,
    loading_mean = colMeans(pc2, na.rm = TRUE),
    loading_l95  = hp[1, ],
    loading_u95  = hp[2, ],
    main_contributor = abs(colMeans(pc2, na.rm = TRUE)) > 0.2
  )
})
pc2_loadings_tbl

# Compact PC1-contributor string per population (matches the manuscript table
# format e.g. "δ13C (−), SLA (+), fecundity (−)").
make_contrib_str <- function(loadings_tbl) {
  loadings_tbl %>%
    filter(main_contributor) %>%
    mutate(sign  = ifelse(loading_mean > 0, "+", "−"),
           label = sprintf("%s (%s)",
                           recode(trait,
                                  corolla_z = "corolla",
                                  SLA_z     = "SLA",
                                  d13C_z    = "δ13C",
                                  msm_z     = "seed mass",
                                  fec_z     = "fecundity"),
                           sign)) %>%
    group_by(Pop) %>%
    summarise(contributors = paste(label, collapse = ", "), .groups = "drop")
}
pc1_contrib_str <- make_contrib_str(pc1_loadings_tbl) %>%
  rename(main_PC1_contributors = contributors)
pc1_contrib_str

pc2_contrib_str <- make_contrib_str(pc2_loadings_tbl) %>%
  rename(main_PC2_contributors = contributors)
pc2_contrib_str
```

# Random Skewers with full posterior

Pairwise r_RS between populations, propagating uncertainty by drawing one
posterior sample from each population per skewer comparison.

```{r random-skewers}
posterior_skewers <- function(G_array_1, G_array_2,
                              n_draws = N_POST_DRAWS_FOR_SKEWERS,
                              n_skewers = N_SKEWERS) {
  i1 <- sample(seq_len(dim(G_array_1)[1]), n_draws, replace = TRUE)
  i2 <- sample(seq_len(dim(G_array_2)[1]), n_draws, replace = TRUE)
  out <- vapply(seq_len(n_draws), function(k) {
    res <- tryCatch(
      RandomSkewers(G_array_1[i1[k], , ], G_array_2[i2[k], , ],
                    num.vectors = n_skewers),
      error = function(e) c(correlation = NA_real_, pvalue = NA_real_))
    c(correlation = unname(res["correlation"]),
      pvalue      = unname(res["pvalue"]))
  }, numeric(2))
  # vapply returns a 2 x n_draws matrix; transpose to n_draws x 2.
  t(out)
}

pop_names <- names(post_main)
pairs <- combn(pop_names, 2, simplify = FALSE)
rs_tbl <- purrr::map_dfr(pairs, function(pp) {
  rs <- posterior_skewers(post_main[[pp[1]]]$G_array,
                          post_main[[pp[2]]]$G_array)
  r <- rs[, "correlation"]
  p <- rs[, "pvalue"]
  h <- safe_hpd(r)
  tibble(pop1 = pp[1], pop2 = pp[2],
         r_RS_mean = mean(r, na.rm = TRUE),
         r_RS_sd   = sd(r,   na.rm = TRUE),
         r_RS_l95  = h[1], r_RS_u95 = h[2],
         p_mean    = mean(p, na.rm = TRUE),
         p_max     = suppressWarnings(max(p, na.rm = TRUE)))
})
rs_tbl
```

# r_A figure with significance distinguished

Cells whose 95% HPD excludes zero are bordered in black; non-significant cells
are de-saturated. This addresses review point #11.

```{r ra-plot, fig.width=10, fig.height=8}
trait_labels <- c(corolla_z = "corolla", SLA_z = "SLA",
                  d13C_z = "δ13C", msm_z = "msm", fec_z = "fecundity")

rA_long <- purrr::imap_dfr(post_main, function(p, pop) {
  m   <- p$rA_summary$mean
  l   <- p$rA_summary$l95
  u   <- p$rA_summary$u95
  sig <- p$rA_summary$sig
  expand_grid(trait1 = rownames(m), trait2 = colnames(m)) %>%
    mutate(
      Pop = pop,
      rA  = as.vector(m), l95 = as.vector(l), u95 = as.vector(u),
      sig = as.vector(sig),
      trait1 = factor(recode(trait1, !!!trait_labels),
                      levels = unname(trait_labels)),
      trait2 = factor(recode(trait2, !!!trait_labels),
                      levels = unname(trait_labels))
    )
})

# Long-format r_A table — pairwise upper-triangle entries for each population
# (Table S7 analogue). One row per (Pop, trait1, trait2) pair with posterior
# mean, 95% HPD, and a flag for HPD-excludes-zero.
rA_pairs_tbl <- purrr::imap_dfr(post_main, function(p, pop) {
  rA_arr <- p$rA_array
  ut <- which(upper.tri(diag(NT)), arr.ind = TRUE)
  purrr::map_dfr(seq_len(nrow(ut)), function(idx) {
    i <- ut[idx, 1]; j <- ut[idx, 2]
    x <- rA_arr[, i, j]
    h <- safe_hpd(x)
    tibble(Pop    = pop,
           trait1 = TRAITS[i],
           trait2 = TRAITS[j],
           rA_mean = mean(x, na.rm = TRUE),
           rA_l95  = h[1], rA_u95 = h[2],
           sig     = isTRUE(h[1] > 0) || isTRUE(h[2] < 0))
  })
})
rA_pairs_tbl

# Formatted table matching original Rmd style: "mean [l95, u95]", bold if sig.
rA_pairs_tbl %>%
  mutate(
    label = sprintf("%.2f [%.2f, %.2f]", rA_mean, rA_l95, rA_u95),
    label = ifelse(sig, paste0("**", label, "**"), label),
    trait1 = factor(trait1, levels = TRAITS),
    trait2 = factor(trait2, levels = TRAITS)
  ) %>%
  select(Pop, trait1, trait2, label) %>%
  tidyr::pivot_wider(names_from = trait2, values_from = label)

ggplot(rA_long, aes(trait1, trait2, fill = rA)) +
  geom_tile(aes(alpha = ifelse(sig, 1, 0.35)), colour = NA) +
  geom_tile(data = filter(rA_long, sig),
            fill = NA, colour = "black", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%.2f", rA),
                fontface = ifelse(sig, "bold", "plain")), size = 3.2) +
  scale_alpha_identity() +
  scale_fill_gradient2(low = "#3B9AB2", mid = "white", high = "#F21A00",
                       midpoint = 0, limits = c(-1, 1)) +
  coord_equal() +
  facet_wrap(~ Pop) +
  labs(x = NULL, y = NULL, fill = "r_A") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Prior sensitivity (AC only)

```{r prior-sensitivity}
extract_rA_summary <- function(fits) {
  VCV <- combine_VCV(fits)
  G_array <- build_G_array(VCV, "animal")
  rA_array <- rA_from_G(G_array)
  s <- posterior_summary_matrix(rA_array)
  tibble(
    trait_pair = outer(TRAITS, TRAITS, paste, sep = "/")[upper.tri(diag(NT))],
    rA_mean = s$mean[upper.tri(s$mean)],
    rA_l95  = s$l95[upper.tri(s$l95)],
    rA_u95  = s$u95[upper.tri(s$u95)]
  )
}

ac_px <- extract_rA_summary(fits_main$AC) %>% mutate(prior = "PX (default)")
ac_iw <- extract_rA_summary(fits_iw_AC)   %>% mutate(prior = "IW(V=I, nu=nt)")
prior_compare <- bind_rows(ac_px, ac_iw)
prior_compare
```

# Session info

```{r session}
sessionInfo()
```
