---
title: "Multivariate Animal Model"
author: "Helen Payne"
date: "2026-05-20"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
---

```{r setup, include=FALSE}
# Persist figures as PNGs in figures/ alongside the embedded HTML versions.
# Tables are written to tables/ in the save-outputs chunk near the end.
dir.create("figures", showWarnings = FALSE)
dir.create("tables",  showWarnings = FALSE)
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.path = "figures/", dpi = 300, dev = "png"
)
```

# 1 · Motivation

This document fits a multivariate Bayesian animal model to four *Nemophila
menziesii* populations (AC, BB, BO, HR) and uses the resulting **G** matrices
to characterise the genetic architecture of five focal traits — corolla
diameter, specific leaf area (SLA), δ¹³C, mean seed mass (msm), and
fecundity — and the selection acting on them.

It is a re-implementation of an earlier analysis that the manuscript review
flagged on a number of methodological points. The structure of this document
walks through each step of the analysis in the order in which it must be
executed, with a short explanation before each block of code.

The five biggest changes versus the original analysis are:

1. **Maternal (dam) random effect** — separates V_A from V_maternal in the
   paternal half-sib design.
2. **Block and within-block spatial random effects** — the field design has 3
   blocks per site; treating spatial structure as fixed factors blew up at BB
   (158 Locations × 5 traits ≈ 775 parameters for ~441 plants).
3. **Genetic selection gradient β_G** — computed from the posterior of **G**
   via the Robertson–Price identity, alongside the standard Lande–Arnold β_P.
4. **Full posterior propagation** — evolvability metrics, eigenanalysis, and
   Random Skewers are computed per draw, not on the posterior mean of **G**.
5. **Deterministic data cleaning** — no unseeded `sample()` calls; the
   F-plant substitution rule is fully reproducible.

The remainder is laid out as: setup → data → model → diagnostics → posterior
quantities → genetic correlations → eigenanalysis → evolvability → population
comparison → selection → viability → sensitivity analyses.

# 2 · Setup

## 2.1 Libraries and namespace hygiene

`evolqg` (for Random Skewers) loads `plyr`, which masks several `dplyr`
verbs. The `conflicted` package below forces `dplyr` to win whenever both
versions are on the search path.

```{r libs}
options(rgl.useNULL = TRUE)   # evolqg pulls in rgl; avoid X11 dependency

suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(MCMCglmm)
  library(nadiv)
  library(coda)
  library(evolqg)
  library(broom)
  library(patchwork)
  library(car)        # vif()
  library(sandwich)   # HC3 robust variance
  library(lmtest)     # coeftest / waldtest with vcov.
  library(multcompView) # compact letter displays from Tukey HSD
  library(ggh4x)        # per-facet scale overrides (δ¹³C y-axis inversion)
})

suppressPackageStartupMessages(library(conflicted))
for (fn in c("mutate", "rename", "filter", "select", "summarise", "summarize",
             "arrange", "group_by", "ungroup", "transmute", "count", "tally",
             "distinct", "bind_rows", "bind_cols", "recode")) {
  try(conflicted::conflict_prefer(fn, "dplyr", quiet = TRUE), silent = TRUE)
}

set.seed(20260520)
```

## 2.2 Run-time configuration

`TEST_RUN <- TRUE` runs short MCMC chains (1–2 min total) for debugging.
Set to `FALSE` for the publication-quality fit; the difference is roughly
two orders of magnitude in run time.

```{r runtime-config}
TEST_RUN <- FALSE

if (TEST_RUN) {
  N_CHAINS <- 2
  NITT     <- 6000
  BURNIN   <- 1000
  THIN     <- 5
  N_SKEWERS <- 200
  N_POST_DRAWS_FOR_SKEWERS <- 100   # subsample posterior to speed up
} else {
  N_CHAINS <- 3
  NITT     <- 200000
  BURNIN   <- 20000
  THIN     <- 100
  N_SKEWERS <- 2000
  N_POST_DRAWS_FOR_SKEWERS <- 1000
}
```

## 2.2.1 Fit caching

The full-resolution MCMC fits take ~3–5 h on this machine. To make it
practical to re-knit just the downstream sections (β_G, plots, sensitivity)
without re-running MCMC, every fit is saved to `cache/` as an RDS file and
reloaded on subsequent knits. Set `FORCE_REFIT <- TRUE` to bypass the cache
(e.g. after changing the model formula or prior).

```{r fit-cache-setup}
FORCE_REFIT <- FALSE
CACHE_DIR   <- here::here("cache")
dir.create(CACHE_DIR, showWarnings = FALSE, recursive = TRUE)

cache_fit <- function(name, expr) {
  fpath <- file.path(CACHE_DIR, paste0(name, ".rds"))
  if (!FORCE_REFIT && file.exists(fpath)) {
    message("Loading cached fit: ", name)
    return(readRDS(fpath))
  }
  message("Fitting (not cached): ", name)
  obj <- force(expr)
  saveRDS(obj, fpath)
  obj
}
```

## 2.3 Population configuration and trait list

Each population uses a slightly different field layout. AC has 10 seeds per
1-m segment and uses transects; BB has 8 seeds and uses 158 named
Locations; BO and HR each have 8 seeds and use transects. AC also includes
a watered-treatment block ("W") that is excluded from this analysis.

```{r pop-config}
POP_CONFIG <- list(
  AC = list(file = "data/AC_compiledsheet_full_2022.csv",
            spatial = "Transect", seeds_sown = 10, exclude_blocks = "W"),
  BB = list(file = "data/BB_compiledsheet_full_2022.csv",
            spatial = "Location", seeds_sown = 8,  exclude_blocks = "W"),
  BO = list(file = "data/BO_compiledsheet_full_2022.csv",
            spatial = "Transect", seeds_sown = 8,  exclude_blocks = "W"),
  HR = list(file = "data/HR_compiledsheet_full_2022.csv",
            spatial = "Transect", seeds_sown = 8,  exclude_blocks = "W")
)

TRAITS <- c("corolla_z", "SLA_z", "d13C_z", "msm_z", "fec_z")
NT     <- length(TRAITS)
```

## 2.4 Pull data from Dryad

`fetch_dryad()` downloads the four compiled sheets from Dryad
(doi:10.5061/dryad.pvmcvdp1p) into `data/` if they are not already there.
When the Dryad package is still private, files placed in `data/` by hand
take precedence and the fetch is a no-op.

```{r fetch-data}
source(here::here("fetch_dryad.R"))
fetch_dryad(c("AC_compiledsheet_full_2022.csv",
              "BB_compiledsheet_full_2022.csv",
              "BO_compiledsheet_full_2022.csv",
              "HR_compiledsheet_full_2022.csv"))
```

# 3 · Data preparation

## 3.1 The per-population cleaning function

`prep_population()` does four things, in order:

1. **Normalises donor and recipient labels** so the same dam appears under
   one name in the pedigree (the raw sheets use mixed prefixes).
2. **Repairs the survival flag**. If a plant has `total_fruits > 0` but is
   marked `surv_to_flower == 0`, it clearly survived to flowering — flip it
   to 1. The `!is.na(total_fruits) &` guard keeps NAs as NAs (otherwise
   `ifelse(NA, …)` would silently turn recorded 0s into NAs).
3. **Builds a true segment key**: `paste(Block, Recipient, spatial_unit)`.
   Earlier versions used `ceiling(row_number() / 8)` to bin segments, which
   silently mixed AC segments together (AC has 10 seeds per segment, not 8).
   The current key is robust across populations.
4. **Picks the F-plant deterministically**. The original protocol designated
   one focal plant per segment for trait measurement, but if that plant
   died before flowering a sibling was substituted. We reproduce that rule
   by promoting the **first surviving sibling** in row order, and as a last
   resort the first row (which will then enter the model with W = 0).

The output is a list with `all` (all rows, used for the viability GLM) and
`fp` (F-plants only, used for the animal model).

```{r prep-fn}
prep_population <- function(pop_code, cfg) {
  fpath <- here::here(cfg[["file"]])
  raw <- readr::read_csv(fpath, show_col_types = FALSE)
  spatial <- cfg$spatial

  if (length(cfg$exclude_blocks) > 0) {
    raw <- raw %>% filter(!(Block %in% cfg$exclude_blocks))
  }

  raw <- raw %>%
    mutate(
      Donor     = paste0("Donor_",     pop_code, "_",
                         str_remove(Donor,
                                    paste0("^Donor_", pop_code, "_|^",
                                           pop_code, "_"))),
      Recipient = paste0("Recipient_", pop_code, "_",
                         str_remove(Recipient,
                                    paste0("^Recipient_", pop_code, "_|^",
                                           pop_code, "_"))),
      F_plant   = ifelse(F_plant == TRUE, 1L, 0L),
      F_plant   = ifelse(is.na(F_plant), 0L, F_plant),
      surv_to_flower = ifelse(!is.na(total_fruits) & total_fruits > 0 &
                                surv_to_flower == 0,
                              1, surv_to_flower)
    ) %>%
    rename(spatial_unit = !!sym(spatial)) %>%
    mutate(spatial_unit = suppressWarnings(as.numeric(spatial_unit))) %>%
    filter(!is.na(spatial_unit)) %>%
    mutate(
      sample_ID = str_c(Year, Donor, Recipient, Gen,
                        str_pad(spatial_unit, width = 3, pad = "0"),
                        Plant_ID, sep = "_"),
      sample_ID  = str_replace(sample_ID, "^([^_]*)_", "\\1-"),
      segment_id = paste(Block, Recipient, spatial_unit, sep = "_")
    )

  # Deterministic F_plant assignment: where no plant in a segment is flagged
  # as F_plant, take the first surviving sibling. If none survived, take the
  # first row (which will contribute W = 0 to the selection analysis).
  raw <- raw %>%
    group_by(segment_id) %>%
    mutate(
      seg_has_fplant = sum(F_plant) > 0,
      seg_first_surv = which(surv_to_flower == 1)[1],
      seg_pick_idx   = ifelse(!is.na(seg_first_surv), seg_first_surv, 1L),
      F_plant        = ifelse(!seg_has_fplant & row_number() == seg_pick_idx,
                              1L, F_plant)
    ) %>%
    ungroup() %>%
    select(-seg_has_fplant, -seg_first_surv, -seg_pick_idx)

  # Segment-level survival count (number alive out of seeds_sown).
  raw <- raw %>%
    group_by(segment_id) %>%
    mutate(total_surv_to_flower = sum(surv_to_flower, na.rm = TRUE)) %>%
    ungroup()

  fp <- raw %>% filter(F_plant == 1L) %>%
    mutate(
      animal     = sample_ID,
      dam        = Recipient,
      sire       = Donor,
      block      = factor(Block),
      transect_f = factor(spatial_unit),
      corolla    = corolla_diam_mm_SEG,
      SLA        = SLA_SEG,
      d13C       = d13C_SEG,
      msm        = msm_all,
      fec        = est_fecundity,
      surv       = surv_to_flower
    )

  # Centre+scale to mean 0 / sd 1 (z-scaling).
  z <- function(x) {
    m <- mean(x, na.rm = TRUE); s <- sd(x, na.rm = TRUE)
    if (!is.finite(s) || s == 0) return(rep(NA_real_, length(x)))
    (x - m) / s
  }
  fp <- fp %>%
    mutate(
      corolla_z = z(corolla),
      SLA_z     = z(SLA),
      d13C_z    = z(d13C),
      msm_z     = z(msm),
      fec_z     = z(fec)
    )

  list(all = raw, fp = fp, pop = pop_code, spatial = spatial)
}
```

## 3.2 Building the pedigree

`MCMCglmm` needs an inverse-relatedness matrix Aⁱⁿᵛ. We construct it
deterministically from the F-plants: each plant's parents are its `Donor`
(sire) and `Recipient` (dam); founder dams and sires that never appear as
offspring are added as parentless rows so the pedigree closes.

```{r prep-ped-fn}
build_pedigree <- function(fp) {
  ped <- tibble(
    animal = as.character(fp$sample_ID),
    dam    = fp$Recipient,
    sire   = fp$Donor
  )
  missing_parents <- setdiff(unique(c(ped$dam, ped$sire)), ped$animal)
  ped2 <- bind_rows(
    ped,
    tibble(animal = missing_parents,
           dam = NA_character_, sire = NA_character_)
  ) %>% distinct(animal, .keep_all = TRUE)
  ped_ord <- prepPed(as.data.frame(ped2))
  Ainv <- MCMCglmm::inverseA(ped_ord)$Ainv
  list(ped = ped_ord, Ainv = Ainv)
}
```

## 3.3 Run the cleaning across all four populations

```{r load-all}
pop_data <- purrr::map(names(POP_CONFIG),
                       ~ prep_population(.x, POP_CONFIG[[.x]]))
names(pop_data) <- names(POP_CONFIG)
```

## 3.4 Sample sizes

How many sown plants, F-plants, sires, dams, and blocks per population?
This table grounds every effective-sample-size discussion downstream — if
ESS looks low, look here first.

**Table 1.** Sample sizes per population. Columns report total sown
plants, survivors to flowering, F-plants entering the animal model,
number of unique sires (donors), dams (recipients), blocks, and
transects, per-trait observation counts, and the number of complete
cases across all five focal traits (corolla, SLA, δ¹³C, mean seed mass,
fecundity).

```{r sample-sizes}
sample_sizes <- purrr::map_dfr(pop_data, function(d) {
  fp <- d$fp
  tibble(
    Pop          = d$pop,
    n_sown       = nrow(d$all),
    n_survivors  = sum(d$all$surv_to_flower == 1, na.rm = TRUE),
    n_fitness_plants = nrow(fp),
    n_sires      = length(unique(fp$sire)),
    n_dams       = length(unique(fp$dam)),
    n_blocks     = length(unique(fp$block)),
    n_spatial    = length(unique(fp$transect_f)),
    n_corolla    = sum(!is.na(fp$corolla_z)),
    n_SLA        = sum(!is.na(fp$SLA_z)),
    n_d13C       = sum(!is.na(fp$d13C_z)),
    n_msm        = sum(!is.na(fp$msm_z)),
    n_fec        = sum(!is.na(fp$fec_z)),
    n_complete_4traits_fec =
      sum(complete.cases(fp[, c("corolla_z", "SLA_z", "d13C_z",
                                "msm_z", "fec_z")]))
  )
})
sample_sizes
```

## 3.5 Missingness per trait

MCMCglmm imputes missing trait values under the joint multivariate Gaussian
likelihood (conditional on G, V_M, V_block, V_transect, V_R). Traits with
high missingness contribute correspondingly less direct information to G
and its derived quantities. Note especially BB (SLA ~69%, msm ~75%) and
BO (SLA ~49%, msm ~66%) — the BB section below has a dedicated trait-drop
sensitivity check.

**Table 2.** Per-trait missingness by population. `pct_missing` is the
fraction of F-plants with no measurement for that trait. MCMCglmm
imputes these values under the joint multivariate Gaussian likelihood;
high-missingness traits (BB SLA 69%, BB msm 75%, BO msm 66%) therefore
contribute less direct information to **G** than the column counts suggest.

```{r missingness-table}
missingness_tbl <- purrr::imap_dfr(pop_data, function(d, pop) {
  fp <- d$fp
  tibble(
    Pop         = pop,
    trait       = TRAITS,
    n_observed  = sapply(TRAITS, function(t) sum(!is.na(fp[[t]]))),
    n_missing   = sapply(TRAITS, function(t) sum(is.na(fp[[t]]))),
    pct_missing = round(100 * sapply(TRAITS,
                                     function(t) mean(is.na(fp[[t]]))), 1)
  )
})
missingness_tbl
```

## 3.6 Raw trait means (Figure 4 / Table S3)

Before standardising, glance at the raw-scale means. These mirror Figure 4
and Table S3 of the manuscript.

**Table 3.** Raw (un-standardised) trait means ± SE per population, with
per-trait sample size. Units: corolla diameter (mm), SLA (mm²/mg), δ¹³C
(‰), mean seed mass (mg), fecundity (filled seeds per plant).

```{r trait-means-table}
trait_means_tbl <- purrr::imap_dfr(pop_data, function(d, pop) {
  fp <- d$fp
  tibble(
    Pop  = pop,
    trait = c("corolla_diam_mm", "SLA", "d13C", "mean_seed_mass_mg",
              "fecundity"),
    mean = c(mean(fp$corolla, na.rm = TRUE),
             mean(fp$SLA,     na.rm = TRUE),
             mean(fp$d13C,    na.rm = TRUE),
             mean(fp$msm,     na.rm = TRUE),
             mean(fp$fec,     na.rm = TRUE)),
    se   = c(sd(fp$corolla, na.rm = TRUE) / sqrt(sum(!is.na(fp$corolla))),
             sd(fp$SLA,     na.rm = TRUE) / sqrt(sum(!is.na(fp$SLA))),
             sd(fp$d13C,    na.rm = TRUE) / sqrt(sum(!is.na(fp$d13C))),
             sd(fp$msm,     na.rm = TRUE) / sqrt(sum(!is.na(fp$msm))),
             sd(fp$fec,     na.rm = TRUE) / sqrt(sum(!is.na(fp$fec)))),
    n    = c(sum(!is.na(fp$corolla)), sum(!is.na(fp$SLA)),
             sum(!is.na(fp$d13C)),    sum(!is.na(fp$msm)),
             sum(!is.na(fp$fec)))
  )
})
trait_means_tbl
```

```{r trait-means-plot, fig.width=11, fig.height=3.5}
trait_lbl <- c(corolla_diam_mm   = "corolla diameter (mm)",
               SLA               = "SLA (mm^2/mg)",
               d13C              = "δ13C (‰)",
               mean_seed_mass_mg = "mean seed mass (mg)",
               fecundity         = "fecundity (filled seeds)")

# One-way ANOVA + Tukey HSD per trait on raw values, with compact letter
# display. Two populations sharing a letter do not differ at α = 0.05.
trait_raw_long <- purrr::imap_dfr(pop_data, function(d, pop) {
  fp <- d$fp
  tibble(Pop = pop,
         corolla_diam_mm   = fp$corolla,
         SLA               = fp$SLA,
         d13C              = fp$d13C,
         mean_seed_mass_mg = fp$msm,
         fecundity         = fp$fec)
}) %>%
  tidyr::pivot_longer(-Pop, names_to = "trait", values_to = "value") %>%
  filter(!is.na(value))

trait_letters <- trait_raw_long %>%
  group_by(trait) %>%
  group_modify(~ {
    mod   <- aov(value ~ Pop, data = .x)
    pvals <- TukeyHSD(mod)$Pop[, "p adj"]
    lets  <- multcompView::multcompLetters(pvals)$Letters
    # Order letters by population mean (ascending) so "a" is the lowest mean.
    pop_means <- .x %>% group_by(Pop) %>%
      summarise(m = mean(value, na.rm = TRUE), .groups = "drop") %>%
      arrange(m)
    lets_ordered <- lets[pop_means$Pop]
    unique_lets  <- unique(unlist(strsplit(paste(lets_ordered, collapse = ""), "")))
    relabel      <- setNames(letters[seq_along(unique_lets)], unique_lets)
    tibble(Pop = names(lets_ordered),
           letter = vapply(lets_ordered, function(s) {
             paste(relabel[strsplit(s, "")[[1]]], collapse = "")
           }, character(1)))
  }) %>%
  ungroup()

plot_dat <- trait_means_tbl %>%
  left_join(trait_letters, by = c("Pop", "trait")) %>%
  group_by(trait) %>%
  mutate(
    panel_pad = 0.08 * (max(mean + se, na.rm = TRUE) -
                        min(mean - se, na.rm = TRUE)),
    # For δ¹³C the y-axis is reversed, so "above the bar" visually means
    # below in data-coordinates (i.e. *more negative*). Push letters down
    # past mean − se in that panel.
    letter_y  = if (trait[1] == "d13C") mean - se - panel_pad
                else                    mean + se + panel_pad
  ) %>%
  ungroup() %>%
  mutate(trait = factor(recode(trait, !!!trait_lbl),
                        levels = unname(trait_lbl)))

ggplot(plot_dat, aes(Pop, mean, fill = Pop)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.25) +
  geom_text(aes(y = letter_y, label = letter),
            size = 3.5, vjust = 0, fontface = "bold") +
  facet_wrap(~ trait, scales = "free_y", nrow = 1) +
  # Positional list: one entry per facet in the order set by
  # `levels = unname(trait_lbl)` — corolla, SLA, δ¹³C, msm, fec.
  ggh4x::facetted_pos_scales(
    y = list(NULL, NULL, scale_y_reverse(), NULL, NULL)
  ) +
  labs(x = NULL, y = "Mean ± SE") +
  theme_classic(base_size = 11) +
  theme(legend.position = "none",
        strip.text = element_text(size = 9))
```

**Figure 1.** Raw-scale trait means ± SE per population, one panel per
trait. Letters above each bar are compact-letter-display annotations from
trait-wise one-way ANOVAs (`aov(trait ~ Pop)`) followed by Tukey HSD
pairwise comparisons at α = 0.05; populations sharing a letter within a
panel do not differ significantly. Letters are ordered by ascending
population mean within each trait, **except for the δ¹³C panel**, whose
y-axis is reversed so that more-negative values (lower carbon-isotope
discrimination, higher water-use efficiency) appear at the top; the
letter `a` in that panel therefore corresponds to AC (the most-negative
δ¹³C population). Y-axes are panel-specific (free scales). Mirrors Figure
4 / Table S3 of the manuscript.

# 4 · The animal model

## 4.1 Model structure

For each population we fit the same multivariate model:

$$
\mathbf{y} \;=\; \text{trait intercepts} \;+\; \mathbf{a} \;+\; \mathbf{m}
\;+\; \mathbf{b} \;+\; \mathbf{t} \;+\; \mathbf{e}
$$

with the following random-effect structure:

- `us(trait):animal` — additive genetic effects via the pedigree, giving
  the unstructured 5×5 V_A matrix (this is **G**).
- `us(trait):dam` — maternal/dam effects, separating V_A from V_maternal.
- `idh(trait):block` — between-block environmental variance (diagonal).
- `idh(trait):transect_f` — within-block spatial variance (diagonal).
- `us(trait):units` — residual covariance (the 5×5 R matrix).

The diagonal (`idh`) form for block and transect avoids estimating dozens
of off-diagonal nuisance covariances we have no use for.

All five traits — including fecundity — are modelled with a Gaussian
family on the within-population z-scale. Fecundity is biologically a
zero-inflated, right-skewed count, so the Gaussian likelihood is a
convenience choice that will attenuate V_A(fec) and Cov_A(z, fec)
relative to a Poisson or negative-binomial family. We test sensitivity
to this in §13.1 by refitting AC with `log1p(fec)` z-scored as the
fecundity column; a Poisson/NB animal model is left for future work
because it would also require revisiting the joint Gaussian imputation
of missing trait values that MCMCglmm uses for the other four traits.

## 4.2 Priors: parameter-expanded vs inverse-Wishart

Two priors are used:

- **PX (parameter-expanded)** — weakly informative on correlations; this is
  the default for the main analysis. `alpha.V = diag(NT) * 1000` lets the
  scale of V_A and V_M be effectively unbounded. Applied to all four G
  components (animal, dam, block, transect). With only 3 blocks per site
  the inverse-Gamma-like prior used previously (`nu = 0.002`) caused R̂ > 1.2
  for HR block components; the PX form removes that pathology without
  committing to high or low V_block.
- **IW (inverse-Wishart, V = I, ν = NT)** — the original analysis's prior,
  retained as a sensitivity check (see §13.3).

```{r mcmc-fn}
fit_animal_model <- function(prep, peds, prior_type = "px",
                             nitt = NITT, burnin = BURNIN, thin = THIN,
                             n_chains = N_CHAINS) {

  dat  <- as.data.frame(prep$fp)
  Ainv <- peds$Ainv

  dat$animal     <- factor(trimws(as.character(dat$animal)),
                           levels = rownames(Ainv))
  dat$animal     <- droplevels(dat$animal)
  dat$dam        <- factor(dat$dam)
  dat$block      <- factor(dat$block)
  dat$transect_f <- factor(dat$transect_f)

  if (prior_type == "px") {
    # Parameter expansion on every G component, including block and transect.
    # With 3 blocks per site, the inverse-Gamma-like (nu=0.002) prior on
    # V_block previously caused R-hat > 1.2 for HR block components. PX
    # widens the support without committing to a high or low variance,
    # which is what Gelman (2006) recommends for variance components with
    # few random-effect levels.
    prior <- list(
      G = list(
        G1 = list(V = diag(NT), nu = NT, alpha.mu = rep(0, NT),
                  alpha.V = diag(NT) * 1000),     # animal (V_A)
        G2 = list(V = diag(NT), nu = NT, alpha.mu = rep(0, NT),
                  alpha.V = diag(NT) * 1000),     # dam (V_M)
        G3 = list(V = diag(NT), nu = NT, alpha.mu = rep(0, NT),
                  alpha.V = diag(NT) * 1000),     # block (idh, PX)
        G4 = list(V = diag(NT), nu = NT, alpha.mu = rep(0, NT),
                  alpha.V = diag(NT) * 1000)      # transect (idh, PX)
      ),
      R = list(V = diag(NT), nu = NT)
    )
  } else if (prior_type == "iw") {
    prior <- list(
      G = list(
        G1 = list(V = diag(NT), nu = NT),
        G2 = list(V = diag(NT), nu = NT),
        G3 = list(V = diag(NT), nu = 0.002),
        G4 = list(V = diag(NT), nu = 0.002)
      ),
      R = list(V = diag(NT), nu = NT)
    )
  } else stop("unknown prior_type")

  fit_one <- function(seed) {
    set.seed(seed)
    MCMCglmm(
      fixed    = cbind(corolla_z, SLA_z, d13C_z, msm_z, fec_z) ~ trait - 1,
      random   = ~ us(trait):animal + us(trait):dam
                 + idh(trait):block + idh(trait):transect_f,
      rcov     = ~ us(trait):units,
      family   = rep("gaussian", NT),
      ginverse = list(animal = Ainv),
      data     = dat,
      prior    = prior,
      nitt     = nitt, burnin = burnin, thin = thin,
      verbose  = FALSE
    )
  }

  purrr::map(seq_len(n_chains), fit_one)
}
```

## 4.3 Build the per-population pedigrees

```{r build-peds}
peds_list <- purrr::map(pop_data, ~ build_pedigree(.x$fp))
```

## 4.4 Fit the main models (PX prior, all four populations)

```{r fit-main}
fits_main <- cache_fit("fits_main_px", {
  purrr::map2(pop_data, peds_list,
              ~ fit_animal_model(.x, .y, prior_type = "px"))
})
```

## 4.5 Fit sensitivity models with the IW prior

AC is our "reference" population (most complete data). BB has the thinnest
spatial structure (156 Locations with ~1–3 F-plants each) and highest
missingness, making it the population where the PX prior does the most
work. We refit both with the original IW prior so we can compare in §13.3.

```{r fit-iw-sensitivity}
fits_iw_AC <- cache_fit("fits_iw_AC",
  fit_animal_model(pop_data$AC, peds_list$AC, prior_type = "iw"))
fits_iw_BB <- cache_fit("fits_iw_BB",
  fit_animal_model(pop_data$BB, peds_list$BB, prior_type = "iw"))
```

# 5 · MCMC diagnostics

The earlier implementation only checked R̂ on the V_A diagonals (~12% of
monitored parameters). The off-diagonals of G, the maternal covariance
matrix, and the residual covariance matrix all drive downstream quantities
(r_A, β_G, eigenanalysis), and they typically mix worse than diagonals.
Here we diagnose **every** monitored parameter — fixed effects (`Sol`) plus
every entry of every variance/covariance component (`VCV`).

## 5.1 Gelman–Rubin R̂

R̂ > 1.1 is the conventional concern threshold; > 1.2 is bad mixing. With
`TEST_RUN = TRUE`, expect some R̂ above threshold simply because the chains
are short.

**Table 4.** Gelman–Rubin convergence diagnostic (R̂) across all Sol +
VCV parameters per population. `conv_health` summarises the count of
parameters with R̂ > 1.1 and > 1.2; the second printed table lists any
parameter with R̂ > 1.1. Under `TEST_RUN = FALSE` the target is all
parameters with R̂ < 1.1.

```{r convergence-rhat}
gelman_summary <- function(fits, label) {
  mc <- mcmc.list(lapply(fits,
                         function(f) as.mcmc(cbind(f$Sol, f$VCV))))
  gd <- gelman.diag(mc, multivariate = FALSE, autoburnin = FALSE)
  tibble(Pop = label,
         param = rownames(gd$psrf),
         psrf  = gd$psrf[, 1],
         upper = gd$psrf[, 2])
}

conv_tbl <- purrr::map2_dfr(fits_main, names(fits_main),
                            ~ gelman_summary(.x, .y))

conv_health <- conv_tbl %>%
  group_by(Pop) %>%
  summarise(
    n_params         = n(),
    max_psrf         = max(psrf, na.rm = TRUE),
    n_psrf_above_1_1 = sum(psrf > 1.1, na.rm = TRUE),
    n_psrf_above_1_2 = sum(psrf > 1.2, na.rm = TRUE),
    .groups = "drop"
  )
conv_health
conv_tbl %>% filter(psrf > 1.1) %>% arrange(desc(psrf))
```

## 5.2 Effective sample size

ESS < 100 means the chain is essentially uninformative for that parameter;
< 400 is borderline. With `TEST_RUN = FALSE`, the target is ≥ 1000 for V_A
diagonals and the leading r_A entries.

**Table 5.** Effective sample size (ESS) per parameter, summarised
per population. `ess_health` reports the minimum ESS and the count of
parameters with ESS < 100 (uninformative) or < 400 (borderline). The
second printed table lists the 20 lowest-ESS parameters.

```{r convergence-ess}
ess_summary <- function(fits, label) {
  combined <- as.mcmc(do.call(rbind,
                              lapply(fits,
                                     function(f) cbind(f$Sol, f$VCV))))
  ess <- effectiveSize(combined)
  tibble(Pop = label, param = names(ess), ESS = unname(ess))
}
ess_tbl <- purrr::map2_dfr(fits_main, names(fits_main),
                           ~ ess_summary(.x, .y))

ess_health <- ess_tbl %>%
  group_by(Pop) %>%
  summarise(
    n_params        = n(),
    min_ESS         = min(ESS, na.rm = TRUE),
    n_ESS_below_100 = sum(ESS < 100, na.rm = TRUE),
    n_ESS_below_400 = sum(ESS < 400, na.rm = TRUE),
    .groups = "drop"
  )
ess_health
ess_tbl %>% arrange(ESS) %>% head(20)
```

# 6 · Extracting posterior quantities

The fit object is a list of `MCMCglmm` chains. Everything downstream (r_A,
h², eigenanalysis, β_G, evolvability, Random Skewers) is computed from the
VCV columns, so the next few helpers shape those columns into arrays we
can index by trait pair and posterior draw.

## 6.1 Small numerical helpers

`safe_hpd()` returns a 95% HPD interval but tolerates NAs and tiny samples
gracefully. `cov2cor_mat()` converts a covariance matrix to a correlation
matrix without R's `cov2cor` warning machinery.

```{r helpers}
safe_hpd <- function(x, prob = 0.95) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(c(NA_real_, NA_real_))
  as.numeric(HPDinterval(as.mcmc(x), prob = prob))
}

cov2cor_mat <- function(S) {
  d <- sqrt(diag(S))
  S / (d %o% d)
}
```

## 6.2 Reshaping the VCV output into trait × trait arrays

`build_G_array()` takes the VCV matrix (rows = posterior draws, columns =
named like `traitA:traitB.animal`) and returns a 3-D array indexed as
`[draw, trait_i, trait_j]`. The `suffix` argument selects which variance
component to extract — `"animal"` for **G**, `"dam"` for the maternal
covariance, `"units"` for the residual covariance.

```{r vcv-helpers}
combine_VCV <- function(fits) {
  do.call(rbind, lapply(fits, function(f) as.matrix(f$VCV)))
}

build_G_array <- function(VCV, suffix = "animal") {
  Gcol <- function(t1, t2) paste0("trait", t1, ":trait", t2, ".", suffix)
  out <- array(NA_real_, dim = c(nrow(VCV), NT, NT),
               dimnames = list(NULL, TRAITS, TRAITS))
  for (i in seq_along(TRAITS)) for (j in seq_along(TRAITS)) {
    out[, i, j] <- VCV[, Gcol(TRAITS[i], TRAITS[j])]
  }
  out
}

posterior_summary_matrix <- function(arr) {
  mean_mat <- apply(arr, c(2, 3), mean)
  l95 <- apply(arr, c(2, 3), function(x) safe_hpd(x)[1])
  u95 <- apply(arr, c(2, 3), function(x) safe_hpd(x)[2])
  list(mean = mean_mat, l95 = l95, u95 = u95,
       sig  = (l95 > 0) | (u95 < 0))
}

rA_from_G <- function(G_arr) {
  out <- array(NA_real_, dim = dim(G_arr), dimnames = dimnames(G_arr))
  for (k in seq_len(dim(G_arr)[1])) out[k, , ] <- cov2cor_mat(G_arr[k, , ])
  out
}
```

## 6.3 Per-population extraction including h²

`extract_pop()` packages the per-population posterior into one list. h² is
computed *per draw* with the proper phenotypic-variance denominator:

$$
h^2_i \;=\; \frac{V_{A,i}}{V_{A,i} + V_{M,i} + V_{\text{block},i}
+ V_{\text{transect},i} + V_{R,i}}
$$

```{r extract-pop-fn}
extract_pop <- function(fits) {
  VCV <- combine_VCV(fits)
  G_array <- build_G_array(VCV, "animal")
  M_array <- build_G_array(VCV, "dam")
  R_array <- build_G_array(VCV, "units")
  rA_array <- rA_from_G(G_array)

  block_cols    <- grep("\\.block$",      colnames(VCV), value = TRUE)
  transect_cols <- grep("\\.transect_f$", colnames(VCV), value = TRUE)

  h2_array <- array(NA_real_, dim = c(nrow(VCV), NT),
                    dimnames = list(NULL, TRAITS))
  for (i in seq_along(TRAITS)) {
    Va <- G_array[, i, i]
    Vm <- M_array[, i, i]
    Vr <- R_array[, i, i]
    bcol <- grep(paste0("trait", TRAITS[i], "\\.block$"),
                 block_cols,    value = TRUE)
    tcol <- grep(paste0("trait", TRAITS[i], "\\.transect_f$"),
                 transect_cols, value = TRUE)
    Vb   <- if (length(bcol) == 1) VCV[, bcol] else 0
    Vt   <- if (length(tcol) == 1) VCV[, tcol] else 0
    h2_array[, i] <- Va / (Va + Vm + Vb + Vt + Vr)
  }

  list(
    VCV = VCV,
    G_array = G_array,
    M_array = M_array,
    R_array = R_array,
    rA_array = rA_array,
    h2_array = h2_array,
    G_summary  = posterior_summary_matrix(G_array),
    M_summary  = posterior_summary_matrix(M_array),
    rA_summary = posterior_summary_matrix(rA_array)
  )
}
```

## 6.4 Run the extraction on every population

```{r extract-all}
post_main <- purrr::map(fits_main, extract_pop)
```

## 6.5 V_A diagonals

**Table 6.** Additive genetic variance (V_A) diagonals of **G** per
population and trait. `Va_mean` is the posterior mean, `Va_l95` /
`Va_u95` are the 95% HPD bounds. Traits are on the z-scored within-
population scale, so V_A is dimensionless and comparable across traits
and populations.

```{r tbl-Va}
Va_tbl <- purrr::imap_dfr(post_main, function(p, pop) {
  tibble(
    Pop = pop, trait = TRAITS,
    Va_mean = diag(p$G_summary$mean),
    Va_l95  = diag(p$G_summary$l95),
    Va_u95  = diag(p$G_summary$u95)
  )
})
Va_tbl
```

## 6.6 V_M diagonals (maternal/dam variance)

V_M from `us(trait):dam` is the variance among dams — it captures
maternal genetic effects, maternal common-environment effects, and
seed-provisioning effects together, because the design has no grand-dam
replication that could separate them. Compare against V_A in the
previous table: if V_M is large for a trait, the original analysis
(which lacked a dam effect) was almost certainly partitioning some of
that maternal/dam variance into V_A. A large V_M does not, on its own,
imply heritable maternal effects.

**Table 7.** Maternal/dam variance (V_M) diagonals per population and
trait, with 95% HPD intervals. Compare against V_A in Table 6 — large
V_M flags traits where the previous (no-dam) animal model would have
absorbed maternal variance into V_A.

```{r tbl-Vm}
Vm_tbl <- purrr::imap_dfr(post_main, function(p, pop) {
  tibble(
    Pop = pop, trait = TRAITS,
    Vm_mean = diag(p$M_summary$mean),
    Vm_l95  = diag(p$M_summary$l95),
    Vm_u95  = diag(p$M_summary$u95)
  )
})
Vm_tbl
```

## 6.7 Narrow-sense heritability

**Table 8.** Narrow-sense heritability (*h*² = V_A / (V_A + V_M +
V_block + V_transect + V_R)) per population and trait, computed per
posterior draw. `h2_mean` is the posterior mean; `h2_l95` / `h2_u95`
are the 95% HPD bounds.

```{r tbl-h2}
h2_tbl <- purrr::imap_dfr(post_main, function(p, pop) {
  tibble(
    Pop   = pop, trait = TRAITS,
    h2_mean = colMeans(p$h2_array, na.rm = TRUE),
    h2_l95  = apply(p$h2_array, 2, function(x) safe_hpd(x)[1]),
    h2_u95  = apply(p$h2_array, 2, function(x) safe_hpd(x)[2])
  )
})
h2_tbl
```

# 7 · Genetic correlations (r_A)

Per-draw genetic correlations come from converting each posterior G to a
correlation matrix. The upper triangle gives one row per trait pair per
population; the `sig` flag marks pairs whose 95% HPD excludes zero.

## 7.1 Pairwise r_A table

**Table 9.** Pairwise additive genetic correlations (r_A) per
population. One row per upper-triangle trait pair. `rA_mean` is the
posterior mean of the per-draw correlation; `rA_l95` / `rA_u95` give
the 95% HPD interval; `sig = TRUE` flags pairs whose HPD excludes 0.

```{r rA-table}
rA_pairs_tbl <- purrr::imap_dfr(post_main, function(p, pop) {
  rA_arr <- p$rA_array
  ut <- which(upper.tri(diag(NT)), arr.ind = TRUE)
  purrr::map_dfr(seq_len(nrow(ut)), function(idx) {
    i <- ut[idx, 1]; j <- ut[idx, 2]
    x <- rA_arr[, i, j]
    h <- safe_hpd(x)
    tibble(Pop    = pop,
           trait1 = TRAITS[i],
           trait2 = TRAITS[j],
           rA_mean = mean(x, na.rm = TRUE),
           rA_l95  = h[1], rA_u95 = h[2],
           sig     = isTRUE(h[1] > 0) || isTRUE(h[2] < 0))
  })
})
rA_pairs_tbl
```

Same data, formatted as `"mean [l95, u95]"` with **bold** for HPDs that
exclude zero (matches the original Rmd table style):

**Table 10.** Pairwise r_A in wide trait-by-trait format. Each cell
shows `"mean [l95, u95]"`; **bold** entries are correlations whose 95%
HPD excludes zero. Matches the manuscript table layout.

```{r rA-table-formatted}
rA_pairs_tbl %>%
  mutate(
    label = sprintf("%.2f [%.2f, %.2f]", rA_mean, rA_l95, rA_u95),
    label = ifelse(sig, paste0("**", label, "**"), label),
    trait1 = factor(trait1, levels = TRAITS),
    trait2 = factor(trait2, levels = TRAITS)
  ) %>%
  select(Pop, trait1, trait2, label) %>%
  tidyr::pivot_wider(names_from = trait2, values_from = label)
```

## 7.2 r_A heatmap

Cells whose 95% HPD excludes zero are bordered in black and shown at full
saturation; non-significant cells are de-saturated.

```{r rA-plot, fig.width=10, fig.height=8}
trait_labels <- c(corolla_z = "corolla", SLA_z = "SLA",
                  d13C_z = "δ13C", msm_z = "msm", fec_z = "fecundity")

rA_long <- purrr::imap_dfr(post_main, function(p, pop) {
  m   <- p$rA_summary$mean
  l   <- p$rA_summary$l95
  u   <- p$rA_summary$u95
  sig <- p$rA_summary$sig
  expand_grid(trait1 = rownames(m), trait2 = colnames(m)) %>%
    mutate(
      Pop = pop,
      rA  = as.vector(m), l95 = as.vector(l), u95 = as.vector(u),
      sig = as.vector(sig),
      trait1 = factor(recode(trait1, !!!trait_labels),
                      levels = unname(trait_labels)),
      trait2 = factor(recode(trait2, !!!trait_labels),
                      levels = unname(trait_labels))
    )
})

ggplot(rA_long, aes(trait1, trait2, fill = rA)) +
  geom_tile(aes(alpha = ifelse(sig, 1, 0.35)), colour = NA) +
  geom_tile(data = filter(rA_long, sig),
            fill = NA, colour = "black", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%.2f", rA),
                fontface = ifelse(sig, "bold", "plain")), size = 3.2) +
  scale_alpha_identity() +
  scale_fill_gradient2(low = "#3B9AB2", mid = "white", high = "#F21A00",
                       midpoint = 0, limits = c(-1, 1)) +
  coord_equal() +
  facet_wrap(~ Pop) +
  labs(x = NULL, y = NULL, fill = "r_A") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Figure 2.** Pairwise additive genetic correlation (r_A) heatmaps, one
panel per population. Cell colour encodes the posterior mean of r_A
(blue = negative, red = positive). Cells whose 95% HPD excludes zero
are bordered in black and shown at full saturation with bold labels;
non-significant cells are de-saturated.

# 8 · Eigenanalysis of G

The eigendecomposition of G summarises which directions in trait space
hold most additive genetic variance. We work per draw so HPDs reflect
genuine posterior uncertainty.

## 8.1 Per-draw eigendecomposition

Eigenvectors flip sign arbitrarily between draws, so within each
population we sign-align each draw's eigenvector to the corresponding
eigenvector of the posterior-mean G before averaging.

```{r eigen-fn}
eigen_per_draw <- function(G_array) {
  niter <- dim(G_array)[1]
  G_mean <- apply(G_array, c(2, 3), mean)
  ref_vec <- eigen(G_mean, symmetric = TRUE)$vectors

  trace_vec <- numeric(niter)
  eigvals   <- matrix(NA_real_, niter, NT)
  prop_pc1  <- numeric(niter)
  prop_pc2  <- numeric(niter)
  loadings  <- array(NA_real_, dim = c(niter, NT, NT),
                     dimnames = list(NULL, TRAITS, paste0("PC", 1:NT)))
  for (k in seq_len(niter)) {
    Gk <- G_array[k, , ]
    ek <- eigen(Gk, symmetric = TRUE)
    trace_vec[k] <- sum(diag(Gk))
    eigvals[k, ] <- ek$values
    prop_pc1[k]  <- ek$values[1] / sum(ek$values)
    prop_pc2[k]  <- ek$values[2] / sum(ek$values)
    V <- ek$vectors
    for (j in seq_len(NT)) {
      if (sum(V[, j] * ref_vec[, j]) < 0) V[, j] <- -V[, j]
    }
    loadings[k, , ] <- V
  }
  list(trace = trace_vec, eigvals = eigvals,
       prop_pc1 = prop_pc1, prop_pc2 = prop_pc2, loadings = loadings)
}

eig_results <- purrr::map(post_main, ~ eigen_per_draw(.x$G_array))
```

## 8.2 Eigenvalue summary (Table 5 / S9 analogue)

Per population: trace(G), λ₁ … λ₅, and the proportion of variance on PC1
and PC2.

**Table 11.** Eigenvalue summary of **G** per population: posterior
mean of trace(**G**) (total genetic variance), eigenvalues λ₁–λ₅ with
95% HPD intervals (λ₁–λ₃ shown with HPDs; λ₄, λ₅ as means only), and
the proportion of variance on PC1 and PC2 with HPDs. Per-draw
quantities so HPDs reflect posterior uncertainty in **G**.

```{r eigen-summary-table}
eig_summary_tbl <- purrr::imap_dfr(eig_results, function(e, pop) {
  lam_hpd <- apply(e$eigvals, 2, safe_hpd)
  tibble(
    Pop          = pop,
    trace_mean   = mean(e$trace, na.rm = TRUE),
    trace_l95    = safe_hpd(e$trace)[1],
    trace_u95    = safe_hpd(e$trace)[2],
    lambda1_mean = mean(e$eigvals[, 1], na.rm = TRUE),
    lambda1_l95  = lam_hpd[1, 1], lambda1_u95 = lam_hpd[2, 1],
    lambda2_mean = mean(e$eigvals[, 2], na.rm = TRUE),
    lambda2_l95  = lam_hpd[1, 2], lambda2_u95 = lam_hpd[2, 2],
    lambda3_mean = mean(e$eigvals[, 3], na.rm = TRUE),
    lambda3_l95  = lam_hpd[1, 3], lambda3_u95 = lam_hpd[2, 3],
    lambda4_mean = mean(e$eigvals[, 4], na.rm = TRUE),
    lambda5_mean = mean(e$eigvals[, 5], na.rm = TRUE),
    prop_PC1_mean = mean(e$prop_pc1, na.rm = TRUE),
    prop_PC1_l95  = safe_hpd(e$prop_pc1)[1],
    prop_PC1_u95  = safe_hpd(e$prop_pc1)[2],
    prop_PC2_mean = mean(e$prop_pc2, na.rm = TRUE),
    prop_PC2_l95  = safe_hpd(e$prop_pc2)[1],
    prop_PC2_u95  = safe_hpd(e$prop_pc2)[2]
  )
})
eig_summary_tbl
```

## 8.3 PC2 reliability check

PC1 is almost always well separated from PC2 by eigenvalue magnitude, but
PC2 vs PC3 may not be — and when their HPDs overlap, sign-alignment is no
longer enough to stabilise their eigenvectors across draws. The averaged
loadings then pull toward the reference and understate uncertainty. This
chunk flags any population where λ₂ and λ₃ HPDs overlap so PC2 results
can be suppressed or caveated in the manuscript.

**Table 12.** PC2 reliability check per population. `pc2_reliable` is
TRUE if the 95% HPD for λ₂ does not overlap the HPD for λ₃; when FALSE,
PC2 eigenvectors are not stably identified across draws and the
averaged PC2 loadings should not be interpreted.

```{r pc2-reliable}
pc2_reliable_tbl <- eig_summary_tbl %>%
  mutate(
    pc2_reliable = lambda2_l95 > lambda3_u95,
    overlap_note = ifelse(pc2_reliable,
                          "λ₂ and λ₃ HPDs do not overlap — PC2 loadings stable",
                          "λ₂ and λ₃ HPDs OVERLAP — PC2 loadings unreliable")
  ) %>%
  select(Pop, lambda2_mean, lambda2_l95, lambda2_u95,
         lambda3_mean, lambda3_l95, lambda3_u95,
         pc2_reliable, overlap_note)
pc2_reliable_tbl
```

## 8.4 PC1 trait loadings

**Table 13.** PC1 trait loadings per population, with 95% HPD bounds
from per-draw sign-aligned eigenvectors. `main_contributor` flags
traits with |loading| > 0.2.

```{r pc1-loadings}
pc1_loadings_tbl <- purrr::imap_dfr(eig_results, function(e, pop) {
  pc1 <- e$loadings[, , 1]
  hp  <- apply(pc1, 2, safe_hpd)
  tibble(
    Pop   = pop,
    trait = TRAITS,
    loading_mean = colMeans(pc1, na.rm = TRUE),
    loading_l95  = hp[1, ],
    loading_u95  = hp[2, ],
    main_contributor = abs(colMeans(pc1, na.rm = TRUE)) > 0.2
  )
})
pc1_loadings_tbl
```

## 8.5 PC2 trait loadings (caveated for unreliable populations)

**Table 14.** PC2 trait loadings per population, with HPD bounds and
the `pc2_reliable` flag from Table 12. Loadings for populations with
`pc2_reliable = FALSE` are reported but should not be interpreted —
λ₂/λ₃ overlap means PC2 eigenvectors are not stably identified across
draws.

```{r pc2-loadings}
pc2_loadings_tbl <- purrr::imap_dfr(eig_results, function(e, pop) {
  pc2 <- e$loadings[, , 2]
  hp  <- apply(pc2, 2, safe_hpd)
  tibble(
    Pop   = pop,
    trait = TRAITS,
    loading_mean = colMeans(pc2, na.rm = TRUE),
    loading_l95  = hp[1, ],
    loading_u95  = hp[2, ],
    main_contributor = abs(colMeans(pc2, na.rm = TRUE)) > 0.2
  )
}) %>%
  left_join(pc2_reliable_tbl %>% select(Pop, pc2_reliable), by = "Pop")
pc2_loadings_tbl

unreliable_pops <- pc2_reliable_tbl %>%
  filter(!pc2_reliable) %>%
  pull(Pop)
if (length(unreliable_pops) > 0)
  message("PC2 loadings are unreliable for: ",
          paste(unreliable_pops, collapse = ", "),
          " (λ₂ and λ₃ HPDs overlap). ",
          "Suppress or caveat PC2 results for these populations.")
```

## 8.6 Compact contributor strings (manuscript table format)

A trait counts as a "main contributor" to a PC if `|loading| > 0.2`. We
also flag whether the PC is **reliable** at each population: a PC's
loadings are considered reliable only if at least one trait loading
has a 95% HPD that excludes zero. PC1 at AC, BB and BO has wide enough
loading HPDs that no individual loading excludes zero, so the
contributor strings for those populations should be read as posterior
means of an underdetermined eigenvector rather than as resolved trait
contributions. Only HR PC1 (d13C and msm both with HPDs excluding 0)
is loading-reliable. PC2 reliability is also flagged via the λ₂/λ₃
overlap criterion from Table 12 (unreliable at all four populations).

**Table 15.** Compact string summary of PC1 (`pc1_contrib_str`) and
PC2 (`pc2_contrib_str`) main contributors per population, in the
manuscript table format. A trait is listed if |loading| > 0.2 with the
sign of its mean loading. The `pc1_reliable` / `pc2_reliable` columns
flag whether the PC is interpretable at that population
(`pc1_reliable = TRUE` iff ≥ 1 PC1 loading HPD excludes 0;
`pc2_reliable` from Table 12).

```{r pc-contributor-strings}
make_contrib_str <- function(loadings_tbl) {
  loadings_tbl %>%
    filter(main_contributor) %>%
    mutate(sign  = ifelse(loading_mean > 0, "+", "−"),
           label = sprintf("%s (%s)",
                           recode(trait,
                                  corolla_z = "corolla",
                                  SLA_z     = "SLA",
                                  d13C_z    = "δ13C",
                                  msm_z     = "seed mass",
                                  fec_z     = "fecundity"),
                           sign)) %>%
    group_by(Pop) %>%
    summarise(contributors = paste(label, collapse = ", "),
              .groups = "drop")
}

pc1_reliable_tbl <- pc1_loadings_tbl %>%
  group_by(Pop) %>%
  summarise(pc1_reliable = any((loading_l95 > 0) | (loading_u95 < 0),
                               na.rm = TRUE),
            .groups = "drop")

pc1_contrib_str <- make_contrib_str(pc1_loadings_tbl) %>%
  rename(main_PC1_contributors = contributors) %>%
  left_join(pc1_reliable_tbl, by = "Pop")
pc1_contrib_str

pc2_contrib_str <- make_contrib_str(pc2_loadings_tbl) %>%
  rename(main_PC2_contributors = contributors) %>%
  left_join(pc2_reliable_tbl %>% select(Pop, pc2_reliable), by = "Pop")
pc2_contrib_str
```

# 9 · Evolvability metrics

For a unit vector **β** in trait space, the *evolvability* of G in
direction **β** is

$$
e(\boldsymbol{\beta}) \;=\; \boldsymbol{\beta}^{\top} \mathbf{G}
\boldsymbol{\beta}
$$

i.e. the additive genetic variance projected onto **β**. Three summaries
across many random unit vectors are useful:

- **`e_mean`** — average evolvability (mean of e(**β**) over directions).
- **`e_min`** — evolvability in the *most-constrained* direction (the
  worst-case selection axis). Approaches λ_min(**G**) as the number of
  random vectors increases.
- **`e_max`** — evolvability in the *least-constrained* direction (the
  best-case selection axis). Approaches λ_max(**G**).
- **`c_mean`** — conditional evolvability, the variance in **β** that
  remains when all other axes are held fixed. Lower than `e_mean` whenever
  there are genetic constraints.
- **`a_mean`** — autonomy `c / e`. 1 means no constraint; near 0 means
  every direction is strongly constrained by correlations with other traits.

The ratio **`e_max / e_min`** is a scale-free index of multivariate
anisotropy in **G** — a value near 1 means G is roughly spherical (no
preferred direction of evolutionary response); a large ratio means G
is highly anisotropic and selection responses depend strongly on
direction.

We compute these per posterior draw using a fixed set of `N_SKEWERS`
random unit vectors **B**.

## 9.1 Per-draw computation

**Table 16.** Evolvability metrics per population in long format.
`e_mean` = average evolvability β'**G**β over `N_SKEWERS` random unit
vectors; `c_mean` = conditional evolvability (1 / β'**G**⁻¹β); `a_mean`
= autonomy c/e (1 = no constraint, 0 = fully constrained); `e_min`,
`e_max` = range of e(β) across direction vectors. Computed per
posterior draw; `l95`, `u95` are 95% HPD bounds.

```{r evolvability-fn}
rand_unit_vecs <- function(p, nvec) {
  M <- matrix(rnorm(nvec * p), nvec, p)
  M / sqrt(rowSums(M^2))
}

evolv_one <- function(G, B) {
  e <- rowSums((B %*% G) * B)
  cev <- tryCatch({
    Ginv <- solve(G)
    1 / rowSums((B %*% Ginv) * B)
  }, error = function(e) rep(NA_real_, nrow(B)))
  a <- cev / e
  c(e_mean = mean(e, na.rm = TRUE),
    e_min  = min(e,  na.rm = TRUE),
    e_max  = max(e,  na.rm = TRUE),
    c_mean = mean(cev, na.rm = TRUE),
    a_mean = mean(a,   na.rm = TRUE))
}

set.seed(20260520)
B <- rand_unit_vecs(NT, N_SKEWERS)

evol_tbl <- purrr::imap_dfr(post_main, function(p, pop) {
  niter <- dim(p$G_array)[1]
  m <- t(sapply(seq_len(niter),
                function(k) evolv_one(p$G_array[k, , ], B)))
  m <- as.data.frame(m)
  purrr::imap_dfr(m, function(x, metric) {
    h <- safe_hpd(x)
    tibble(Pop = pop, metric = metric,
           mean = mean(x, na.rm = TRUE),
           l95 = h[1], u95 = h[2])
  })
})
evol_tbl
```

### 9.1.1 Wide-format summary (manuscript table)

**Table 16b.** Evolvability metrics per population in wide format with
the anisotropy ratio `e_max / e_min`. Each cell is "posterior mean
[95% HPD lower, upper]".

```{r evolvability-wide-table}
fmt_mh <- function(mean, l95, u95)
  sprintf("%.3f [%.3f, %.3f]", mean, l95, u95)

evol_wide_tbl <- evol_tbl %>%
  mutate(cell = fmt_mh(mean, l95, u95)) %>%
  select(Pop, metric, cell, mean) %>%
  pivot_wider(names_from = metric,
              values_from = c(cell, mean),
              names_glue = "{metric}_{.value}") %>%
  mutate(
    anisotropy_ratio = round(e_max_mean / e_min_mean, 1)
  ) %>%
  select(Pop,
         e_mean = e_mean_cell,
         e_min  = e_min_cell,
         e_max  = e_max_cell,
         c_mean = c_mean_cell,
         a_mean = a_mean_cell,
         anisotropy_ratio)
evol_wide_tbl
```

## 9.2 Figure 7 — three-panel evolvability plot

```{r evolvability-plot, fig.width=12, fig.height=4}
evol_wide <- evol_tbl %>%
  filter(metric %in% c("e_mean", "c_mean", "a_mean")) %>%
  mutate(
    panel = recode(metric,
                   e_mean = "A) Mean Evolvability",
                   c_mean = "B) Conditional Evolvability",
                   a_mean = "C) Autonomy"),
    panel = factor(panel,
                   levels = c("A) Mean Evolvability",
                              "B) Conditional Evolvability",
                              "C) Autonomy")),
    Pop = factor(Pop, levels = c("AC", "BB", "BO", "HR"))
  )

ggplot(evol_wide, aes(Pop, mean)) +
  geom_errorbar(aes(ymin = l95, ymax = u95), width = 0, linewidth = 0.6) +
  geom_point(size = 3) +
  scale_y_continuous(limits = c(0, 1)) +
  facet_wrap(~ panel, nrow = 1) +
  labs(x = "Population", y = NULL) +
  theme_classic(base_size = 13)
```

**Figure 3.** Evolvability metrics by population. (A) Mean evolvability
e(β) = β'**G**β averaged over random unit vectors. (B) Conditional
evolvability c(β) = 1 / β'**G**⁻¹β. (C) Autonomy a = c/e. Points are
posterior means; error bars are 95% HPDs. All three metrics are on the
[0, 1] range under z-scored traits.

# 10 · Population comparison via Random Skewers

The Random Skewers metric `r_RS` measures how similarly two **G**
matrices respond to random selection vectors: r_RS = 1 means identical
response, r_RS = 0 means orthogonal. We propagate posterior uncertainty
by drawing one posterior sample from each of the two populations per
comparison (rather than comparing only the posterior means).

**Power note.** With these data the Random Skewers HPDs span roughly
[0.15, 0.87] for every pairwise comparison (Table 17), so the metric
cannot distinguish "similar **G**" from "weakly correlated **G**" — both
are inside every HPD. The point estimates (r_RS ≈ 0.50–0.63) and the
permutation p-values from `evolqg::RandomSkewers()` (p_mean ≈ 0.10–0.17,
max ≈ 0.50) provide no statistical evidence that any pair of **G**
matrices is more or less similar than any other pair. We report the
metric for completeness but do not draw conclusions about
**G**-matrix divergence from these comparisons.

**Table 17.** Random Skewers comparison of **G** matrices between all
pairs of populations. `r_RS_mean` is the posterior mean of the response
correlation across `N_POST_DRAWS_FOR_SKEWERS` paired posterior draws,
each evaluated against `N_SKEWERS` random unit vectors. `r_RS_l95` /
`r_RS_u95` are 95% HPD bounds; `p_mean` / `p_max` are the mean and
maximum across-draw permutation p-values from
`evolqg::RandomSkewers()`.

```{r random-skewers}
# evolqg::RandomSkewers() returns a named numeric c(correlation, probability)
# when called on a pair of matrices — NOT c(correlation, pvalue). The earlier
# version pulled res["pvalue"], which is always NA, so the rs_tbl p-value
# columns came out as NaN / -Inf. Fixed below.
posterior_skewers <- function(G_array_1, G_array_2,
                              n_draws = N_POST_DRAWS_FOR_SKEWERS,
                              n_skewers = N_SKEWERS) {
  i1 <- sample(seq_len(dim(G_array_1)[1]), n_draws, replace = TRUE)
  i2 <- sample(seq_len(dim(G_array_2)[1]), n_draws, replace = TRUE)
  out <- vapply(seq_len(n_draws), function(k) {
    res <- tryCatch(
      RandomSkewers(G_array_1[i1[k], , ], G_array_2[i2[k], , ],
                    num.vectors = n_skewers),
      error = function(e) c(correlation = NA_real_, probability = NA_real_))
    c(correlation = unname(res["correlation"]),
      probability = unname(res["probability"]))
  }, numeric(2))
  t(out)
}

pop_names <- names(post_main)
pairs <- combn(pop_names, 2, simplify = FALSE)
rs_tbl <- purrr::map_dfr(pairs, function(pp) {
  rs <- posterior_skewers(post_main[[pp[1]]]$G_array,
                          post_main[[pp[2]]]$G_array)
  r <- rs[, "correlation"]
  p <- rs[, "probability"]
  h <- safe_hpd(r)
  tibble(pop1 = pp[1], pop2 = pp[2],
         r_RS_mean = mean(r, na.rm = TRUE),
         r_RS_sd   = sd(r,   na.rm = TRUE),
         r_RS_l95  = h[1], r_RS_u95 = h[2],
         p_mean    = mean(p, na.rm = TRUE),
         p_max     = suppressWarnings(max(p, na.rm = TRUE)))
})
rs_tbl
```

# 11 · Selection — phenotypic vs genetic gradients

The central added analysis: comparing the **phenotypic** selection
gradient β_P (the Lande–Arnold OLS regression) against the **genetic**
selection gradient β_G computed from **G** via the Robertson–Price
identity.

## 11.1 Lande–Arnold β_P (phenotypic)

Lande–Arnold's β_P is just the OLS regression of relative fitness on
standardised traits. Two adjustments to the textbook version are worth
calling out:

- **Relative-fitness denominator**: `w_rel = W / mean(W)`. Earlier
  iterations used `mean(W[W > 0 | surv == 1])`, which made `mean(w_rel)`
  not equal to 1 and gave an asymmetric estimator.
- **HC3 sandwich SEs**: residual variance in `w_rel` scales with fitness
  magnitude (a zero-inflated outcome guarantees heteroscedasticity), so
  the conventional OLS SEs are wrong. We use HC3 sandwich SEs and a
  HC3-based Wald test for the global F.

```{r beta-pheno-fn}
beta_pheno <- function(prep, total_fitness = FALSE) {
  d <- prep$fp %>%
    mutate(W = if (total_fitness) {
                  ifelse(surv == 1, ifelse(is.na(fec), 0, fec), 0)
              } else {
                  ifelse(is.na(fec), 0, fec)
              }) %>%
    select(corolla_z, SLA_z, d13C_z, msm_z, W) %>%
    drop_na()
  if (nrow(d) < 10) return(NULL)
  d$w_rel <- d$W / mean(d$W, na.rm = TRUE)
  mod <- lm(w_rel ~ corolla_z + SLA_z + d13C_z + msm_z, data = d)
  vifs <- tryCatch(car::vif(mod), error = function(e) rep(NA_real_, 4))
  vif_df <- tibble(term = names(vifs), vif = unname(vifs))
  V_HC3 <- sandwich::vcovHC(mod, type = "HC3")
  ct    <- lmtest::coeftest(mod, vcov. = V_HC3)
  wt    <- lmtest::waldtest(mod, vcov = V_HC3)
  q     <- qt(0.975, df.residual(mod))
  coefs <- tibble(
    term      = rownames(ct),
    estimate  = ct[, "Estimate"],
    std.error = ct[, "Std. Error"],
    statistic = ct[, "t value"],
    p.value   = ct[, "Pr(>|t|)"],
    conf.low  = ct[, "Estimate"] - q * ct[, "Std. Error"],
    conf.high = ct[, "Estimate"] + q * ct[, "Std. Error"],
    n         = nrow(d)
  ) %>%
    filter(term != "(Intercept)") %>%
    left_join(vif_df, by = "term")
  g <- broom::glance(mod) %>%
    select(r.squared, adj.r.squared, df, df.residual, nobs) %>%
    mutate(F_HC3 = wt$F[2], p_HC3 = wt$`Pr(>F)`[2])
  attr(coefs, "model_fit") <- g
  coefs
}
```

### 11.1.1 Fecundity-only β_P (the classic Lande–Arnold)

**Table 18.** Phenotypic selection gradients β_P (fecundity-only
relative fitness) per population and trait, from OLS regression of
`w_rel = fec / mean(fec)` on z-scored corolla, SLA, δ¹³C, and msm
across complete-case rows. `std.error`, `statistic`, `p.value`, and
`conf.low`/`conf.high` use HC3 sandwich SEs; `n` is the per-population
complete-case sample size; `vif` is the variance inflation factor for
each predictor.

```{r beta-P-fec}
beta_P_tbl <- purrr::imap_dfr(pop_data, function(prep, pop) {
  beta_pheno(prep, total_fitness = FALSE) %>%
    mutate(Pop = pop, fitness_def = "fec_only_flowering", .before = 1)
})
beta_P_tbl
```

### 11.1.2 Total-fitness β_P (survival × fecundity)

In this dataset non-survivors don't have trait measurements, so the
fec-only and survival × fec regression samples are nearly identical. The
total-fitness column matters more in datasets where some non-survivors
were measured pre-mortality.

**Table 19.** Phenotypic selection gradients β_P using total fitness
(survival × fecundity, with non-survivors = 0) as the fitness
denominator. `beta_P_all` combines this table with Table 19 for direct
side-by-side comparison with Table 18. In this dataset the fec-only
and total-fitness samples nearly coincide because non-survivors have
no trait data.

```{r beta-P-total}
beta_P_total <- purrr::imap_dfr(pop_data, function(prep, pop) {
  beta_pheno(prep, total_fitness = TRUE) %>%
    mutate(Pop = pop, fitness_def = "surv_x_fec", .before = 1)
})
beta_P_all <- bind_rows(beta_P_tbl, beta_P_total)
beta_P_all
```

### 11.1.3 Fit summary

R², F (HC3), p (HC3), n, and max VIF for the fecundity-only regressions.
This is what the manuscript text "R² ≈ 0.10, all VIF < 1.1" comes from.

**Table 20.** Lande–Arnold β_P fit summary per population for the
fecundity-only regression: R² and adjusted R², HC3-based global F and
its p-value, residual degrees of freedom, complete-case n, and the
maximum VIF across the four trait predictors.

```{r beta-P-fit-summary}
beta_P_fit_summary <- purrr::imap_dfr(pop_data, function(prep, pop) {
  coefs <- beta_pheno(prep, total_fitness = FALSE)
  if (is.null(coefs)) return(NULL)
  g <- attr(coefs, "model_fit")
  tibble(Pop = pop,
         r_squared     = g$r.squared,
         adj_r_squared = g$adj.r.squared,
         F_HC3         = g$F_HC3,
         p_HC3         = g$p_HC3,
         df_resid      = g$df.residual,
         n             = g$nobs,
         max_vif       = max(coefs$vif, na.rm = TRUE))
})
beta_P_fit_summary
```

## 11.2 Robertson–Price β_G (genetic)

For non-fitness traits **z** and fitness trait *w*,

$$
\boldsymbol{\beta}_G \;=\; \mathbf{G}_{zz}^{-1}\,\mathrm{Cov}_A(z, w)
$$

computed draw by draw to propagate posterior uncertainty.

**Scale and likelihood notes.**

1. *Scale (rescaled to the Lande w_rel form).* The animal model fits G
   on z-scored fecundity (`fec_z`), so the off-diagonal column of G
   directly yields Cov_A(z, z_fec). Lande's equation Δz̄ = G β requires
   β = G_zz⁻¹·Cov_A(z, w_rel), where w_rel = W / W̄. Because
   w_rel = 1 + CV(W) · z_fec (with CV(W) = SD(W) / mean(W) on the raw
   scale), Cov_A(z, w_rel) = CV(W) · Cov_A(z, z_fec). We therefore
   multiply each per-draw β_G by CV(fec) per population — a deterministic
   rescaling, so HPD widths scale proportionally. β_G is now on the
   relative-fitness scale and directly comparable to the Lande–Arnold
   β_P.

2. *Gaussian likelihood on count data*: fecundity is a non-negative,
   right-skewed integer. The Gaussian likelihood attenuates V_A(fec) and
   Cov_A(z, fec) relative to a Poisson or negative-binomial likelihood,
   and β_G with it. The `log1p(fec)` sensitivity check in §13.1 gauges
   whether this attenuation matters in practice.

**Table 21.** Genetic selection gradients β_G per population and trait,
from the Robertson–Price identity β_G = **G**_zz⁻¹ Cov_A(z, w_rel)
computed draw by draw and rescaled to the relative-fitness scale by
multiplying each draw by `cv_fec = SD(fec)/mean(fec)` per population
(see §11.2). `beta_G_median` is the posterior median (robust to
heavy-tailed posteriors); `beta_G_mean` is reported only for diagnostic
comparison. `beta_G_l95`/`beta_G_u95` are 95% HPD bounds.
`n_kept`/`n_dropped` count posterior draws retained vs excluded because
**G**_zz was numerically near-singular (`rcond < 1e-10`); in practice
`n_dropped` is ~0 because the threshold is strict, and the heavy tails
that matter for the mean/median gap come from finite but
ill-conditioned draws rather than from singular ones. `sig = TRUE`
flags HPDs that exclude zero.

```{r beta-genetic-fn}
beta_genetic_posterior <- function(G_array, fec_index = 5,
                                   cv_factor = 1,
                                   rcond_threshold = 1e-10) {
  # cv_factor: SD(fitness)/mean(fitness) on the raw scale. Converts
  # G_zz^{-1} Cov_A(z, z_fec) → G_zz^{-1} Cov_A(z, w_rel), since
  # w_rel = 1 + cv_factor * z_fec implies Cov_A(z, w_rel) = cv_factor *
  # Cov_A(z, z_fec). Set cv_factor = 1 to retain the z-scale output.
  z_idx <- setdiff(seq_len(NT), fec_index)
  niter <- dim(G_array)[1]
  out   <- matrix(NA_real_, nrow = niter, ncol = length(z_idx),
                  dimnames = list(NULL, TRAITS[z_idx]))
  n_dropped_rcond <- 0L
  n_dropped_solve <- 0L
  for (k in seq_len(niter)) {
    Gk    <- G_array[k, , ]
    G_zz  <- Gk[z_idx, z_idx]
    G_zw  <- Gk[z_idx, fec_index]
    if (!is.finite(rcond(G_zz)) || rcond(G_zz) < rcond_threshold) {
      n_dropped_rcond <- n_dropped_rcond + 1L
      next
    }
    out[k, ] <- tryCatch(
      cv_factor * as.numeric(solve(G_zz, G_zw)),
      error = function(e) {
        n_dropped_solve <<- n_dropped_solve + 1L
        rep(NA_real_, length(z_idx))
      }
    )
  }
  attr(out, "n_total")         <- niter
  attr(out, "n_dropped_rcond") <- n_dropped_rcond
  attr(out, "n_dropped_solve") <- n_dropped_solve
  attr(out, "n_kept")          <- niter - n_dropped_rcond - n_dropped_solve
  attr(out, "cv_factor")       <- cv_factor
  out
}

beta_G_tbl <- purrr::imap_dfr(post_main, function(p, pop) {
  fec_raw <- pop_data[[pop]]$fp$fec
  cv_fec  <- sd(fec_raw, na.rm = TRUE) / mean(fec_raw, na.rm = TRUE)
  bG <- beta_genetic_posterior(p$G_array,
                               fec_index = which(TRAITS == "fec_z"),
                               cv_factor = cv_fec)
  hpds <- apply(bG, 2, safe_hpd)
  tibble(
    Pop            = pop,
    trait          = colnames(bG),
    cv_fec         = cv_fec,
    beta_G_median  = apply(bG, 2, median, na.rm = TRUE),
    beta_G_mean    = colMeans(bG, na.rm = TRUE),
    beta_G_l95     = hpds[1, ],
    beta_G_u95     = hpds[2, ],
    n_kept         = attr(bG, "n_kept"),
    n_dropped      = attr(bG, "n_total") - attr(bG, "n_kept"),
    sig            = (hpds[1, ] > 0) | (hpds[2, ] < 0)
  )
})
beta_G_tbl
```

`cv_fec` shows the per-population rescaling factor — useful as a sanity
check, since β_G HPD widths scale linearly with it. `beta_G_median`
summarises the posterior because its distribution is heavy-tailed even
when **G**_zz is numerically non-singular: the `rcond(G_zz) < 1e-10`
guard rarely fires (see `n_dropped` ≈ 0 in practice), but moderately
ill-conditioned draws give large but finite values that pull the
posterior mean far from its central mass. The median is robust to these
tails; `beta_G_mean` is reported only for diagnostic comparison.

## 11.3 Side-by-side comparison plot

Filled symbols mark gradients whose 95% CI / HPD excludes zero.

```{r beta-comparison-plot, fig.width=11, fig.height=5}
# Per-population n from β_P (n is constant across traits within a pop after
# listwise deletion). Used to annotate the y-axis so BB's n=36 is visible.
n_by_pop <- beta_P_tbl %>%
  group_by(Pop) %>%
  summarise(n = max(n, na.rm = TRUE), .groups = "drop") %>%
  mutate(pop_label = paste0(Pop, " (n=", n, ")"))

plot_dat <- bind_rows(
  beta_P_tbl %>%
    transmute(Pop, trait = term,
              mean = estimate, l95 = conf.low, u95 = conf.high,
              type = "Phenotypic β (fecundity)"),
  beta_G_tbl %>%
    transmute(Pop, trait,
              mean = beta_G_median, l95 = beta_G_l95, u95 = beta_G_u95,
              type = "Genetic β_G")
) %>%
  left_join(n_by_pop %>% select(Pop, pop_label), by = "Pop") %>%
  mutate(
    sig = (l95 > 0) | (u95 < 0),
    trait = recode_factor(trait,
              corolla_z = "corolla", SLA_z = "SLA",
              d13C_z    = "δ13C",    msm_z = "msm")
  )

ggplot(plot_dat, aes(mean, pop_label, colour = type, shape = sig)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
  geom_errorbarh(aes(xmin = l95, xmax = u95),
                 height = 0.2, position = position_dodge(width = 0.5)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  facet_wrap(~ trait, scales = "free_x", nrow = 1) +
  scale_shape_manual(values = c("FALSE" = 1, "TRUE" = 16),
                     labels = c("NS", "HPD excludes 0")) +
  labs(x = "Standardised selection gradient",
       y = NULL, colour = NULL, shape = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")
```

**Figure 4.** Side-by-side comparison of phenotypic (β_P) and genetic
(β_G) selection gradients on z-scored traits, one panel per non-fitness
trait (corolla, SLA, δ¹³C, msm). Points are posterior medians (β_G) or
OLS estimates (β_P); error bars are 95% HPD intervals (β_G) or HC3 95%
CIs (β_P). Filled symbols mark gradients whose interval excludes zero;
open symbols are non-significant. Y-axis labels include the
complete-case n per population.

# 12 · Sibship-level associations between focal-plant traits and segment survival

The phenotypic β in §11 only sees plants that survived to flowering, so
it can't measure viability selection on the missing dead plants. Trait
data are only available on the F-plant (the designated focal sib, or
the first surviving substitute), so we cannot fit an individual-level
viability gradient. Instead we fit a segment-level binomial GLM
relating the F-plant's traits to the fraction of its segment-mates that
survived. **This is a sibship-level association, not a Lande viability
gradient**: a positive coefficient means segments whose surviving focal
sib happened to have a high trait value also had higher overall
survival, which can arise from family-mean trait differences correlated
with family-mean survival rather than from individual selection on that
trait. We report it as a descriptive complement to §11 and discuss the
distinction in §14.6. Coefficients should not be compared to published
β_v values from individual-level data.

## 12.1 Safety: cap survival counts at seeds sown

Survival counts come from `group_by(segment_id) %>% sum(surv_to_flower)`.
If, for any reason, the count ever exceeds `seeds_sown` (8 or 10
depending on population), something upstream is broken — most likely the
segment key. We cap silently but **warn loudly** so problems can't pass
unnoticed.

```{r cap-seeds-fn}
cap_seeds <- function(df, seeds_sown_val, pop_label) {
  n_over <- sum(df$total_surv_to_flower > seeds_sown_val, na.rm = TRUE)
  if (n_over > 0)
    warning(pop_label, ": ", n_over,
            " row(s) had total_surv_to_flower > seeds_sown and were capped")
  df$total_surv_to_flower <- pmin(df$total_surv_to_flower, seeds_sown_val)
  df
}
```

## 12.2 Segment-level survival GLM (sibship-level associations)

Quasi-binomial because survival counts are overdispersed (see §12.3).
The predictor side uses only the four non-fitness traits. Coefficients
are **trait-of-surviving-focal-sib × segment-survival-fraction**
associations, not individual viability gradients.

**Table 22.** Segment-level survival GLM per population: quasi-binomial
regression of `cbind(total_surv_to_flower, seeds_sown −
total_surv_to_flower)` on z-scored corolla, SLA, δ¹³C, and msm.
Columns from `broom::tidy()` give the coefficient estimate, SE,
z-statistic, p-value, and 95% CI. Trait values come from the surviving
focal sib, so these are sibship-level associations rather than
individual Lande viability gradients (see §14.6). Quasi-binomial family
used because of overdispersion documented in Table 23.

```{r surv-glm}
surv_summary <- purrr::imap_dfr(pop_data, function(prep, pop) {
  cfg <- POP_CONFIG[[pop]]
  d <- cap_seeds(prep$fp, cfg$seeds_sown, pop) %>%
    mutate(seeds_sown = cfg$seeds_sown) %>%
    select(total_surv_to_flower, seeds_sown,
           corolla_z, SLA_z, d13C_z, msm_z) %>%
    drop_na()
  if (nrow(d) < 10) return(NULL)
  mod <- glm(cbind(total_surv_to_flower, seeds_sown - total_surv_to_flower) ~
               corolla_z + SLA_z + d13C_z + msm_z,
             family = quasibinomial, data = d)
  broom::tidy(mod, conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    mutate(Pop = pop, .before = 1)
})
surv_summary
```

## 12.3 Overdispersion diagnostic

Pearson-residual dispersion (Σ rᵖ² / df) — values >> 1 justify the
quasi-binomial choice. The original analysis used quasi-binomial for BO
and HR for precisely this reason.

**Table 23.** Survival GLM overdispersion diagnostic per population:
the Pearson-residual dispersion statistic Σ rᵖ² / df. Values
substantially > 1 indicate binomial overdispersion and justify the
quasi-binomial family used in Table 22.

```{r surv-overdisp}
overdisp_fun <- function(model) {
  rp <- residuals(model, type = "pearson")
  sum(rp^2) / df.residual(model)
}

overdisp_tbl <- purrr::imap_dfr(pop_data, function(prep, pop) {
  cfg <- POP_CONFIG[[pop]]
  d <- cap_seeds(prep$fp, cfg$seeds_sown, pop) %>%
    mutate(seeds_sown = cfg$seeds_sown) %>%
    select(total_surv_to_flower, seeds_sown,
           corolla_z, SLA_z, d13C_z, msm_z) %>%
    drop_na()
  if (nrow(d) < 10) return(NULL)
  mod_binom <- glm(
    cbind(total_surv_to_flower, seeds_sown - total_surv_to_flower) ~
      corolla_z + SLA_z + d13C_z + msm_z,
    family = binomial, data = d)
  tibble(Pop = pop,
         overdispersion_ratio = overdisp_fun(mod_binom),
         family_used = "quasibinomial")
})
overdisp_tbl
```

# 13 · Sensitivity analyses

Three checks, in increasing breadth: (1) likelihood mis-specification on
fecundity (AC only); (2) trait drop / imputation reliance (BB only, where
missingness is highest); (3) prior mis-specification (AC and BB, the
most-complete and the most-vulnerable populations).

## 13.1 log1p(fec) sensitivity for β_G — likelihood specification

Refit AC with `log1p(fec)` z-scored as the fecundity column. Each
spec's β_G is rescaled to its own relative-fitness scale (CV of fec for
the raw spec, CV of log1p(fec) for the log spec) so both rows are
directly comparable as Lande gradients. If β_G is stable across the
two specs the Gaussian-on-raw-counts approximation is defensible; if it
shifts materially, prefer the log1p specification (or move to a Poisson
family in MCMCglmm).

**Table 24.** β_G sensitivity to fecundity likelihood specification at
AC, both rows on the Lande relative-fitness scale. The main analysis
(`spec = "raw fec (main)"`) uses `fec_z` from raw counts rescaled by
CV(fec); the refit (`spec = "log1p(fec) sensitivity"`) uses `fec_z`
from `log1p(fec)` rescaled by CV(log1p(fec)). Posterior medians
reported. Stable β_G across specs supports the Gaussian-on-raw-counts
approximation; large shifts would argue for a log or Poisson
likelihood.

```{r beta-genetic-fec-sensitivity}
prep_AC_logfec <- pop_data$AC
prep_AC_logfec$fp <- prep_AC_logfec$fp %>%
  mutate(fec_z = as.numeric(scale(log1p(fec))))

fits_logfec_AC <- cache_fit("fits_logfec_AC",
  fit_animal_model(prep_AC_logfec, peds_list$AC, prior_type = "px"))
post_logfec_AC <- extract_pop(fits_logfec_AC)

cv_log_AC <- with(prep_AC_logfec$fp,
                  sd(log1p(fec), na.rm = TRUE) /
                  mean(log1p(fec), na.rm = TRUE))

bG_logfec_AC <- beta_genetic_posterior(post_logfec_AC$G_array,
                                       fec_index = which(TRAITS == "fec_z"),
                                       cv_factor = cv_log_AC)
hpd_logfec <- apply(bG_logfec_AC, 2, safe_hpd)

beta_G_fec_compare <- bind_rows(
  beta_G_tbl %>%
    filter(Pop == "AC") %>%
    transmute(trait,
              beta_G = beta_G_median, l95 = beta_G_l95, u95 = beta_G_u95,
              spec   = "raw fec (main)"),
  tibble(trait  = colnames(bG_logfec_AC),
         beta_G = apply(bG_logfec_AC, 2, median, na.rm = TRUE),
         l95    = hpd_logfec[1, ], u95 = hpd_logfec[2, ],
         spec   = "log1p(fec) sensitivity")
)
beta_G_fec_compare
```

## 13.2 BB trait-drop sensitivity — imputation reliance

BB has the highest missingness (SLA 69%, msm 75%). Its five-trait r_A
estimates depend partly on MCMCglmm's multivariate-normal imputation
rather than directly observed data. We refit BB twice — once dropping SLA,
once dropping msm — and compare the surviving r_A entries against the
five-trait posterior. Stable estimates ⇒ data-driven; large shifts or
sign flips ⇒ imputation-driven.

### 13.2.1 Subset-trait refit function

A trimmed version of `fit_animal_model` that takes a `traits_keep`
argument and rebuilds the prior at the matching dimension.

```{r bb-subset-fn}
fit_animal_model_subset <- function(prep, peds, traits_keep,
                                    nitt = NITT, burnin = BURNIN, thin = THIN,
                                    n_chains = N_CHAINS) {
  nt_k <- length(traits_keep)
  dat  <- as.data.frame(prep$fp)
  Ainv <- peds$Ainv
  dat$animal <- factor(trimws(as.character(dat$animal)),
                       levels = rownames(Ainv))
  dat$animal     <- droplevels(dat$animal)
  dat$dam        <- factor(dat$dam)
  dat$block      <- factor(dat$block)
  dat$transect_f <- factor(dat$transect_f)

  prior <- list(
    G = list(
      G1 = list(V = diag(nt_k), nu = nt_k, alpha.mu = rep(0, nt_k),
                alpha.V = diag(nt_k) * 1000),  # animal
      G2 = list(V = diag(nt_k), nu = nt_k, alpha.mu = rep(0, nt_k),
                alpha.V = diag(nt_k) * 1000),  # dam
      G3 = list(V = diag(nt_k), nu = nt_k, alpha.mu = rep(0, nt_k),
                alpha.V = diag(nt_k) * 1000),  # block (PX, matches main)
      G4 = list(V = diag(nt_k), nu = nt_k, alpha.mu = rep(0, nt_k),
                alpha.V = diag(nt_k) * 1000)   # transect (PX, matches main)
    ),
    R = list(V = diag(nt_k), nu = nt_k)
  )

  lhs     <- paste0("cbind(", paste(traits_keep, collapse = ", "), ")")
  fixed_f <- as.formula(paste(lhs, "~ trait - 1"))

  fit_one <- function(seed) {
    set.seed(seed)
    MCMCglmm(
      fixed    = fixed_f,
      random   = ~ us(trait):animal + us(trait):dam
                 + idh(trait):block + idh(trait):transect_f,
      rcov     = ~ us(trait):units,
      family   = rep("gaussian", nt_k),
      ginverse = list(animal = Ainv),
      data     = dat,
      prior    = prior,
      nitt     = nitt, burnin = burnin, thin = thin,
      verbose  = FALSE
    )
  }
  purrr::map(seq_len(n_chains), fit_one)
}

extract_rA_subset <- function(fits, traits_keep) {
  VCV  <- do.call(rbind, lapply(fits, function(f) as.matrix(f$VCV)))
  nt_k <- length(traits_keep)
  G_arr <- array(NA_real_, dim = c(nrow(VCV), nt_k, nt_k),
                 dimnames = list(NULL, traits_keep, traits_keep))
  for (i in seq_along(traits_keep)) for (j in seq_along(traits_keep)) {
    G_arr[, i, j] <- VCV[, paste0("trait", traits_keep[i], ":trait",
                                  traits_keep[j], ".animal")]
  }
  rA_arr <- array(NA_real_, dim = dim(G_arr), dimnames = dimnames(G_arr))
  for (k in seq_len(dim(G_arr)[1])) rA_arr[k, , ] <- cov2cor_mat(G_arr[k, , ])
  rA_arr
}
```

### 13.2.2 Run BB without SLA and without msm

```{r bb-subset-fits}
traits_no_SLA <- setdiff(TRAITS, "SLA_z")
traits_no_msm <- setdiff(TRAITS, "msm_z")

fits_BB_no_SLA <- cache_fit("fits_BB_no_SLA",
  fit_animal_model_subset(pop_data$BB, peds_list$BB, traits_no_SLA))
fits_BB_no_msm <- cache_fit("fits_BB_no_msm",
  fit_animal_model_subset(pop_data$BB, peds_list$BB, traits_no_msm))

rA_BB_main   <- post_main$BB$rA_array
rA_BB_no_SLA <- extract_rA_subset(fits_BB_no_SLA, traits_no_SLA)
rA_BB_no_msm <- extract_rA_subset(fits_BB_no_msm, traits_no_msm)
```

### 13.2.3 Compare overlapping r_A entries

`delta` is the per-pair shift in posterior mean (subset − main). A
well-conditioned five-trait fit should show `max_abs_delta < 0.1`,
`n_HPD_nonoverlap = 0`, and no sign changes.

**Table 25.** BB trait-drop sensitivity for genetic correlations.
`rA_BB_sensitivity` (first printed table) gives per-pair r_A under the
main five-trait fit vs. the drop-SLA and drop-msm refits, with HPD
bounds and `delta = rA_sub − rA_main`. `rA_BB_sens_summary` (second
table) reports per-spec totals: number of overlapping pairs, max and
median |Δ|, HPD-non-overlap count, and sign-change count.

```{r bb-subset-compare}
compare_rA <- function(rA_main, rA_sub, traits_sub, label) {
  ut <- which(upper.tri(diag(length(traits_sub))), arr.ind = TRUE)
  purrr::map_dfr(seq_len(nrow(ut)), function(idx) {
    i  <- ut[idx, 1]; j <- ut[idx, 2]
    t1 <- traits_sub[i]; t2 <- traits_sub[j]
    x_main <- rA_main[, t1, t2]
    x_sub  <- rA_sub[, i, j]
    h_main <- safe_hpd(x_main); h_sub <- safe_hpd(x_sub)
    tibble(spec        = label,
           trait1      = t1, trait2 = t2,
           rA_main     = mean(x_main, na.rm = TRUE),
           rA_main_l95 = h_main[1], rA_main_u95 = h_main[2],
           rA_sub      = mean(x_sub,  na.rm = TRUE),
           rA_sub_l95  = h_sub[1],   rA_sub_u95  = h_sub[2],
           delta       = mean(x_sub, na.rm = TRUE) -
                         mean(x_main, na.rm = TRUE))
  })
}

rA_BB_sensitivity <- bind_rows(
  compare_rA(rA_BB_main, rA_BB_no_SLA, traits_no_SLA, "drop SLA"),
  compare_rA(rA_BB_main, rA_BB_no_msm, traits_no_msm, "drop msm")
)
rA_BB_sensitivity

rA_BB_sens_summary <- rA_BB_sensitivity %>%
  group_by(spec) %>%
  summarise(
    n_pairs          = n(),
    max_abs_delta    = max(abs(delta), na.rm = TRUE),
    median_abs_delta = median(abs(delta), na.rm = TRUE),
    n_HPD_nonoverlap = sum(rA_main_u95 < rA_sub_l95
                           | rA_sub_u95 < rA_main_l95, na.rm = TRUE),
    n_sign_change    = sum(sign(rA_main) != sign(rA_sub), na.rm = TRUE),
    .groups = "drop"
  )
rA_BB_sens_summary
```

## 13.3 Prior sensitivity (PX vs IW) — AC and BB

Three sets of quantities are compared between the PX and IW priors:

- **r_A** (genetic correlations) — the original check.
- **V_A diagonals** — prior sensitivity here would indicate that
  heritability estimates are prior-dominated.
- **V_M diagonals** — with only ~3 full-sibs per dam, V_M is weakly
  identified and the most likely place to see a prior-driven shift.

### 13.3.1 Extraction helper covering all three quantities

```{r prior-sensitivity-fn}
extract_prior_sensitivity <- function(fits) {
  VCV      <- combine_VCV(fits)
  G_array  <- build_G_array(VCV, "animal")
  M_array  <- build_G_array(VCV, "dam")
  rA_array <- rA_from_G(G_array)

  s_rA <- posterior_summary_matrix(rA_array)
  rA_tbl <- tibble(
    quantity = paste0("rA_",
                      outer(TRAITS, TRAITS, paste, sep = "/")[upper.tri(diag(NT))]),
    mean = s_rA$mean[upper.tri(s_rA$mean)],
    l95  = s_rA$l95[upper.tri(s_rA$l95)],
    u95  = s_rA$u95[upper.tri(s_rA$u95)]
  )

  Va_draws <- sapply(seq_len(NT), function(i) G_array[, i, i])
  Va_tbl <- tibble(
    quantity = paste0("Va_", TRAITS),
    mean = colMeans(Va_draws, na.rm = TRUE),
    l95  = apply(Va_draws, 2, function(x) safe_hpd(x)[1]),
    u95  = apply(Va_draws, 2, function(x) safe_hpd(x)[2])
  )

  Vm_draws <- sapply(seq_len(NT), function(i) M_array[, i, i])
  Vm_tbl <- tibble(
    quantity = paste0("Vm_", TRAITS),
    mean = colMeans(Vm_draws, na.rm = TRUE),
    l95  = apply(Vm_draws, 2, function(x) safe_hpd(x)[1]),
    u95  = apply(Vm_draws, 2, function(x) safe_hpd(x)[2])
  )

  bind_rows(rA_tbl, Va_tbl, Vm_tbl)
}
```

### 13.3.2 Long-format comparison

**Table 26.** Prior sensitivity (long format) comparing the
parameter-expanded (PX, main) and inverse-Wishart (IW, V = I, ν = NT)
priors at AC and BB. One row per quantity: 10 r_A entries, 5 V_A
diagonals, and 5 V_M diagonals per population × prior combination.
Columns give posterior mean and 95% HPD bounds.

```{r prior-sensitivity-long}
ac_px <- extract_prior_sensitivity(fits_main$AC) %>%
  mutate(Pop = "AC", prior = "PX (default)")
ac_iw <- extract_prior_sensitivity(fits_iw_AC) %>%
  mutate(Pop = "AC", prior = "IW(V=I, nu=nt)")

bb_px <- extract_prior_sensitivity(fits_main$BB) %>%
  mutate(Pop = "BB", prior = "PX (default)")
bb_iw <- extract_prior_sensitivity(fits_iw_BB) %>%
  mutate(Pop = "BB", prior = "IW(V=I, nu=nt)")

prior_compare <- bind_rows(ac_px, ac_iw, bb_px, bb_iw) %>%
  select(Pop, prior, quantity, mean, l95, u95)
prior_compare
```

### 13.3.3 Wide format with `delta_mean` and HPD-overlap flag

**Table 27.** Prior sensitivity (wide format) pivoted so each row
gives one quantity with PX and IW posterior means and HPDs side by
side, the difference `delta_mean = IW − PX`, and `hpd_overlap` (TRUE
when the two priors' 95% HPDs overlap; FALSE flags prior-sensitive
quantities).

```{r prior-sensitivity-wide}
prior_compare_wide <- prior_compare %>%
  pivot_wider(names_from = prior,
              values_from = c(mean, l95, u95),
              names_glue = "{prior}_{.value}") %>%
  mutate(
    delta_mean = `IW(V=I, nu=nt)_mean` - `PX (default)_mean`,
    hpd_overlap = !(`PX (default)_u95` < `IW(V=I, nu=nt)_l95` |
                    `IW(V=I, nu=nt)_u95` < `PX (default)_l95`)
  ) %>%
  arrange(Pop, quantity)
prior_compare_wide
```

# 14 · Limitations and constraints on interpretation

Several features of the design and posterior structure constrain which
inferences this analysis supports. They are documented here so the strength
of each Results claim can be weighed against them.

## 14.1 Sampling design limits resolution of off-diagonal **G**

Each population is sampled at ~40–49 sires, ~3 dams per sire, and ~3
F-plants per dam (Table 03_sample_sizes). With this design, Falconer's
approximation gives SE(*h*²) ≈ 2/√(n_dams × n_off/dam) ≈ 0.10–0.12 per
trait — sufficient to resolve heritabilities of *h*² ≳ 0.3 but not lower.
The same data must support 10 off-diagonal entries in each 5 × 5 **G**, and
the posteriors confirm the resulting power limit: only 1 of 40 pairwise
*r*_A values has an HPD that excludes zero (HR δ¹³C × mean seed mass,
*r*_A = 0.58 [0.25, 0.87]). The remaining 39 span roughly [−0.8, +0.8],
meaning the data cannot distinguish strong positive from strong negative
correlations. Comparisons of **G** structure across populations (Random
Skewers, eigenanalysis, β_G) operate on this floor of resolution.

## 14.2 Variance components are prior-sensitive in posterior mean

The PX/IW comparison in §13.3 (Table 13_sensitivity_prior_wide) shows
two kinds of prior sensitivity. **Strict HPD non-overlap** is the
narrow case — fewer rows than the manuscript narrative might suggest:
only Va_fec_z at BB (1 of 5 V_A entries), Vm_corolla_z and Vm_fec_z at
BB (2 of 5 V_M entries), Va_SLA_z at AC (1 of 5), and Vm_SLA_z,
Vm_corolla_z, Vm_fec_z at AC (3 of 5). **Posterior-mean shifts**
between priors are more pervasive: IW(V = I, ν = NT = 5) is improper at
this dimension (its mean does not exist for ν ≤ NT + 1) and pulls
posterior means toward V = I, so IW means are systematically larger
than PX means even where their wide HPDs overlap — at BB, V_A means
shift 2–7× between priors for every trait. We adopt PX as the main
analysis because it is the appropriate weakly-informative prior for
variance components with few random-effect levels (Gelman 2006) and
because IW(V = I, ν = NT) is improper here. **Defensible claim**:
δ¹³C V_A at HR and AC, and mean seed mass V_A at HR, have lower HPDs
clearly above zero under both priors and posterior-mean shifts < 2×.
All other reported V_A and V_M values should be read as posterior
summaries under the PX prior rather than data-driven point estimates.
We avoid claims about heritability magnitude per §14.8.

## 14.3 Genetic selection gradients β_G are precision-limited

The Robertson–Price identity is the right object for predicting
evolutionary change, but its precision is bounded by that of off-diagonal
**G**. None of 16 β_G estimates has an HPD that excludes zero, and HPD
widths span 1.5–10 standardised units. We protect against occasional
near-singular **G**_zz draws via `rcond(G_zz) < 1e-10` exclusion and report
posterior medians rather than means (§11.2). Even with these fixes the β_G
intervals are too wide to test alignment with β_P or with environmental
predictors. β_G results are best read as confirming no clear conflict with
β_P given available precision, not as positive evidence of alignment.

## 14.4 BB phenotypic selection is underpowered

After listwise deletion, β_P at BB rests on n = 36 complete cases
(corolla, SLA, δ¹³C, msm, fec all non-missing) versus n = 185 at AC, 62 at
BO, 296 at HR (Table 11_beta_P_fit_summary). The marginal global *F* and
individual coefficient significance for BB are largely small-sample
artefacts and should be read alongside the n column.

## 14.5 Inference is conditional on survival

Trait values are measured only on F-plants — individuals that survived to
flowering. Where no designated F-plant survived in a segment, the first
surviving sibling was substituted; segments with no survivors enter the
fitness regression with W = 0 but contribute no trait data. The **G**
matrices estimated here are therefore **post-viability-selection G**, not
the parental **G** that determines next-generation response. If viability
selection acts on the same traits, our **G** systematically understates
pre-selection genetic variance and the implied response to fecundity
selection.

## 14.6 The viability GLM is a sibship-level analysis

Because F-plants are themselves survivors (or the first row in all-dead
segments), the binomial GLM regressing segment-level survival counts on
F-plant traits is not a Lande viability gradient. It asks whether the
surviving focal sib's trait predicts the survival fraction of the whole
segment — a sibship-level association, not individual viability selection.
This is the closest substitute possible given the protocol (no traits were
measured pre-mortality); it should be reported accordingly.

## 14.7 Other constraints

- **PC2 is unreliable in all four populations** (Table 08_pc2_reliability).
  λ₂ and λ₃ HPDs overlap at every site, so PC2 eigenvectors are not
  stably identified across draws. Only PC1 contributions are interpreted.

- **Block variance is weakly identified from three levels.** With three
  blocks per site, V_block carries little information from the data. An
  earlier version of the model used a near-improper inverse-Gamma-like
  prior (`nu = 0.002`) on the block component, which produced R̂ > 1.2 for
  five HR block parameters. The current PX prior on `idh(trait):block`
  (and on `idh(trait):transect_f`) — `nu = NT`, `alpha.V = 1000·I` —
  **substantially reduces but does not eliminate** the mixing
  pathology: per Table 05_convergence_rhat_above_1_1, 15 parameters
  across the four populations still have R̂ > 1.1 and 9 have R̂ > 1.2,
  almost all of them block-related. The PX prior follows Gelman (2006)
  for variance components with few random-effect levels and is the
  appropriate weakly-informative choice, but the fundamental
  identifiability issue remains: V_block cannot be precisely estimated
  from three random levels regardless of prior. Block remains in the
  model as a necessary nuisance term so that V_block is not absorbed
  into V_A or V_R, but V_block itself is not reported as a target
  quantity and the elevated R̂ values for block parameters should not be
  taken as evidence against the V_A and *r*_A estimates the block
  effect protects.

- **BB depends heavily on multivariate-normal imputation.** SLA, msm, and
  δ¹³C are 49–75% missing at BB. The trait-drop sensitivity in §13.2
  shows surviving *r*_A entries shift by at most 0.025 when SLA or msm is
  omitted, providing some reassurance that the 5-trait **G** is not driven
  by imputation artefacts — but it does not validate the imputed values
  themselves.

## 14.8 Defensible inferences from these data

Taking the constraints above together:

1. **Appreciable additive genetic variance (V_A) in δ¹³C and mean seed
   mass at HR, and δ¹³C at AC.** Posterior means and lower HPD bounds:
   HR δ¹³C V_A = 0.45 [0.24, 0.68]; HR mean seed mass V_A = 0.32 [0.07,
   0.57]; AC δ¹³C V_A = 0.42 [0.12, 0.72]. These three are the V_A
   estimates whose lower HPDs sit clearly above zero under the PX prior
   and whose posterior means change less than 2× between PX and IW. We
   report these as **V_A** rather than as heritabilities because
   per-draw *h*² = V_A / (V_A + V_M + V_block + V_transect + V_R) has a
   highly variable denominator under the wide PX prior on every
   variance component, so *h*² lower HPDs touch zero even when V_A is
   well-resolved (Table 06_h2_narrow_sense versus Table
   06_Va_diagonals). Posterior-mean *h*² for these three is 0.25–0.28,
   but lower HPD bounds are essentially zero; the defensible biological
   claim is the presence of heritable variation, not a specific *h*²
   magnitude.

2. **A credible additive genetic correlation between δ¹³C and mean seed
   mass at HR** (*r*_A = 0.58, HPD [0.27, 0.88]; the only one of 40
   pairwise *r*_A entries whose HPD excludes zero).

3. **Phenotypic selection patterns at AC and HR** favouring more negative
   δ¹³C and lower mean seed mass. β_P at AC has n = 185 and significant
   HC3 coefficients for both δ¹³C and msm; HR has n = 296 and a similar
   pattern.

4. **BB and BO are essentially unpowered** for selection and **G**-matrix
   structure inferences and should be reported as such.

5. **No credible evidence** for or against alignment of β_P with β_G in
   any population — even after rescaling β_G to the relative-fitness
   scale (§11.2), the intervals are too wide to test alignment.

These are the claims the data currently support. The Rmd documents
everything else (full **G** matrices, eigenanalyses, evolvability metrics,
Random Skewers, full β_P/β_G tables, prior sensitivity) so reviewers can
assess them against the limitations above.

# 15 · Save tables to disk

Every named table referenced in the document is written here to `tables/`
as a CSV. Figures are written to `figures/` automatically via the
`fig.path` chunk option set in the setup block (top of file).

```{r save-outputs}
save_tbl <- function(x, name) {
  if (is.null(x)) return(invisible(NULL))
  readr::write_csv(x, file.path("tables", paste0(name, ".csv")))
}

# Section 3 — data + sample sizes
save_tbl(sample_sizes,       "03_sample_sizes")
save_tbl(missingness_tbl,    "03_missingness_per_trait")
save_tbl(trait_means_tbl,    "03_raw_trait_means")

# Section 5 — convergence
save_tbl(conv_health,                       "05_convergence_rhat_health")
save_tbl(conv_tbl %>% filter(psrf > 1.1)
                  %>% arrange(desc(psrf)),  "05_convergence_rhat_above_1_1")
save_tbl(ess_health,                        "05_convergence_ess_health")
save_tbl(ess_tbl  %>% arrange(ESS)
                  %>% head(20),             "05_convergence_ess_lowest20")

# Section 6 — variance components and heritability
save_tbl(Va_tbl, "06_Va_diagonals")
save_tbl(Vm_tbl, "06_Vm_diagonals")
save_tbl(h2_tbl, "06_h2_narrow_sense")

# Section 7 — genetic correlations
save_tbl(rA_pairs_tbl, "07_rA_pairs")

# Section 8 — eigenanalysis
save_tbl(eig_summary_tbl,  "08_eigen_summary")
save_tbl(pc2_reliable_tbl, "08_pc2_reliability")
save_tbl(pc1_loadings_tbl, "08_pc1_loadings")
save_tbl(pc2_loadings_tbl, "08_pc2_loadings")
save_tbl(pc1_contrib_str,  "08_pc1_contributors")
save_tbl(pc2_contrib_str,  "08_pc2_contributors")

# Section 9 — evolvability
save_tbl(evol_tbl,      "09_evolvability_metrics")
save_tbl(evol_wide_tbl, "09_evolvability_metrics_wide")

# Section 10 — Random Skewers
save_tbl(rs_tbl, "10_random_skewers")

# Section 11 — selection
save_tbl(beta_P_tbl,         "11_beta_P_fec_only")
save_tbl(beta_P_total,       "11_beta_P_total_fitness")
save_tbl(beta_P_all,         "11_beta_P_combined")
save_tbl(beta_P_fit_summary, "11_beta_P_fit_summary")
save_tbl(beta_G_tbl,         "11_beta_G")

# Section 12 — viability
save_tbl(surv_summary, "12_survival_glm")
save_tbl(overdisp_tbl, "12_overdispersion")

# Section 13 — sensitivity analyses
save_tbl(beta_G_fec_compare,  "13_sensitivity_log1p_fec")
save_tbl(rA_BB_sensitivity,   "13_sensitivity_BB_trait_drop")
save_tbl(rA_BB_sens_summary,  "13_sensitivity_BB_trait_drop_summary")
save_tbl(prior_compare,       "13_sensitivity_prior_long")
save_tbl(prior_compare_wide,  "13_sensitivity_prior_wide")

message("Wrote ", length(list.files("tables", pattern = "\\.csv$")),
        " CSV files to tables/, ",
        length(list.files("figures", pattern = "\\.png$")),
        " PNG files to figures/")
```

# 16 · Session info

```{r session}
sessionInfo()
```
