---
title: "MCMC_multivariate_animal_model"
author: "Helen Payne"
date: "2026-01-08"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **ANGELO COAST** 

### Load in the data and clean it

```{r}
library(dplyr)
library(tidyverse)
AC_22 <- read_csv(here::here("data_sheets/compiled_sheets/AC_compiledsheet_full_2022.csv")) 

AC_22 <- AC_22 %>%
  dplyr::mutate(
    Donor = paste0(
      "Donor_AC_",
      str_remove(Donor, "^Donor_AC_|^AC_")
    ),
    Recipient = paste0(
      "Recipient_AC_",
      str_remove(Recipient, "^Recipient_AC_|^AC_")
    )
  )

#organize the data
AC_22 <- AC_22 %>%
  dplyr::mutate(
    sample_ID = str_c(Year, Donor, Recipient, Gen, str_pad(Transect, width = 3, pad = "0"), Plant_ID, sep = "_"), #create an id column for each individual
    sample_ID = str_replace(sample_ID, "^([^_]*)_", "\\1-"), #use "_" rather than "-"
    F_plant = ifelse(F_plant == TRUE, 1, 0), #make fitness plant binary (1/0) rather than true/false
    Transect = as.numeric(Transect)) #make sure location is a continuous variable

#clean the data
AC_22 <- AC_22 %>%
  dplyr::mutate(
    surv_to_flower = ifelse(total_fruits > 0 & surv_to_flower == 0, 1, surv_to_flower), #If a plant produced any fruits, it must have survived to flowering, even if the original surv_to_flower entry was coded as 0, This corrects miscodings where survival to flowering was mistakenly marked as 0 despite having fruit production
  ) %>%
    dplyr::filter(!is.na(Transect)) # We are removing W1 and W2 from these analyses because they were part of a different experiment involving a watered treatment. 
## In the original dataset, we included some extra fitness plants (see `anyFit_P`) in other transects for comparison with W1 and W2. 
### For the purposes of these analyses, we will exclude them and focus only on the `F_plant` values that are recorded for every segment.

AC_22$F_plant[is.na(AC_22$F_plant)] <- 0 #we need to make F_plant 0 and not NA

#Now we want to make sure every segment is represented by one F_plant (because we will filter by F_plant soon)
AC_22 <- AC_22 %>%
  dplyr::mutate(sample_prefix = substr(sample_ID, 1, 19)) %>% #create a sample prefix with the first 19 characters of sample ID
  group_by(sample_prefix) %>% #group by the prefix
  dplyr::mutate(
    all_dead = all(surv_to_flower == 0), #flags groups where no plants survived and where there are currently no fitness plants
##if a group has no fitness plants, it randomly assigns one plant in the group to be the fitness plant
    no_f_plant = sum(F_plant) == 0, 
    F_plant = if_else(
      no_f_plant & row_number() == sample(n(), 1),
      1L,
      F_plant
    ) #this will pick different individuals within the group each time unless you set a seed. For the purposes of this analyses, it will not matter because their values will be zero either way.
  ) %>%
  ungroup() %>%
  dplyr::select(-sample_prefix, -all_dead, -no_f_plant) #ungroup and remove the rows that were created\

#Now we want to determine how many individuals within a segment survived to flower
AC_22 <- AC_22 %>%
  dplyr::mutate(group = ceiling(row_number() / 8)) %>%  # Create a grouping variable for every 8 rows
  # Calculate the sum of 'surv_to_flower' within each group
  group_by(group) %>%
  dplyr::mutate(total_surv_to_flower = sum(surv_to_flower, na.rm = TRUE)) %>%
  ungroup() %>% # Ungroup to avoid grouped data frame
  # Remove the temporary 'group' column
  dplyr::select(-group)

# Filter to fitness plants once
AC_22 <- AC_22 %>%
  dplyr::filter(F_plant == 1)
```

```{r}
trait_means_AC <- AC_22 %>%
  dplyr::summarise(
    n = n(),

    corolla_mean = mean(corolla_diam_mm_SEG, na.rm = TRUE),
    corolla_se   = sd(corolla_diam_mm_SEG, na.rm = TRUE) / sqrt(sum(!is.na(corolla_diam_mm_SEG))),

    SLA_mean = mean(SLA_SEG, na.rm = TRUE),
    SLA_se   = sd(SLA_SEG, na.rm = TRUE) / sqrt(sum(!is.na(SLA_SEG))),

    d13C_mean = mean(d13C_SEG, na.rm = TRUE),
    d13C_se   = sd(d13C_SEG, na.rm = TRUE) / sqrt(sum(!is.na(d13C_SEG))),

    msm_mean = mean(msm_all, na.rm = TRUE),
    msm_se   = sd(msm_all, na.rm = TRUE) / sqrt(sum(!is.na(msm_all))),

    fec_mean = mean(est_fecundity, na.rm = TRUE),
    fec_se   = sd(est_fecundity, na.rm = TRUE) / sqrt(sum(!is.na(est_fecundity)))
  )

trait_means_AC
```


### Establish your Pedigree

```{r}
library(MCMCglmm)

#creating a pedigree (this one is missing parents though so we will make another)
ped <- data.frame(
  animal = as.character(AC_22$sample_ID),
  dam    = AC_22$Recipient,
  sire   = AC_22$Donor
)

#This is confusing and weird to me, but basically you need to make sure that the parents are also included in the "animal" category, so you can create a fully functioning pedigree. It is okay that you don't have the parents of the parents info, this will just be NA.
missing_parents <- setdiff(unique(c(ped$dam, ped$sire)), ped$animal)

ped2 <- bind_rows(
  ped,
  tibble(animal = missing_parents, dam = NA_character_, sire = NA_character_)
) %>%
  distinct(animal, .keep_all = TRUE)

library(nadiv) #(Non-Additive InVerses) is used in quantitative genetics to construct genetic relationship matrices (both additive and non-additive) and their inverses from pedigree data. These matrices are essential for running linear mixed effect models, also known as the "animal model," to estimate different sources of trait variation (e.g., additive genetic, dominance, and epistatic variances)

ped_ord <- prepPed(ped2)

Ainv <- MCMCglmm::inverseA(ped_ord)$Ainv

#We built the pedigree!!

#Double-checking to ensure that everything makes sense

##no one is their own parent
ped_ord %>%
  filter((!is.na(dam)  & animal == dam) | (!is.na(sire) & animal == sire))

##parents exist in the animal column
ped_ord %>%
  dplyr::summarise(
    dam_missing  = sum(!is.na(dam)  & !(dam  %in% animal)),
    sire_missing = sum(!is.na(sire) & !(sire %in% animal))
  )

parents <- unique(na.omit(c(ped_ord$dam, ped_ord$sire)))

#how many parents are there? 147
length(parents)
```

### Create your dataset
```{r}
dat_AC <- AC_22 %>%
  dplyr::mutate(
    animal = sample_ID,
    corolla = corolla_diam_mm_SEG,
    SLA     = SLA_SEG,
    d13C    = d13C_SEG,
    msm     = msm_all,
    fec     = est_fecundity
  )
#creating the dataset, simplifying names and renaming the sample ID for the animal model
#use the scale function to standardize the traits, this results in data with a mean of 0 and a standard deviation of 1
dat_AC <- dat_AC %>%
  dplyr::mutate(
    corolla_z = as.numeric(scale(corolla)),
    SLA_z     = as.numeric(scale(SLA)),
    d13C_z    = as.numeric(scale(d13C)),
    msm_z     = as.numeric(scale(msm_all)),    
    fec_z     = as.numeric(scale(fec)))

#Traits were standardized to mean 0 and variance 1 prior to analysis to facilitate comparison of variance components across traits

```

### Set your Prior and build your multivariate model

```{r}
# we are modeling multiple traits simultaneously to get Va and CovVa and rG, to do this we need to bind all the traits together
##this esentially binds multiple traits into a single response matrix

#We used a weakly informative inverse-Wishart prior (V = I, ν = nt) on additive genetic and residual (co)variance matrices after standardizing traits.
nt <- 5 #5 traits
prior <- list(
  G = list(G1 = list(V = diag(nt), nu = nt)),
  R = list(V = diag(nt), nu = nt)
)

dat_AC_model <- as.data.frame(dat_AC)

# make sure it's character, trimmed, then factor with ONLY valid levels
dat_AC_model$animal <- trimws(as.character(dat_AC_model$animal))
dat_AC_model$animal <- factor(dat_AC_model$animal, levels = rownames(Ainv))
dat_AC_model$animal <- droplevels(dat_AC_model$animal)
extra_levels <- setdiff(levels(dat_AC_model$animal), rownames(Ainv))

m <- MCMCglmm(
  fixed    = cbind(corolla_z, SLA_z, d13C_z, msm_z, fec_z) ~ trait - 1 + trait:Transect,
  random   = ~ us(trait):animal,
  rcov     = ~ us(trait):units,
  family   = rep("gaussian", nt),
  ginverse = list(animal = Ainv),
  data     = dat_AC_model,
  prior    = prior,
  nitt     = 115000,
  burnin   = 15000,
  thin     = 50
)

```

### Graphically show additive genetic variance (Va) for each trait is well estimated and the MCMC has converged

```{r}
ncol(m$VCV)
colnames(m$VCV)

graphics.off()

#Additive Genetic Variance
G_cols <- grep("\\.animal$", colnames(m$VCV), value = TRUE)
G_diag <- G_cols[grepl("trait([^:]+):trait\\1\\.animal$", G_cols)]  # diagonals only

par(mfrow = c(3,2), mar = c(4,4,2,1))
for (cc in G_diag) plot(m$VCV[, cc], type="l", main=cc, xlab="iter", ylab="value")

#Residual Variance
R_cols <- grep("\\.units$", colnames(m$VCV), value = TRUE)
R_diag <- R_cols[grepl("trait([^:]+):trait\\1\\.units$", R_cols)]

par(mfrow = c(3,2), mar = c(4,4,2,1))
for (cc in R_diag) plot(m$VCV[, cc], type="l", main=cc, xlab="iter", ylab="value")


library(coda)
effectiveSize(m$VCV[, c(G_diag, R_diag)]) #these ES look great and are big
autocorr.diag(m$VCV[, c(G_diag, R_diag)]) #this looks good too!
```

Obtain additive genetic variance and the likeihood interval for each trait

```{r}
#Extracting Va for all traits
Va_cols <- c(
  corolla = "traitcorolla_z:traitcorolla_z.animal",
  SLA     = "traitSLA_z:traitSLA_z.animal",
  d13C    = "traitd13C_z:traitd13C_z.animal",
  msm     = "traitmsm_z:traitmsm_z.animal",
  fec     = "traitfec_z:traitfec_z.animal"
)

Va_summary <- lapply(Va_cols, function(col) {
  x <- m$VCV[, col]
  c(
    mean = mean(x),
    l95  = HPDinterval(x)[1],
    u95  = HPDinterval(x)[2]
  )
})

Va_summary <- do.call(rbind, Va_summary)
Va_summary #Va is reported on the standardized trait scale
```

### Create a genetic variance-covariance matrices (G)

```{r}
traits <- c("corolla_z","SLA_z","d13C_z","msm_z","fec_z")
nt <- length(traits)

#creating a helper function that builds the specific column names MCMCglmm uses for each variance-covariance component
Gcol <- function(t1, t2) paste0("trait", t1, ":trait", t2, ".animal")

# Build an array: iterations × trait × trait- a 3 dimensional array of full posterior distributions of genetic cov(Va)
G_array <- array(NA_real_, dim = c(nrow(m$VCV), nt, nt),
                 dimnames = list(NULL, traits, traits))

for(i in seq_along(traits)){
  for(j in seq_along(traits)){
    colname <- Gcol(traits[i], traits[j])
    if(!colname %in% colnames(m$VCV)){
      stop("Missing column in m$VCV: ", colname)
    }
    G_array[, i, j] <- m$VCV[, colname]
  }
}
#The nested loops pull variance-covariance estimates from m$VCV (the variance-covariance components from the model) for every pair of traits across all MCMC iterations
#it calculates G_mean by averaging across all MCMC iterations (dimension 1), giving you the posterior mean genetic variance-covariance matrix

# Posterior mean G
G_mean <- apply(G_array, c(2,3), mean)
G_mean

AC_G_mean <- G_mean

#             corolla_z       SLA_z      d13C_z        msm_z        fec_z 
# corolla_z  0.48680864  0.03733133  0.02587508 -0.100500116  0.069900739
# SLA_z      0.03733133  0.41133405 -0.10300514 -0.029726589 -0.012776133
# d13C_z     0.02587508 -0.10300514  0.58441866  0.027827698  0.112543144
# msm_z     -0.10050012 -0.02972659  0.02782770  0.513217789 -0.006079159
# fec_z      0.06990074 -0.01277613  0.11254314 -0.006079159  0.424112739
```

### Convert the genetic variance-covariance matrix (G) into a genetic correlations matrix (rA)

```{r}
# Function: covariance matrix -> correlation matrix
#correlation matrices need to be on a unitless scale- scales covariances to correlations ranging from -1 to +1
cov2cor_mat <- function(S){
  d <- sqrt(diag(S)) #takes the square root of the diagonal elements and calculates the standard deviations
  S / (d %o% d) #Dividing each element by the product of the corresponding standard deviations using the outer product operator
}

# Posterior samples of rA- applies the transformation to every MCMC iteration
rA_array <- array(NA_real_, dim = dim(G_array), dimnames = dimnames(G_array))
for(k in 1:dim(G_array)[1]){
  rA_array[k,,] <- cov2cor_mat(G_array[k,,])
}

rA_mean <- apply(rA_array, c(2,3), mean)

# 95% HPD intervals for each correlation coefficient
rA_l95 <- rA_u95 <- rA_mean
for(i in 1:nt){
  for(j in 1:nt){
    h <- HPDinterval(as.mcmc(rA_array[,i,j]))
    rA_l95[i,j] <- h[1]
    rA_u95[i,j] <- h[2]
  }
}

rA_mean

#             corolla_z       SLA_z      d13C_z       msm_z       fec_z
# corolla_z  1.00000000  0.08692555  0.04554163 -0.20028987  0.15156807
# SLA_z      0.08692555  1.00000000 -0.20561270 -0.06407727 -0.02790244
# d13C_z     0.04554163 -0.20561270  1.00000000  0.04826298  0.21847088
# msm_z     -0.20028987 -0.06407727  0.04826298  1.00000000 -0.01155932
# fec_z      0.15156807 -0.02790244  0.21847088 -0.01155932  1.00000000


 rA_df <- expand.grid(
   trait1 = rownames(rA_mean),
   trait2 = colnames(rA_mean)
 ) %>%
   dplyr::mutate(
     mean = as.vector(rA_mean),
     l95  = as.vector(rA_l95),
     u95  = as.vector(rA_u95),
     
#     # significance (CI does not overlap 0)
     sig = (l95 > 0 & u95 > 0) | (l95 < 0 & u95 < 0),
#     
#     # formatted label (2 decimal places)
     label = sprintf("%.2f [%.2f, %.2f]", mean, l95, u95),
#     
#     # bold significant values (markdown style)
    label = ifelse(sig, paste0("**", label, "**"), label)
   )

rA_df <- rA_df %>%
  dplyr::mutate(
    trait1 = factor(trait1, levels = unique(trait1)),
    trait2 = factor(trait2, levels = unique(trait2))
  ) %>%
  filter(as.numeric(trait1) >= as.numeric(trait2))

rA_table <- rA_df %>%
  dplyr::select(trait1, trait2, label) %>%
  tidyr::pivot_wider(names_from = trait2, values_from = label)

rA_table
```

### Plot the rA matrix

```{r}
# Create a named vector for renaming
trait_names <- c(
  "SLA_z" = "specific leaf area",
  "msm_z" = "mean seed mass",
  "fec_z" = "fecundity",
  "d13C_z" = "δC13",
  "corolla_z" = "corolla diameter"
)

plot_rA <- as.data.frame(rA_mean) %>%
  tibble::rownames_to_column("trait1") %>%
  pivot_longer(-trait1, names_to="trait2", values_to="rA") %>%
  dplyr::mutate(
    trait1 = recode_factor(trait1, !!!trait_names),
    trait2 = recode_factor(trait2, !!!trait_names)
  )

AC_rA<- ggplot(plot_rA, aes(trait1, trait2, fill = rA)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", rA)), size = 7) +
  scale_fill_gradient2(low = "#3B9AB2", mid = "white", high = "#F21A00", 
                       midpoint = 0) +
  coord_equal() +
  labs(x=NULL, y=NULL, fill="rA") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle=45, hjust=1, size = 16),
    axis.text.y = element_text(size = 16)
  )

AC_rA


ggsave("AC_rA.jpeg", #save the plot
       plot = AC_rA, width = 8, height = 8, dpi = 700)
```

### Estimate heritability (h2)
```{r}

# matching residual variance columns
Vr_cols <- c(
  corolla = "traitcorolla_z:traitcorolla_z.units",
  SLA     = "traitSLA_z:traitSLA_z.units",
  d13C    = "traitd13C_z:traitd13C_z.units",
  msm     = "traitmsm_z:traitmsm_z.units",
  fec     = "traitfec_z:traitfec_z.units"
)

h2_summary <- lapply(names(Va_cols), function(tr){

  Va <- m$VCV[, Va_cols[tr]]
  Vr <- m$VCV[, Vr_cols[tr]]

  h2 <- Va / (Va + Vr)

  c(
    mean = mean(h2),
    l95  = HPDinterval(as.mcmc(h2))[1],
    u95  = HPDinterval(as.mcmc(h2))[2]
  )
})

h2_summary <- do.call(rbind, h2_summary)

h2_summary
```

### Determine Evolvability and related metrics

```{r}
set.seed(1)#use the same numbers everytime you rerun the code

#sample random unit vectors in trait space, each vector represents a hypothetical direction of selection (ex. selection for high WUE and big corollas), they are all normalized 
rand_unit_vecs <- function(p, nvec=2000){
  M <- matrix(rnorm(nvec*p), nvec, p)
  M / sqrt(rowSums(M^2))
}

B <- rand_unit_vecs(nt, nvec = 2000)

evolv_metrics_oneG <- function(G, B){ 
  # evolvability per direction
  e <- rowSums((B %*% G) * B)  # beta'Gbeta
  ##how much genetic variance exists in each selection direction
  
  # conditional evolvability
  Ginv <- solve(G)
  c <- 1 / rowSums((B %*% Ginv) * B)  # 1/(beta'G^-1 beta)
  ##expected response if selection acts only on that trait combination (accounting for genetic correlations that cause indirect responses)

  # respondability (one common definition)
  GG <- G %*% G
  r <- sqrt(rowSums((B %*% GG) * B))
  ##Magnitude of the genetic response vector -(response capacity)

  #autonomy
  a <- c / e ## whether traits can evolve independently
 
  #integration
   i <- 1 - a ##opposite of autonomy, high integration = traits are genetically coupled

  c(e_mean = mean(e),
    e_min  = min(e),
    e_max  = max(e),
    c_mean = mean(c),
    r_mean = mean(r),
    a_mean = mean(a),
    i_mean = mean(i))
}

# compute metrics for each posterior draw of G, giving posterior distributions for each metric (mean, min, max)
metrics_post <- t(apply(G_array, 1, evolv_metrics_oneG, B = B))
metrics_post <- as.data.frame(metrics_post)

# posterior summaries like Mattila table
metric_summary <- metrics_post %>%
  dplyr::summarise(across(everything(), list(
    mean = mean,
    l95  = ~HPDinterval(as.mcmc(.x))[1],
    u95  = ~HPDinterval(as.mcmc(.x))[2]
  )))

metric_summary

G_AC <- G_mean
G_array_AC <- G_array
```

###Clear the environment
```{r}
#rm(list = ls())
```


# **BODEGA BAY** 

###Load in the data and clean it
```{r}
library(dplyr)
library(tidyverse)

BB_22 <- read_csv(here::here("data_sheets/compiled_sheets/BB_compiledsheet_full_2022.csv"))

BB_22 <- BB_22 %>%
  dplyr::mutate(
    Donor = paste0(
      "Donor_BB_",
      str_remove(Donor, "^Donor_BB_|^BB_")
    ),
    Recipient = paste0(
      "Recipient_BB_",
      str_remove(Recipient, "^Recipient_BB_|^BB_")
    )
  )

# organize the data
BB_22 <- BB_22 %>%
  dplyr::mutate(
    sample_ID = str_c(Year, Donor, Recipient, Gen, str_pad(Location, width = 3, pad = "0"), Plant_ID, sep = "_"),
    sample_ID = str_replace(sample_ID, "^([^_]*)_", "\\1-"),
    F_plant   = ifelse(F_plant == TRUE, 1, 0),
    Location  = as.numeric(Location)
  )

# clean the data
BB_22 <- BB_22 %>%
  dplyr::mutate(
    surv_to_flower = ifelse(total_fruits > 0 & surv_to_flower == 0, 1, surv_to_flower)
  ) %>%
  filter(!is.na(Location))

BB_22$F_plant[is.na(BB_22$F_plant)] <- 0

# ensure every segment is represented by one F_plant
BB_22 <- BB_22 %>%
  dplyr::mutate(sample_prefix = substr(sample_ID, 1, 19)) %>%
  group_by(sample_prefix) %>%
  dplyr::mutate(
    all_dead   = all(surv_to_flower == 0),
    no_f_plant = sum(F_plant) == 0,
    F_plant    = if_else(
      no_f_plant & row_number() == sample(n(), 1),
      1L,
      F_plant
    )
  ) %>%
  ungroup() %>%
  dplyr::select(-sample_prefix, -all_dead, -no_f_plant)

# how many individuals within a segment survived to flower
BB_22 <- BB_22 %>%
  dplyr::mutate(group = ceiling(row_number() / 8)) %>%
  group_by(group) %>%
  dplyr::mutate(total_surv_to_flower = sum(surv_to_flower, na.rm = TRUE)) %>%
  ungroup() %>%
  dplyr::select(-group)

# Filter to fitness plants once
BB_22 <- BB_22 %>%
  dplyr::filter(F_plant == 1)
```

```{r}
trait_means_BB <- BB_22 %>%
  dplyr::summarise(
    n = n(),

    corolla_mean = mean(corolla_diam_mm_SEG, na.rm = TRUE),
    corolla_se   = sd(corolla_diam_mm_SEG, na.rm = TRUE) / sqrt(sum(!is.na(corolla_diam_mm_SEG))),

    SLA_mean = mean(SLA_SEG, na.rm = TRUE),
    SLA_se   = sd(SLA_SEG, na.rm = TRUE) / sqrt(sum(!is.na(SLA_SEG))),

    d13C_mean = mean(d13C_SEG, na.rm = TRUE),
    d13C_se   = sd(d13C_SEG, na.rm = TRUE) / sqrt(sum(!is.na(d13C_SEG))),

    msm_mean = mean(msm_all, na.rm = TRUE),
    msm_se   = sd(msm_all, na.rm = TRUE) / sqrt(sum(!is.na(msm_all))),

    fec_mean = mean(est_fecundity, na.rm = TRUE),
    fec_se   = sd(est_fecundity, na.rm = TRUE) / sqrt(sum(!is.na(est_fecundity)))
  )

trait_means_BB
```



### Establish your pedigree
```{r}

library(MCMCglmm)

ped <- data.frame(
  animal = as.character(BB_22$sample_ID),
  dam    = BB_22$Recipient,
  sire   = BB_22$Donor
)

missing_parents <- setdiff(unique(c(ped$dam, ped$sire)), ped$animal)

ped2 <- bind_rows(
  ped,
  tibble(animal = missing_parents, dam = NA_character_, sire = NA_character_)
) %>%
  distinct(animal, .keep_all = TRUE)

library(nadiv)

ped_ord <- prepPed(ped2)
Ainv <- MCMCglmm::inverseA(ped_ord)$Ainv

# sanity checks
ped_ord %>%
  dplyr::filter((!is.na(dam)  & animal == dam) | (!is.na(sire) & animal == sire))

ped_ord %>%
  summarise(
    dam_missing  = sum(!is.na(dam)  & !(dam  %in% animal)),
    sire_missing = sum(!is.na(sire) & !(sire %in% animal))
  )

parents <- unique(na.omit(c(ped_ord$dam, ped_ord$sire)))
length(parents)
```

### Create your dataset
```{r}
dat_BB <- BB_22 %>%
  dplyr::mutate(
    animal  = sample_ID,
    corolla = corolla_diam_mm_SEG,
    SLA     = SLA_SEG,
    d13C    = d13C_SEG,
    msm     = msm_all,
    fec     = est_fecundity
  )

dat_BB <- dat_BB %>%
  dplyr::mutate(
    corolla_z = as.numeric(scale(corolla)),
    SLA_z     = as.numeric(scale(SLA)),
    d13C_z    = as.numeric(scale(d13C)),
    msm_z     = as.numeric(scale(msm_all)),
    fec_z     = as.numeric(scale(fec))
  )
```

### Set your Prior and build your multivariate model

```{r}
# we are modeling multiple traits simultaneously to get Va and CovVa and rG, to do this we need to bind all the traits together
##this esentially binds multiple traits into a single response matrix

#We used a weakly informative inverse-Wishart prior (V = I, ν = nt) on additive genetic and residual (co)variance matrices after standardizing traits.
nt <- 5 #5 traits
prior <- list(
  G = list(G1 = list(V = diag(nt), nu = nt)),
  R = list(V = diag(nt), nu = nt)
)

dat_BB_model <- as.data.frame(dat_BB)

# make sure it's character, trimmed, then factor with ONLY valid levels
dat_BB_model$animal <- trimws(as.character(dat_BB_model$animal))
dat_BB_model$animal <- factor(dat_BB_model$animal, levels = rownames(Ainv))
dat_BB_model$animal <- droplevels(dat_BB_model$animal)
extra_levels <- setdiff(levels(dat_BB_model$animal), rownames(Ainv))

m <- MCMCglmm(
  fixed    = cbind(corolla_z, SLA_z, d13C_z, msm_z, fec_z) ~ trait - 1 + trait:Location,
  random   = ~ us(trait):animal,
  rcov     = ~ us(trait):units,
  family   = rep("gaussian", nt),
  ginverse = list(animal = Ainv),
  data     = dat_BB_model,
  prior    = prior,
  nitt     = 150000,
  burnin   = 15000,
  thin     = 100
)

```

### Graphically show additive genetic variance (Va) for each trait is well estimated and the MCMC has converged

```{r}
ncol(m$VCV)
colnames(m$VCV)

graphics.off()

#Additive Genetic Variance
G_cols <- grep("\\.animal$", colnames(m$VCV), value = TRUE)
G_diag <- G_cols[grepl("trait([^:]+):trait\\1\\.animal$", G_cols)]  # diagonals only

par(mfrow = c(3,2), mar = c(4,4,2,1))
for (cc in G_diag) plot(m$VCV[, cc], type="l", main=cc, xlab="iter", ylab="value")

#Residual Variance
R_cols <- grep("\\.units$", colnames(m$VCV), value = TRUE)
R_diag <- R_cols[grepl("trait([^:]+):trait\\1\\.units$", R_cols)]

par(mfrow = c(3,2), mar = c(4,4,2,1))
for (cc in R_diag) plot(m$VCV[, cc], type="l", main=cc, xlab="iter", ylab="value")
#trait_z.animal <- additive genetic variance
#trait_z.units <- residual variance

library(coda)
effectiveSize(m$VCV[, c(G_diag, R_diag)]) #these ES look great and are big (ESS > 200 → acceptable, ESS > 500 → good, ESS > 1000 → excellent / publication-quality)
autocorr.diag(m$VCV[, c(G_diag, R_diag)]) #this looks good too!
```


Obtain additive genetic variance and the likeihood interval for each trait

```{r}
#Extracting Va for all traits
Va_cols <- c(
  corolla = "traitcorolla_z:traitcorolla_z.animal",
  SLA     = "traitSLA_z:traitSLA_z.animal",
  d13C    = "traitd13C_z:traitd13C_z.animal",
  msm     = "traitmsm_z:traitmsm_z.animal",
  fec     = "traitfec_z:traitfec_z.animal"
)

Va_summary <- lapply(Va_cols, function(col) {
  x <- m$VCV[, col]
  c(
    mean = mean(x),
    l95  = HPDinterval(x)[1],
    u95  = HPDinterval(x)[2]
  )
})

Va_summary <- do.call(rbind, Va_summary)
Va_summary #Va is reported on the standardized trait scale

#             mean       l95       u95
# corolla 0.4176219 0.2235215 0.6310133
# SLA     0.5712026 0.2977519 0.8820673
# d13C    0.5138537 0.2851097 0.7597251
# msm     0.6488303 0.3153332 0.9756159
# fec     0.3348900 0.1922488 0.5069683

```

### Create a genetic variance-covariance matrices (G)

```{r}
traits <- c("corolla_z","SLA_z","d13C_z","msm_z","fec_z")
nt <- length(traits)

#creating a helper function that builds the specific column names MCMCglmm uses for each variance-covariance component
Gcol <- function(t1, t2) paste0("trait", t1, ":trait", t2, ".animal")

# Build an array: iterations × trait × trait- a 3 dimensional array of full posterior distributions of genetic cov(Va)
G_array <- array(NA_real_, dim = c(nrow(m$VCV), nt, nt),
                 dimnames = list(NULL, traits, traits))

for(i in seq_along(traits)){
  for(j in seq_along(traits)){
    colname <- Gcol(traits[i], traits[j])
    if(!colname %in% colnames(m$VCV)){
      stop("Missing column in m$VCV: ", colname)
    }
    G_array[, i, j] <- m$VCV[, colname]
  }
}
#The nested loops pull variance-covariance estimates from m$VCV (the variance-covariance components from the model) for every pair of traits across all MCMC iterations
#it calculates G_mean by averaging across all MCMC iterations (dimension 1), giving you the posterior mean genetic variance-covariance matrix

# Posterior mean G
G_mean <- apply(G_array, c(2,3), mean)
G_mean

BB_G_mean <- G_mean

#             corolla_z       SLA_z        d13C_z       msm_z         fec_z
# corolla_z  0.41762187 -0.04315665 -0.0316116671 -0.06679687  0.0726176414
# SLA_z     -0.04315665  0.57120255 -0.0806296165  0.07060006 -0.0174755245
# d13C_z    -0.03161167 -0.08062962  0.5138537273 -0.05710089 -0.0002741675
# msm_z     -0.06679687  0.07060006 -0.0571008939  0.64883030 -0.0641557470
# fec_z      0.07261764 -0.01747552 -0.0002741675 -0.06415575  0.3348899800
```



### Convert the genetic variance-covariance matrix (G) into a genetic correlations matrix (rA)

```{r}
# Function: covariance matrix -> correlation matrix
#correlation matrices need to be on a unitless scale- scales covariances to correlations ranging from -1 to +1
cov2cor_mat <- function(S){
  d <- sqrt(diag(S)) #takes the square root of the diagonal elements and calculates the standard deviations
  S / (d %o% d) #Dividing each element by the product of the corresponding standard deviations using the outer product operator
}

# Posterior samples of rA- applies the transformation to every MCMC iteration
rA_array <- array(NA_real_, dim = dim(G_array), dimnames = dimnames(G_array))
for(k in 1:dim(G_array)[1]){
  rA_array[k,,] <- cov2cor_mat(G_array[k,,])
}

rA_mean <- apply(rA_array, c(2,3), mean)

# 95% HPD intervals for each correlation coefficient
rA_l95 <- rA_u95 <- rA_mean
for(i in 1:nt){
  for(j in 1:nt){
    h <- HPDinterval(as.mcmc(rA_array[,i,j]))
    rA_l95[i,j] <- h[1]
    rA_u95[i,j] <- h[2]
  }
}

rA_mean

#            corolla_z       SLA_z        d13C_z       msm_z         fec_z
# corolla_z  1.00000000 -0.08337582 -6.597962e-02 -0.12399835  1.852250e-01
# SLA_z     -0.08337582  1.00000000 -1.487747e-01  0.11200270 -3.691358e-02
# d13C_z    -0.06597962 -0.14877473  1.000000e+00 -0.09947554  8.697927e-05
# msm_z     -0.12399835  0.11200270 -9.947554e-02  1.00000000 -1.360929e-01
# fec_z      0.18522502 -0.03691358  8.697927e-05 -0.13609285  1.000000e+00

rA_df <- expand.grid(
   trait1 = rownames(rA_mean),
   trait2 = colnames(rA_mean)
 ) %>%
   dplyr::mutate(
     mean = as.vector(rA_mean),
     l95  = as.vector(rA_l95),
     u95  = as.vector(rA_u95),
     
#     # significance (CI does not overlap 0)
     sig = (l95 > 0 & u95 > 0) | (l95 < 0 & u95 < 0),
#     
#     # formatted label (2 decimal places)
     label = sprintf("%.2f [%.2f, %.2f]", mean, l95, u95),
#     
#     # bold significant values (markdown style)
    label = ifelse(sig, paste0("**", label, "**"), label)
   )

rA_df <- rA_df %>%
  dplyr::mutate(
    trait1 = factor(trait1, levels = unique(trait1)),
    trait2 = factor(trait2, levels = unique(trait2))
  ) %>%
  filter(as.numeric(trait1) >= as.numeric(trait2))

rA_table <- rA_df %>%
  select(trait1, trait2, label) %>%
  tidyr::pivot_wider(names_from = trait2, values_from = label)

rA_table
```

### Plot the rA matrix

```{r}
# Create a named vector for renaming
trait_names <- c(
  "SLA_z" = "specific leaf area",
  "msm_z" = "mean seed mass",
  "fec_z" = "fecundity",
  "d13C_z" = "δC13",
  "corolla_z" = "corolla diameter"
)

plot_rA <- as.data.frame(rA_mean) %>%
  tibble::rownames_to_column("trait1") %>%
  pivot_longer(-trait1, names_to="trait2", values_to="rA") %>%
  dplyr::mutate(
    trait1 = recode_factor(trait1, !!!trait_names),
    trait2 = recode_factor(trait2, !!!trait_names)
  )

BB_rA<- ggplot(plot_rA, aes(trait1, trait2, fill = rA)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", rA)), size = 7) +
  scale_fill_gradient2(low = "#3B9AB2", mid = "white", high = "#F21A00", 
                       midpoint = 0) +
  coord_equal() +
  labs(x=NULL, y=NULL, fill="rA") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle=45, hjust=1, size = 16),
    axis.text.y = element_text(size = 16)
  )

BB_rA


ggsave("BB_rA.jpeg", #save the plot
       plot = BB_rA, width = 8, height = 8, dpi = 700)
```

### Estimate heritability (h2)
```{r}

# matching residual variance columns
Vr_cols <- c(
  corolla = "traitcorolla_z:traitcorolla_z.units",
  SLA     = "traitSLA_z:traitSLA_z.units",
  d13C    = "traitd13C_z:traitd13C_z.units",
  msm     = "traitmsm_z:traitmsm_z.units",
  fec     = "traitfec_z:traitfec_z.units"
)

h2_summary <- lapply(names(Va_cols), function(tr){

  Va <- m$VCV[, Va_cols[tr]]
  Vr <- m$VCV[, Vr_cols[tr]]

  h2 <- Va / (Va + Vr)

  c(
    mean = mean(h2),
    l95  = HPDinterval(as.mcmc(h2))[1],
    u95  = HPDinterval(as.mcmc(h2))[2]
  )
})

h2_summary <- do.call(rbind, h2_summary)

h2_summary
```


### Determine Evolvability and related metrics

```{r}
set.seed(1)#use the same numbers everytime you rerun the code

#sample random unit vectors in trait space, each vector represents a hypothetical direction of selection (ex. selection for high WUE and big corollas), they are all normalized 
rand_unit_vecs <- function(p, nvec=2000){
  M <- matrix(rnorm(nvec*p), nvec, p)
  M / sqrt(rowSums(M^2))
}

B <- rand_unit_vecs(nt, nvec = 2000)

evolv_metrics_oneG <- function(G, B){ 
  # evolvability per direction
  e <- rowSums((B %*% G) * B)  # beta'Gbeta
  ##how much genetic variance exists in each selection direction
  
  # conditional evolvability
  Ginv <- solve(G)
  c <- 1 / rowSums((B %*% Ginv) * B)  # 1/(beta'G^-1 beta)
  ##expected response if selection acts only on that trait combination (accounting for genetic correlations that cause indirect responses)

  # respondability (one common definition)
  GG <- G %*% G
  r <- sqrt(rowSums((B %*% GG) * B))
  ##Magnitude of the genetic response vector -(response capacity)

  #autonomy
  a <- c / e ## whether traits can evolve independently
 
  #integration
   i <- 1 - a ##opposite of autonomy, high integration = traits are genetically coupled

  c(e_mean = mean(e),
    e_min  = min(e),
    e_max  = max(e),
    c_mean = mean(c),
    r_mean = mean(r),
    a_mean = mean(a),
    i_mean = mean(i))
}

# compute metrics for each posterior draw of G, giving posterior distributions for each metric (mean, min, max)
metrics_post <- t(apply(G_array, 1, evolv_metrics_oneG, B = B))
metrics_post <- as.data.frame(metrics_post)

# posterior summaries like Mattila table
metric_summary <- metrics_post %>%
  dplyr::summarise(across(everything(), list(
    mean = mean,
    l95  = ~HPDinterval(as.mcmc(.x))[1],
    u95  = ~HPDinterval(as.mcmc(.x))[2]
  )))

metric_summary

G_BB <- G_mean

G_array_BB <- G_array

```

###Clear the environment
```{r}
#rm(list = ls())
```

# **BLUE OAK** 

###Load in the data and clean it
```{r}
library(dplyr)
library(tidyverse)


BO_22 <- read_csv(here::here("data_sheets/compiled_sheets/BO_compiledsheet_full_2022.csv"))

BO_22 <- BO_22 %>%
  mutate(
    Donor = paste0(
      "Donor_BO_",
      str_remove(Donor, "^Donor_BO_|^BO_")
    ),
    Recipient = paste0(
      "Recipient_BO_",
      str_remove(Recipient, "^Recipient_BO_|^BO_")
    )
  )

# organize the data
BO_22 <- BO_22 %>%
  mutate(
    sample_ID = str_c(Year, Donor, Recipient, Gen, str_pad(Transect, width = 3, pad = "0"), Plant_ID, sep = "_"),
    sample_ID = str_replace(sample_ID, "^([^_]*)_", "\\1-"),
    F_plant   = ifelse(F_plant == TRUE, 1, 0),
    Transect  = as.numeric(Transect)
  )

# clean the data
BO_22 <- BO_22 %>%
  mutate(
    surv_to_flower = ifelse(total_fruits > 0 & surv_to_flower == 0, 1, surv_to_flower)
  ) %>%
  filter(!is.na(Transect))

BO_22$F_plant[is.na(BO_22$F_plant)] <- 0

# ensure every segment is represented by one F_plant
BO_22 <- BO_22 %>%
  dplyr::mutate(sample_prefix = substr(sample_ID, 1, 19)) %>%
  dplyr::group_by(sample_prefix) %>%
  dplyr::mutate(
    all_dead   = all(surv_to_flower == 0),
    no_f_plant = sum(F_plant) == 0,
    F_plant    = if_else(
      no_f_plant & dplyr::row_number() == sample(dplyr::n(), 1),
      1L,
      F_plant
    )
  ) %>%
  dplyr::ungroup() %>%
  dplyr::select(-sample_prefix, -all_dead, -no_f_plant)

# how many individuals within a segment survived to flower
BO_22 <- BO_22 %>%
  dplyr::mutate(group = ceiling(row_number() / 8)) %>%
  dplyr::group_by(group) %>%
  dplyr::mutate(total_surv_to_flower = sum(surv_to_flower, na.rm = TRUE)) %>%
  ungroup() %>%
  dplyr::select(-group)

# Filter to fitness plants once
BO_22 <- BO_22 %>%
  filter(F_plant == 1)
```

```{r}
trait_means_BO <- BO_22 %>%
  dplyr::summarise(
    n = n(),

    corolla_mean = mean(corolla_diam_mm_SEG, na.rm = TRUE),
    corolla_se   = sd(corolla_diam_mm_SEG, na.rm = TRUE) / sqrt(sum(!is.na(corolla_diam_mm_SEG))),

    SLA_mean = mean(SLA_SEG, na.rm = TRUE),
    SLA_se   = sd(SLA_SEG, na.rm = TRUE) / sqrt(sum(!is.na(SLA_SEG))),

    d13C_mean = mean(d13C_SEG, na.rm = TRUE),
    d13C_se   = sd(d13C_SEG, na.rm = TRUE) / sqrt(sum(!is.na(d13C_SEG))),

    msm_mean = mean(msm_all, na.rm = TRUE),
    msm_se   = sd(msm_all, na.rm = TRUE) / sqrt(sum(!is.na(msm_all))),

    fec_mean = mean(est_fecundity, na.rm = TRUE),
    fec_se   = sd(est_fecundity, na.rm = TRUE) / sqrt(sum(!is.na(est_fecundity)))
  )

trait_means_BO
```

### Establish your pedigree
```{r}
library(MCMCglmm)

ped <- data.frame(
  animal = as.character(BO_22$sample_ID),
  dam    = BO_22$Recipient,
  sire   = BO_22$Donor
)

missing_parents <- setdiff(unique(c(ped$dam, ped$sire)), ped$animal)

ped2 <- bind_rows(
  ped,
  tibble(animal = missing_parents, dam = NA_character_, sire = NA_character_)
) %>%
  distinct(animal, .keep_all = TRUE)

library(nadiv)

ped_ord <- prepPed(ped2)
Ainv <- MCMCglmm::inverseA(ped_ord)$Ainv

# sanity checks
ped_ord %>%
  filter((!is.na(dam)  & animal == dam) | (!is.na(sire) & animal == sire))

ped_ord %>%
  summarise(
    dam_missing  = sum(!is.na(dam)  & !(dam  %in% animal)),
    sire_missing = sum(!is.na(sire) & !(sire %in% animal))
  )

parents <- unique(na.omit(c(ped_ord$dam, ped_ord$sire)))
length(parents)
```

### Create your dataset
```{r}
dat_BO <- BO_22 %>%
  mutate(
    animal  = sample_ID,
    corolla = corolla_diam_mm_SEG,
    SLA     = SLA_SEG,
    d13C    = d13C_SEG,
    msm     = msm_all,
    fec     = est_fecundity
  )

dat_BO <- dat_BO %>%
  mutate(
    corolla_z = as.numeric(scale(corolla)),
    SLA_z     = as.numeric(scale(SLA)),
    d13C_z    = as.numeric(scale(d13C)),
    msm_z     = as.numeric(scale(msm_all)),
    fec_z     = as.numeric(scale(fec))
  )
```

### Set your Prior and build your multivariate model

```{r}
# we are modeling multiple traits simultaneously to get Va and CovVa and rG, to do this we need to bind all the traits together
##this esentially binds multiple traits into a single response matrix

#We used a weakly informative inverse-Wishart prior (V = I, ν = nt) on additive genetic and residual (co)variance matrices after standardizing traits.
nt <- 5 #5 traits
prior <- list(
  G = list(G1 = list(V = diag(nt), nu = nt)),
  R = list(V = diag(nt), nu = nt)
)

dat_BO_model <- as.data.frame(dat_BO)

# make sure it's character, trimmed, then factor with ONLY valid levels
dat_BO_model$animal <- trimws(as.character(dat_BO_model$animal))
dat_BO_model$animal <- factor(dat_BO_model$animal, levels = rownames(Ainv))
dat_BO_model$animal <- droplevels(dat_BO_model$animal)
extra_levels <- setdiff(levels(dat_BO_model$animal), rownames(Ainv))

m <- MCMCglmm(
  fixed    = cbind(corolla_z, SLA_z, d13C_z, msm_z, fec_z) ~ trait - 1 + trait:Transect,
  random   = ~ us(trait):animal,
  rcov     = ~ us(trait):units,
  family   = rep("gaussian", nt),
  ginverse = list(animal = Ainv),
  data     = dat_BO_model,
  prior    = prior,
  nitt     = 115000,
  burnin   = 15000,
  thin     = 50
)

```

### Graphically show additive genetic variance (Va) for each trait is well estimated and the MCMC has converged

```{r}
ncol(m$VCV)
colnames(m$VCV)

graphics.off()

#Additive Genetic Variance
G_cols <- grep("\\.animal$", colnames(m$VCV), value = TRUE)
G_diag <- G_cols[grepl("trait([^:]+):trait\\1\\.animal$", G_cols)]  # diagonals only

par(mfrow = c(3,2), mar = c(4,4,2,1))
for (cc in G_diag) plot(m$VCV[, cc], type="l", main=cc, xlab="iter", ylab="value")

#Residual Variance
R_cols <- grep("\\.units$", colnames(m$VCV), value = TRUE)
R_diag <- R_cols[grepl("trait([^:]+):trait\\1\\.units$", R_cols)]

par(mfrow = c(3,2), mar = c(4,4,2,1))
for (cc in R_diag) plot(m$VCV[, cc], type="l", main=cc, xlab="iter", ylab="value")


library(coda)
effectiveSize(m$VCV[, c(G_diag, R_diag)]) #these ES look great and are big (ESS > 200 → acceptable, ESS > 500 → good, ESS > 1000 → excellent / publication-quality)
autocorr.diag(m$VCV[, c(G_diag, R_diag)]) #this looks good too!
```

Obtain additive genetic variance and the likeihood interval for each trait

```{r}
#Extracting Va for all traits
Va_cols <- c(
  corolla = "traitcorolla_z:traitcorolla_z.animal",
  SLA     = "traitSLA_z:traitSLA_z.animal",
  d13C    = "traitd13C_z:traitd13C_z.animal",
  msm     = "traitmsm_z:traitmsm_z.animal",
  fec     = "traitfec_z:traitfec_z.animal"
)

Va_summary <- lapply(Va_cols, function(col) {
  x <- m$VCV[, col]
  c(
    mean = mean(x),
    l95  = HPDinterval(x)[1],
    u95  = HPDinterval(x)[2]
  )
})

Va_summary <- do.call(rbind, Va_summary)
Va_summary #Va is reported on the standardized trait scale

#             mean       l95       u95
# corolla 0.4121287 0.2330165 0.6338884
# SLA     0.4661304 0.2299056 0.7215188
# d13C    0.4836679 0.2562423 0.7274870
# msm     0.5798554 0.3076512 0.8878574
# fec     0.3808473 0.2038289 0.5698233


```

### Create a genetic variance-covariance matrices (G)

```{r}
traits <- c("corolla_z","SLA_z","d13C_z","msm_z","fec_z")
nt <- length(traits)

#creating a helper function that builds the specific column names MCMCglmm uses for each variance-covariance component
Gcol <- function(t1, t2) paste0("trait", t1, ":trait", t2, ".animal")

# Build an array: iterations × trait × trait- a 3 dimensional array of full posterior distributions of genetic cov(Va)
G_array <- array(NA_real_, dim = c(nrow(m$VCV), nt, nt),
                 dimnames = list(NULL, traits, traits))

for(i in seq_along(traits)){
  for(j in seq_along(traits)){
    colname <- Gcol(traits[i], traits[j])
    if(!colname %in% colnames(m$VCV)){
      stop("Missing column in m$VCV: ", colname)
    }
    G_array[, i, j] <- m$VCV[, colname]
  }
}
#The nested loops pull variance-covariance estimates from m$VCV (the variance-covariance components from the model) for every pair of traits across all MCMC iterations
#it calculates G_mean by averaging across all MCMC iterations (dimension 1), giving you the posterior mean genetic variance-covariance matrix

# Posterior mean G
G_mean <- apply(G_array, c(2,3), mean)
G_mean

BO_G_mean <- G_mean

#              corolla_z        SLA_z        d13C_z        msm_z       fec_z
# corolla_z  0.4121286870 -0.012109264 -0.0009242919  0.029335482  0.04475777
# SLA_z     -0.0121092637  0.466130406  0.0251612948 -0.004791275 -0.03068266
# d13C_z    -0.0009242919  0.025161295  0.4836678737  0.036673078 -0.01226578
# msm_z      0.0293354819 -0.004791275  0.0366730776  0.579855359 -0.01552157
# fec_z      0.0447577681 -0.030682665 -0.0122657762 -0.015521566  0.38084727
```

### Convert the genetic variance-covariance matrix (G) into a genetic correlations matrix (rA)

```{r}
# Function: covariance matrix -> correlation matrix
#correlation matrices need to be on a unitless scale- scales covariances to correlations ranging from -1 to +1
cov2cor_mat <- function(S){
  d <- sqrt(diag(S)) #takes the square root of the diagonal elements and calculates the standard deviations
  S / (d %o% d) #Dividing each element by the product of the corresponding standard deviations using the outer product operator
}

# Posterior samples of rA- applies the transformation to every MCMC iteration
rA_array <- array(NA_real_, dim = dim(G_array), dimnames = dimnames(G_array))
for(k in 1:dim(G_array)[1]){
  rA_array[k,,] <- cov2cor_mat(G_array[k,,])
}

rA_mean <- apply(rA_array, c(2,3), mean)

# 95% HPD intervals for each correlation coefficient
rA_l95 <- rA_u95 <- rA_mean
for(i in 1:nt){
  for(j in 1:nt){
    h <- HPDinterval(as.mcmc(rA_array[,i,j]))
    rA_l95[i,j] <- h[1]
    rA_u95[i,j] <- h[2]
  }
}

rA_mean

#             corolla_z        SLA_z        d13C_z        msm_z       fec_z
# corolla_z  1.000000e+00 -0.029313670  4.318832e-05  0.061556968  0.11070757
# SLA_z     -2.931367e-02  1.000000000  5.273155e-02 -0.009248131 -0.07294173
# d13C_z     4.318832e-05  0.052731546  1.000000e+00  0.069892257 -0.02822551
# msm_z      6.155697e-02 -0.009248131  6.989226e-02  1.000000000 -0.03300220
# fec_z      1.107076e-01 -0.072941728 -2.822551e-02 -0.033002195  1.00000000

rA_df <- expand.grid(
   trait1 = rownames(rA_mean),
   trait2 = colnames(rA_mean)
 ) %>%
   mutate(
     mean = as.vector(rA_mean),
     l95  = as.vector(rA_l95),
     u95  = as.vector(rA_u95),
     
#     # significance (CI does not overlap 0)
     sig = (l95 > 0 & u95 > 0) | (l95 < 0 & u95 < 0),
#     
#     # formatted label (2 decimal places)
     label = sprintf("%.2f [%.2f, %.2f]", mean, l95, u95),
#     
#     # bold significant values (markdown style)
    label = ifelse(sig, paste0("**", label, "**"), label)
   )

rA_df <- rA_df %>%
  mutate(
    trait1 = factor(trait1, levels = unique(trait1)),
    trait2 = factor(trait2, levels = unique(trait2))
  ) %>%
  filter(as.numeric(trait1) >= as.numeric(trait2))

rA_table <- rA_df %>%
  select(trait1, trait2, label) %>%
  tidyr::pivot_wider(names_from = trait2, values_from = label)

rA_table
```

### Plot the rA matrix

```{r}
# Create a named vector for renaming
trait_names <- c(
  "SLA_z" = "specific leaf area",
  "msm_z" = "mean seed mass",
  "fec_z" = "fecundity",
  "d13C_z" = "δC13",
  "corolla_z" = "corolla diameter"
)

plot_rA <- as.data.frame(rA_mean) %>%
  tibble::rownames_to_column("trait1") %>%
  pivot_longer(-trait1, names_to="trait2", values_to="rA") %>%
  mutate(
    trait1 = recode_factor(trait1, !!!trait_names),
    trait2 = recode_factor(trait2, !!!trait_names)
  )

BO_rA<- ggplot(plot_rA, aes(trait1, trait2, fill = rA)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", rA)), size = 7) +
  scale_fill_gradient2(low = "#3B9AB2", mid = "white", high = "#F21A00", 
                       midpoint = 0) +
  coord_equal() +
  labs(x=NULL, y=NULL, fill="rA") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle=45, hjust=1, size = 16),
    axis.text.y = element_text(size = 16)
  )

BO_rA


ggsave("BO_rA.jpeg", #save the plot
       plot = BO_rA, width = 8, height = 8, dpi = 700)
```

### Estimate heritability (h2)
```{r}

# matching residual variance columns
Vr_cols <- c(
  corolla = "traitcorolla_z:traitcorolla_z.units",
  SLA     = "traitSLA_z:traitSLA_z.units",
  d13C    = "traitd13C_z:traitd13C_z.units",
  msm     = "traitmsm_z:traitmsm_z.units",
  fec     = "traitfec_z:traitfec_z.units"
)

h2_summary <- lapply(names(Va_cols), function(tr){

  Va <- m$VCV[, Va_cols[tr]]
  Vr <- m$VCV[, Vr_cols[tr]]

  h2 <- Va / (Va + Vr)

  c(
    mean = mean(h2),
    l95  = HPDinterval(as.mcmc(h2))[1],
    u95  = HPDinterval(as.mcmc(h2))[2]
  )
})

h2_summary <- do.call(rbind, h2_summary)

h2_summary
```

### Determine Evolvability and related metrics

```{r}
set.seed(1)#use the same numbers everytime you rerun the code

#sample random unit vectors in trait space, each vector represents a hypothetical direction of selection (ex. selection for high WUE and big corollas), they are all normalized 
rand_unit_vecs <- function(p, nvec=2000){
  M <- matrix(rnorm(nvec*p), nvec, p)
  M / sqrt(rowSums(M^2))
}

B <- rand_unit_vecs(nt, nvec = 2000)

evolv_metrics_oneG <- function(G, B){ 
  # evolvability per direction
  e <- rowSums((B %*% G) * B)  # beta'Gbeta
  ##how much genetic variance exists in each selection direction
  
  # conditional evolvability
  Ginv <- solve(G)
  c <- 1 / rowSums((B %*% Ginv) * B)  # 1/(beta'G^-1 beta)
  ##expected response if selection acts only on that trait combination (accounting for genetic correlations that cause indirect responses)

  # respondability (one common definition)
  GG <- G %*% G
  r <- sqrt(rowSums((B %*% GG) * B))
  ##Magnitude of the genetic response vector -(response capacity)

  #autonomy
  a <- c / e ## whether traits can evolve independently
 
  #integration
   i <- 1 - a ##opposite of autonomy, high integration = traits are genetically coupled

  c(e_mean = mean(e),
    e_min  = min(e),
    e_max  = max(e),
    c_mean = mean(c),
    r_mean = mean(r),
    a_mean = mean(a),
    i_mean = mean(i))
}

# compute metrics for each posterior draw of G, giving posterior distributions for each metric (mean, min, max)
metrics_post <- t(apply(G_array, 1, evolv_metrics_oneG, B = B))
metrics_post <- as.data.frame(metrics_post)

# posterior summaries like Mattila table
metric_summary <- metrics_post %>%
  dplyr::summarise(across(everything(), list(
    mean = mean,
    l95  = ~HPDinterval(as.mcmc(.x))[1],
    u95  = ~HPDinterval(as.mcmc(.x))[2]
  )))

metric_summary

G_BO <- G_mean
G_array_BO <- G_array


```

###Clear the environment
```{r}
#rm(list = ls())
```

# **HASTINGS RANCH** 


###Load in the data and clean it
```{r}
library(dplyr)
library(tidyverse)

HR_22 <- read_csv(here::here("data_sheets/compiled_sheets/HR_compiledsheet_full_2022.csv"))

HR_22 <- HR_22 %>%
  dplyr::mutate(
    Donor = paste0(
      "Donor_HR_",
      str_remove(Donor, "^Donor_HR_|^HR_")
    ),
    Recipient = paste0(
      "Recipient_HR_",
      str_remove(Recipient, "^Recipient_HR_|^HR_")
    )
  )

# organize the data
HR_22 <- HR_22 %>%
  dplyr::mutate(
    sample_ID = str_c(Year, Donor, Recipient, Gen, str_pad(Transect, width = 3, pad = "0"), Plant_ID, sep = "_"),
    sample_ID = str_replace(sample_ID, "^([^_]*)_", "\\1-"),
    F_plant   = ifelse(F_plant == TRUE, 1, 0),
    Transect  = as.numeric(Transect)
  )

# clean the data
HR_22 <- HR_22 %>%
  dplyr::mutate(
    surv_to_flower = ifelse(total_fruits > 0 & surv_to_flower == 0, 1, surv_to_flower)
  ) %>%
  dplyr::filter(!is.na(Transect))

HR_22$F_plant[is.na(HR_22$F_plant)] <- 0

# ensure every segment is represented by one F_plant
HR_22 <- HR_22 %>%
  dplyr::mutate(sample_prefix = substr(sample_ID, 1, 19)) %>%
group_by(sample_prefix) %>%
  dplyr::mutate(
    all_dead   = all(surv_to_flower == 0),
    no_f_plant = sum(F_plant) == 0,
    F_plant    = if_else(
      no_f_plant & row_number() == sample(n(), 1),
      1L,
      F_plant
    )
  ) %>%
  ungroup() %>%
  dplyr::select(-sample_prefix, -all_dead, -no_f_plant)

# how many individuals within a segment survived to flower
HR_22 <- HR_22 %>%
  dplyr::mutate(group = ceiling(row_number() / 8)) %>%
group_by(group) %>%
  dplyr::mutate(total_surv_to_flower = sum(surv_to_flower, na.rm = TRUE)) %>%
  ungroup() %>%
  dplyr::select(-group)

# Filter to fitness plants once
HR_22 <- HR_22 %>%
  filter(F_plant == 1)
```

```{r}
trait_means_HR <- HR_22 %>%
  dplyr::summarise(
    n = n(),

    corolla_mean = mean(corolla_diam_mm_SEG, na.rm = TRUE),
    corolla_se   = sd(corolla_diam_mm_SEG, na.rm = TRUE) / sqrt(sum(!is.na(corolla_diam_mm_SEG))),

    SLA_mean = mean(SLA_SEG, na.rm = TRUE),
    SLA_se   = sd(SLA_SEG, na.rm = TRUE) / sqrt(sum(!is.na(SLA_SEG))),

    d13C_mean = mean(d13C_SEG, na.rm = TRUE),
    d13C_se   = sd(d13C_SEG, na.rm = TRUE) / sqrt(sum(!is.na(d13C_SEG))),

    msm_mean = mean(msm_all, na.rm = TRUE),
    msm_se   = sd(msm_all, na.rm = TRUE) / sqrt(sum(!is.na(msm_all))),

    fec_mean = mean(est_fecundity, na.rm = TRUE),
    fec_se   = sd(est_fecundity, na.rm = TRUE) / sqrt(sum(!is.na(est_fecundity)))
  )

trait_means_HR
```

### Establish your pedigree
```{r}
library(MCMCglmm)

ped <- data.frame(
  animal = as.character(HR_22$sample_ID),
  dam    = HR_22$Recipient,
  sire   = HR_22$Donor
)

missing_parents <- setdiff(unique(c(ped$dam, ped$sire)), ped$animal)

ped2 <- bind_rows(
  ped,
  tibble(animal = missing_parents, dam = NA_character_, sire = NA_character_)
) %>%
  distinct(animal, .keep_all = TRUE)

library(nadiv)

ped_ord <- prepPed(ped2)
Ainv <- MCMCglmm::inverseA(ped_ord)$Ainv

# sanity checks
ped_ord %>%
  filter((!is.na(dam)  & animal == dam) | (!is.na(sire) & animal == sire))

ped_ord %>%
  summarise(
    dam_missing  = sum(!is.na(dam)  & !(dam  %in% animal)),
    sire_missing = sum(!is.na(sire) & !(sire %in% animal))
  )

parents <- unique(na.omit(c(ped_ord$dam, ped_ord$sire)))
length(parents)
```

### Create your dataset
```{r}
dat_HR <- HR_22 %>%
  mutate(
    animal  = sample_ID,
    corolla = corolla_diam_mm_SEG,
    SLA     = SLA_SEG,
    d13C    = d13C_SEG,
    msm     = msm_all,
    fec     = est_fecundity
  )

dat_HR <- dat_HR %>%
  mutate(
    corolla_z = as.numeric(scale(corolla)),
    SLA_z     = as.numeric(scale(SLA)),
    d13C_z    = as.numeric(scale(d13C)),
    msm_z     = as.numeric(scale(msm_all)),
    fec_z     = as.numeric(scale(fec))
  )
```

### Set your Prior and build your multivariate model

```{r}
# we are modeling multiple traits simultaneously to get Va and CovVa and rG, to do this we need to bind all the traits together
##this esentially binds multiple traits into a single response matrix

#We used a weakly informative inverse-Wishart prior (V = I, ν = nt) on additive genetic and residual (co)variance matrices after standardizing traits.
nt <- 5 #5 traits
prior <- list(
  G = list(G1 = list(V = diag(nt), nu = nt)),
  R = list(V = diag(nt), nu = nt)
)

dat_HR_model <- as.data.frame(dat_HR)

# make sure it's character, trimmed, then factor with ONLY valid levels
dat_HR_model$animal <- trimws(as.character(dat_HR_model$animal))
dat_HR_model$animal <- factor(dat_HR_model$animal, levels = rownames(Ainv))
dat_HR_model$animal <- droplevels(dat_HR_model$animal)
extra_levels <- setdiff(levels(dat_HR_model$animal), rownames(Ainv))

m <- MCMCglmm(
  fixed    = cbind(corolla_z, SLA_z, d13C_z, msm_z, fec_z) ~ trait - 1 + trait:Transect,
  random   = ~ us(trait):animal,
  rcov     = ~ us(trait):units,
  family   = rep("gaussian", nt),
  ginverse = list(animal = Ainv),
  data     = dat_HR_model,
  prior    = prior,
  nitt     = 115000,
  burnin   = 15000,
  thin     = 50
)

```

### Graphically show additive genetic variance (Va) for each trait is well estimated and the MCMC has converged

```{r}
ncol(m$VCV)
colnames(m$VCV)

graphics.off()

#Additive Genetic Variance
G_cols <- grep("\\.animal$", colnames(m$VCV), value = TRUE)
G_diag <- G_cols[grepl("trait([^:]+):trait\\1\\.animal$", G_cols)]  # diagonals only

par(mfrow = c(3,2), mar = c(4,4,2,1))
for (cc in G_diag) plot(m$VCV[, cc], type="l", main=cc, xlab="iter", ylab="value")

#Residual Variance
R_cols <- grep("\\.units$", colnames(m$VCV), value = TRUE)
R_diag <- R_cols[grepl("trait([^:]+):trait\\1\\.units$", R_cols)]

par(mfrow = c(3,2), mar = c(4,4,2,1))
for (cc in R_diag) plot(m$VCV[, cc], type="l", main=cc, xlab="iter", ylab="value")


library(coda)
effectiveSize(m$VCV[, c(G_diag, R_diag)]) #these ES look great and are big (ESS > 200 → acceptable, ESS > 500 → good, ESS > 1000 → excellent / publication-quality)
autocorr.diag(m$VCV[, c(G_diag, R_diag)]) #this looks good too!
```

Obtain additive genetic variance and the likeihood interval for each trait

```{r}
#Extracting Va for all traits
Va_cols <- c(
  corolla = "traitcorolla_z:traitcorolla_z.animal",
  SLA     = "traitSLA_z:traitSLA_z.animal",
  d13C    = "traitd13C_z:traitd13C_z.animal",
  msm     = "traitmsm_z:traitmsm_z.animal",
  fec     = "traitfec_z:traitfec_z.animal"
)

Va_summary <- lapply(Va_cols, function(col) {
  x <- m$VCV[, col]
  c(
    mean = mean(x),
    l95  = HPDinterval(x)[1],
    u95  = HPDinterval(x)[2]
  )
})

Va_summary <- do.call(rbind, Va_summary)
Va_summary #Va is reported on the standardized trait scale

#             mean       l95       u95
# corolla 0.4126368 0.2363783 0.6100629
# SLA     0.3415553 0.1956410 0.5187924
# d13C    0.5366817 0.3342637 0.7502045
# msm     0.5328403 0.3239066 0.7722104
# fec     0.3639742 0.2102014 0.5370444


```

### Create a genetic variance-covariance matrices (G)

```{r}
traits <- c("corolla_z","SLA_z","d13C_z","msm_z","fec_z")
nt <- length(traits)

#creating a helper function that builds the specific column names MCMCglmm uses for each variance-covariance component
Gcol <- function(t1, t2) paste0("trait", t1, ":trait", t2, ".animal")

# Build an array: iterations × trait × trait- a 3 dimensional array of full posterior distributions of genetic cov(Va)
G_array <- array(NA_real_, dim = c(nrow(m$VCV), nt, nt),
                 dimnames = list(NULL, traits, traits))

for(i in seq_along(traits)){
  for(j in seq_along(traits)){
    colname <- Gcol(traits[i], traits[j])
    if(!colname %in% colnames(m$VCV)){
      stop("Missing column in m$VCV: ", colname)
    }
    G_array[, i, j] <- m$VCV[, colname]
  }
}
#The nested loops pull variance-covariance estimates from m$VCV (the variance-covariance components from the model) for every pair of traits across all MCMC iterations
#it calculates G_mean by averaging across all MCMC iterations (dimension 1), giving you the posterior mean genetic variance-covariance matrix

# Posterior mean G
G_mean <- apply(G_array, c(2,3), mean)
G_mean

HR_G_mean <- G_mean

#             corolla_z        SLA_z       d13C_z        msm_z       fec_z
# corolla_z 0.412636812  0.012650585  0.009891039  0.018931653  0.07053319
# SLA_z     0.012650585  0.341555255 -0.031832067 -0.008788597  0.00166747
# d13C_z    0.009891039 -0.031832067  0.536681669  0.236085151 -0.06357722
# msm_z     0.018931653 -0.008788597  0.236085151  0.532840340 -0.10214363
# fec_z     0.070533193  0.001667470 -0.063577220 -0.102143629  0.36397416
```

### Convert the genetic variance-covariance matrix (G) into a genetic correlations matrix (rA)

```{r}
# Function: covariance matrix -> correlation matrix
#correlation matrices need to be on a unitless scale- scales covariances to correlations ranging from -1 to +1
cov2cor_mat <- function(S){
  d <- sqrt(diag(S)) #takes the square root of the diagonal elements and calculates the standard deviations
  S / (d %o% d) #Dividing each element by the product of the corresponding standard deviations using the outer product operator
}

# Posterior samples of rA- applies the transformation to every MCMC iteration
rA_array <- array(NA_real_, dim = dim(G_array), dimnames = dimnames(G_array))
for(k in 1:dim(G_array)[1]){
  rA_array[k,,] <- cov2cor_mat(G_array[k,,])
}

rA_mean <- apply(rA_array, c(2,3), mean)

# 95% HPD intervals for each correlation coefficient
rA_l95 <- rA_u95 <- rA_mean
for(i in 1:nt){
  for(j in 1:nt){
    h <- HPDinterval(as.mcmc(rA_array[,i,j]))
    rA_l95[i,j] <- h[1]
    rA_u95[i,j] <- h[2]
  }
}

rA_mean

#           corolla_z        SLA_z      d13C_z       msm_z        fec_z
# corolla_z 1.00000000  0.033680769  0.01670331  0.03751055  0.179851155
# SLA_z     0.03368077  1.000000000 -0.07361229 -0.02012276  0.004177986
# d13C_z    0.01670331 -0.073612289  1.00000000  0.43631628 -0.147786734
# msm_z     0.03751055 -0.020122762  0.43631628  1.00000000 -0.229398897
# fec_z     0.17985115  0.004177986 -0.14778673 -0.22939890  1.000000000

rA_df <- expand.grid(
   trait1 = rownames(rA_mean),
   trait2 = colnames(rA_mean)
 ) %>%
   mutate(
     mean = as.vector(rA_mean),
     l95  = as.vector(rA_l95),
     u95  = as.vector(rA_u95),
     
#     # significance (CI does not overlap 0)
     sig = (l95 > 0 & u95 > 0) | (l95 < 0 & u95 < 0),
#     
#     # formatted label (2 decimal places)
     label = sprintf("%.2f [%.2f, %.2f]", mean, l95, u95),
#     
#     # bold significant values (markdown style)
    label = ifelse(sig, paste0("**", label, "**"), label)
   )

rA_df <- rA_df %>%
  mutate(
    trait1 = factor(trait1, levels = unique(trait1)),
    trait2 = factor(trait2, levels = unique(trait2))
  ) %>%
  filter(as.numeric(trait1) >= as.numeric(trait2))

rA_table <- rA_df %>%
  select(trait1, trait2, label) %>%
  tidyr::pivot_wider(names_from = trait2, values_from = label)

rA_table
```

### Plot the rA matrix

```{r}
# Create a named vector for renaming
trait_names <- c(
  "SLA_z" = "SLA",
  "msm_z" = "msm",
  "fec_z" = "fecundity",
  "d13C_z" = "δC13",
  "corolla_z" = "corolla diameter"
)

plot_rA <- as.data.frame(rA_mean) %>%
  tibble::rownames_to_column("trait1") %>%
  pivot_longer(-trait1, names_to="trait2", values_to="rA") %>%
  mutate(
    trait1 = recode_factor(trait1, !!!trait_names),
    trait2 = recode_factor(trait2, !!!trait_names)
  )

HR_rA<- ggplot(plot_rA, aes(trait1, trait2, fill = rA)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", rA)), size = 7) +
  scale_fill_gradient2(low = "#3B9AB2", mid = "white", high = "#F21A00", 
                       midpoint = 0) +
  coord_equal() +
  labs(x=NULL, y=NULL, fill="rA") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle=45, hjust=1, size = 16),
    axis.text.y = element_text(size = 16)
  )

HR_rA


ggsave("HR_rA.jpeg", #save the plot
       plot = HR_rA, width = 8, height = 8, dpi = 700)
```


### Estimate heritability (h2)
```{r}

# matching residual variance columns
Vr_cols <- c(
  corolla = "traitcorolla_z:traitcorolla_z.units",
  SLA     = "traitSLA_z:traitSLA_z.units",
  d13C    = "traitd13C_z:traitd13C_z.units",
  msm     = "traitmsm_z:traitmsm_z.units",
  fec     = "traitfec_z:traitfec_z.units"
)

h2_summary <- lapply(names(Va_cols), function(tr){

  Va <- m$VCV[, Va_cols[tr]]
  Vr <- m$VCV[, Vr_cols[tr]]

  h2 <- Va / (Va + Vr)

  c(
    mean = mean(h2),
    l95  = HPDinterval(as.mcmc(h2))[1],
    u95  = HPDinterval(as.mcmc(h2))[2]
  )
})

h2_summary <- do.call(rbind, h2_summary)

h2_summary
```

### Determine Evolvability and related metrics

```{r}
set.seed(1)#use the same numbers everytime you rerun the code

#sample random unit vectors in trait space, each vector represents a hypothetical direction of selection (ex. selection for high WUE and big corollas), they are all normalized 
rand_unit_vecs <- function(p, nvec=2000){
  M <- matrix(rnorm(nvec*p), nvec, p)
  M / sqrt(rowSums(M^2))
}

B <- rand_unit_vecs(nt, nvec = 2000)

evolv_metrics_oneG <- function(G, B){ 
  # evolvability per direction
  e <- rowSums((B %*% G) * B)  # beta'Gbeta
  ##how much genetic variance exists in each selection direction
  
  # conditional evolvability
  Ginv <- solve(G)
  c <- 1 / rowSums((B %*% Ginv) * B)  # 1/(beta'G^-1 beta)
  ##expected response if selection acts only on that trait combination (accounting for genetic correlations that cause indirect responses)

  # respondability (one common definition)
  GG <- G %*% G
  r <- sqrt(rowSums((B %*% GG) * B))
  ##Magnitude of the genetic response vector -(response capacity)

  #autonomy
  a <- c / e ## whether traits can evolve independently
 
  #integration
   i <- 1 - a ##opposite of autonomy, high integration = traits are genetically coupled

  c(e_mean = mean(e),
    e_min  = min(e),
    e_max  = max(e),
    c_mean = mean(c),
    r_mean = mean(r),
    a_mean = mean(a),
    i_mean = mean(i))
}

# compute metrics for each posterior draw of G, giving posterior distributions for each metric (mean, min, max)
metrics_post <- t(apply(G_array, 1, evolv_metrics_oneG, B = B))
metrics_post <- as.data.frame(metrics_post)

# posterior summaries like Mattila table
metric_summary <- metrics_post %>%
  summarise(across(everything(), list(
    mean = mean,
    l95  = ~HPDinterval(as.mcmc(.x))[1],
    u95  = ~HPDinterval(as.mcmc(.x))[2]
  )))

metric_summary

G_HR <- G_mean

G_array_HR <- G_array

```

###Clear the environment
```{r}
#rm(list = ls())
```

#**COMPARE POPULATIONS**

```{r}
G_list <- list(
  AC = G_AC,
  BB = G_BB,
  BO = G_BO,
  HR = G_HR
)

G_summary <- purrr::map_dfr(names(G_list), function(pop) {
  G <- G_list[[pop]]
  eig <- eigen(G)

  tibble(
    Pop = pop,
    trace_G = sum(diag(G)),
    lambda1 = eig$values[1],
    lambda2 = eig$values[2],
    lambda3 = eig$values[3],
    lambda4 = eig$values[4],
    lambda5 = eig$values[5],
    prop_PC1 = eig$values[1] / sum(eig$values),
    prop_PC2 = eig$values[2] / sum(eig$values)
  )
})

G_summary


PC1_leading_eigenvectors <- purrr::map_dfr(names(G_list), function(pop) {
  eig <- eigen(G_list[[pop]])

  tibble(
    Pop = pop,
    trait = rownames(G_list[[pop]]),
    loading_PC1 = eig$vectors[, 1]
  )
})

PC1_leading_eigenvectors

PC2_leading_eigenvectors <- purrr::map_dfr(names(G_list), function(pop) {
  eig <- eigen(G_list[[pop]])

  tibble(
    Pop = pop,
    trait = rownames(G_list[[pop]]),
    loading_PC2 = eig$vectors[, 2]
  )
})

PC2_leading_eigenvectors

```

```{r}
library(coda)
library(dplyr)
library(purrr)

get_G_metrics <- function(G_array, pop_name){

  n_iter <- dim(G_array)[1]

  metrics <- map_dfr(1:n_iter, function(i){

    G <- G_array[i,,]

    eig <- eigen(G)

    tibble(
      Pop = pop_name,
      trace_G = sum(diag(G)),
      lambda1 = eig$values[1],
      lambda2 = eig$values[2],
      lambda3 = eig$values[3],
      lambda4 = eig$values[4],
      lambda5 = eig$values[5],
      prop_PC1 = eig$values[1] / sum(eig$values)
    )
  })

  # summarize with mean + HPD
  metrics %>%
    summarise(across(-Pop, list(
      mean = mean,
      l95  = ~HPDinterval(as.mcmc(.x))[1],
      u95  = ~HPDinterval(as.mcmc(.x))[2]
    ))) %>%
    mutate(Pop = pop_name, .before = 1)
}

G_metrics_all <- bind_rows(
  get_G_metrics(G_array_HR, "HR")
)

G_metrics_all <- bind_rows(
  get_G_metrics(G_array_AC, "AC"),
  get_G_metrics(G_array_BB, "BB"),
  get_G_metrics(G_array_BO, "BO"),
  get_G_metrics(G_array_HR, "HR")
)

format_ci <- function(mean, l95, u95){
  sprintf("%.3f [%.3f, %.3f]", mean, l95, u95)
}

G_table <- G_metrics_all %>%
  mutate(
    trace_G = format_ci(trace_G_mean, trace_G_l95, trace_G_u95),
    lambda1 = format_ci(lambda1_mean, lambda1_l95, lambda1_u95),
    prop_PC1 = format_ci(prop_PC1_mean, prop_PC1_l95, prop_PC1_u95)
  ) %>%
  select(Pop, trace_G, lambda1, prop_PC1)

G_table
```

```{r}
#One rA figure
rA_list <- list(
  AC = rA_AC,
  BB = rA_BB,
  BO = rA_BO,
  HR = rA_HR
)

library(dplyr)
library(tidyr)
library(purrr)

rA_all <- map_dfr(names(rA_list), function(pop){

  as.data.frame(rA_list[[pop]]) %>%
    tibble::rownames_to_column("trait1") %>%
    pivot_longer(-trait1, names_to = "trait2", values_to = "rA") %>%
    mutate(Pop = pop)

})

rA_all <- rA_all %>%
  mutate(
    trait1 = dplyr::recode_factor(trait1, !!!trait_names),
    trait2 = dplyr::recode_factor(trait2, !!!trait_names)
  )

rA_plot <- ggplot(rA_all, aes(trait1, trait2, fill = rA)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", rA)), size = 3) +
  scale_fill_gradient2(
    low = "#3B9AB2",
    mid = "white",
    high = "#F21A00",
    midpoint = 0
  ) +
  coord_equal() +
  facet_wrap(~Pop) +
  labs(x = NULL, y = NULL, fill = "rA") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    strip.text = element_text(size = 14, face = "bold")
  )

ggsave(
  "rA_all_populations.jpeg",
  plot = rA_plot,
  width = 12,
  height = 10,
  dpi = 700
)
```

```{r}

library(tidyverse)
library(ggh4x)

  # Create dataframe
  trait_means <- tibble(
    Pop = rep(c("AC","BB","BO","HR"), each = 5),
    trait = rep(c("corolla diameter","SLA","δ13C","mean seed mass","fecundity"), times = 4),
    mean = c(
      20.55,17.72,-31.29,1.90,25.47,
      22.89,17.17,-29.64,3.92,3.98,
      23.50,15.52,-28.31,3.77,5.30,
      21.94,17.29,-30.77,2.72,17.73
    ),
    se = c(
      0.17,0.34,0.05,0.04, 1.91,
      0.28,0.57,0.08,0.23, 0.95,
      0.22,0.34,0.08,0.15, 0.80,
      0.14,0.31,0.04,0.05, 1.23
    )
  )

  # Keep original values (no flip)
  trait_means <- trait_means %>%
    mutate(mean_plot = mean,
           se_plot   = se)

  trait_means$trait <- factor(trait_means$trait,
    levels = c("corolla diameter","SLA","δ13C","mean seed mass","fecundity"))

  # Plot with reversed y-axis for δ13C only
  trait_fig <- ggplot(trait_means, aes(Pop, mean_plot, fill = Pop)) +
    geom_col(width = 0.7) +
    geom_errorbar(aes(ymin = mean_plot - se_plot, ymax = mean_plot + se_plot),
                  width = 0.2, na.rm = TRUE) +
    facet_wrap(~trait, scales = "free_y", nrow = 1) +
    facetted_pos_scales(
      y = list(
        trait == "δ13C" ~ scale_y_reverse()
      )
    ) +
    theme_classic(base_size = 14) +
    labs(x = NULL, y = "Mean ± SE") +
    theme(
      legend.position = "none",
      strip.text = element_text(face = "bold", size = 12)
    )

  trait_fig

  # Save
  ggsave("trait_means_figure.jpeg",
         plot = trait_fig,
         width = 14, height = 4, dpi = 600)
```


```{r}
library(tidyverse)

# Create dataframe
trait_means <- tibble(
  Pop = rep(c("AC","BB","BO","HR"), each = 5),
  trait = rep(c("corolla diameter","SLA","δ13C","mean seed mass","fecundity"), times = 4),
  mean = c(
    20.55,17.72,-31.29,1.90,25.47,
    22.89,17.17,-29.64,3.92,3.98,
    23.50,15.52,-28.31,3.77,5.30,
    21.94,17.29,-30.77,2.72,17.73
  ),
  se = c(
    0.17,0.34,0.05,0.04, 1.91,
    0.28,0.57,0.08,0.23, 0.95,
    0.22,0.34,0.08,0.15, 0.80,
    0.14,0.31,0.04,0.05, 1.23
  )
)

# Flip ONLY δ13C
trait_means <- trait_means %>%
  mutate(mean_plot = ifelse(trait == "δ13C", -mean, mean),
         se_plot   = se)

trait_means$trait <- factor(trait_means$trait,
  levels = c("corolla diameter","SLA","δ13C","mean seed mass","fecundity"))

# Plot
trait_fig <- ggplot(trait_means, aes(Pop, mean_plot, fill = Pop)) +
  geom_col(width = 0.7) +
  geom_errorbar(aes(ymin = mean_plot - se_plot, ymax = mean_plot + se_plot),
                width = 0.2, na.rm = TRUE) +
  facet_wrap(~trait, scales = "free_y", nrow = 1) +   # ← all in one line
  theme_classic(base_size = 14) +
  labs(x = NULL, y = "Mean ± SE") +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 12)
  )

trait_fig

# Save
ggsave("trait_means_figure.jpeg",
       plot = trait_fig,
       width = 14, height = 4, dpi = 600)
```

#**Selection Gradients**
```{r}
#FECUNDITY = FITNESS
#should I include transect in the model?

dat_AC <- dat_AC %>%
  mutate(
    w_rel = est_fecundity / mean(est_fecundity, na.rm = TRUE)
  )

dat_BB <- dat_BB %>%
  mutate(
    w_rel = est_fecundity / mean(est_fecundity, na.rm = TRUE)
  )

dat_BO <- dat_BO %>%
  mutate(
    w_rel = est_fecundity / mean(est_fecundity, na.rm = TRUE)
  )

dat_HR <- dat_HR %>%
  mutate(
    w_rel = est_fecundity / mean(est_fecundity, na.rm = TRUE)
  )

mod1 <- lm(w_rel ~ corolla_z + SLA_z + d13C_z + msm_z, data = dat_AC)

mod2 <- lm(w_rel ~ corolla_z + SLA_z + d13C_z + msm_z + Transect, data = dat_AC)

AIC(mod1, mod2)

anova(mod1, mod2)
#not necessary for AC


mod1 <- lm(w_rel ~ corolla_z + SLA_z + d13C_z + msm_z, data = dat_BB)

mod2 <- lm(w_rel ~ corolla_z + SLA_z + d13C_z + msm_z + Location, data = dat_BB)

AIC(mod1, mod2)

anova(mod1, mod2)
#not necessary for BB, but its close..

mod1 <- lm(w_rel ~ corolla_z + SLA_z + d13C_z + msm_z, data = dat_BO)

mod2 <- lm(w_rel ~ corolla_z + SLA_z + d13C_z + msm_z + Transect, data = dat_BO)

AIC(mod1, mod2)

anova(mod1, mod2)
#not necessary

mod1 <- lm(w_rel ~ corolla_z + SLA_z + d13C_z + msm_z, data = dat_HR)

mod2 <- lm(w_rel ~ corolla_z + SLA_z + d13C_z + msm_z + Transect, data = dat_HR)

AIC(mod1, mod2)

anova(mod1, mod2)
#not necessary for HR, but its close..

```

```{r}
get_selection_gradients <- function(dat, pop_name) {
  d <- dat %>%
    select(corolla_z, SLA_z, d13C_z, msm_z, w_rel) %>%
    drop_na()

  mod <- lm(w_rel ~ corolla_z + SLA_z + d13C_z + msm_z, data = d)

  broom::tidy(mod, conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    mutate(Pop = pop_name, .before = 1)
}

library(broom)

beta_all <- bind_rows(
  get_selection_gradients(dat_AC, "AC"),
  get_selection_gradients(dat_BB, "BB"),
  get_selection_gradients(dat_BO, "BO"),
  get_selection_gradients(dat_HR, "HR")
)

beta_all
```


```{r}
#check model fit

d <- dat_AC %>%
  select(corolla_z, SLA_z, d13C_z, msm_z, w_rel) %>%
  drop_na()

mod <- lm(w_rel ~ corolla_z + SLA_z + d13C_z + msm_z, data = d)

library(car)
vif(mod) # < 3 = great! 

par(mfrow = c(2,2))
plot(mod)

summary(mod)$r.squared
```

```{r}
get_selection_gradients <- function(dat, pop_name) {
  d <- dat %>%
    select(corolla_z, SLA_z, d13C_z, msm_z, w_rel, Transect) %>%
    drop_na()

  mod <- lm(w_rel ~ corolla_z + SLA_z + d13C_z + msm_z + Transect, data = d)

  broom::tidy(mod, conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    mutate(Pop = pop_name, .before = 1)
}

library(broom)

beta_all <- bind_rows(
  get_selection_gradients(dat_AC, "AC"),
  get_selection_gradients(dat_BO, "BO"),
  get_selection_gradients(dat_HR, "HR")
)

beta_all
```

```{r}
get_selection_gradients <- function(dat, pop_name) {
  d <- dat %>%
    select(corolla_z, SLA_z, d13C_z, msm_z, w_rel, Location) %>%
    drop_na()

  mod <- lm(w_rel ~ corolla_z + SLA_z + d13C_z + msm_z + Location, data = d)

  broom::tidy(mod, conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    mutate(Pop = pop_name, .before = 1)
}

library(broom)

beta_all <- bind_rows(
  get_selection_gradients(dat_BB, "BB")
)

beta_all
```

```{r}
#SURVIVAL?
##Do segment-level/family-level trait values among flowering plants predict survival of that segment/family?

dat_AC <- dat_AC %>%
  mutate(seeds_sown = 10)

surv_model_AC <- glm(
  cbind(total_surv_to_flower, seeds_sown - total_surv_to_flower) ~
    corolla_z + SLA_z + d13C_z + msm_z,
  family = binomial,
  data = dat_AC
)
summary(surv_model_AC)

dat_BB <- dat_BB %>%
  mutate(seeds_sown = 8)

surv_model_BB <- glm(
  cbind(total_surv_to_flower, seeds_sown - total_surv_to_flower) ~
    corolla_z + SLA_z + d13C_z + msm_z,
  family = binomial,
  data = dat_BB
)
summary(surv_model_BB)

dat_BO <- dat_BO %>%
  mutate(seeds_sown = 8)

surv_model_BO <- glm(
  cbind(total_surv_to_flower, seeds_sown - total_surv_to_flower) ~
    corolla_z + SLA_z + d13C_z + msm_z,
  family = quasibinomial, #because of overdispersion
  data = dat_BO
)
summary(surv_model_BO)

dat_HR <- dat_HR %>%
  mutate(seeds_sown = 8)

surv_model_HR <- glm(
  cbind(total_surv_to_flower, seeds_sown - total_surv_to_flower) ~
    corolla_z + SLA_z + d13C_z + msm_z,
  family = quasibinomial, #because of overdispersion
  data = dat_HR
)
summary(surv_model_HR)

```


```{r}
#test the models

overdisp_fun <- function(model) {
  rdf <- df.residual(model)
  rp  <- residuals(model, type = "pearson")
  sum(rp^2) / rdf
}

overdisp_fun(surv_model_AC) #0.9597097 > good
overdisp_fun(surv_model_BB) #1.230493 > maybe a little overdispersed, but ok
overdisp_fun(surv_model_BO) #1.74148 > overdispersed
overdisp_fun(surv_model_HR) #1.747229 > overdispersed


library(car)
crPlots(surv_model_AC)
crPlots(surv_model_BB)
crPlots(surv_model_BO)
crPlots(surv_model_HR)

```

```{r}
library(ggplot2)
library(patchwork)

big_theme <- theme_minimal(base_size = 22) +  # increase base font
  theme(
    plot.title = element_text(size = 26, face = "bold"),
    axis.title = element_text(size = 22),
    axis.text = element_text(size = 18),
    legend.title = element_text(size = 20),
    legend.text = element_text(size = 18)
  )

set.seed(123)

n <- 200

data <- data.frame(
  trait1 = rnorm(n, 0, 1),
  trait2 = rnorm(n, 0, 1),
  trait3 = rnorm(n, 0, 1)
)

# Create fitness with a true signal + noise (phenotypic selection)
data$fitness <- 0.5 * data$trait1 - 0.3 * data$trait2 + rnorm(n, 0, 1)

# Panel A: Phenotypic selection
p1 <- ggplot(data, aes(trait1, fitness)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "black") +
  labs(title = "A. Phenotypic selection (β)",
       x = "Trait value", y = "Fitness (W)") +
  big_theme


p1

# Fake additive genetic variance-covariance matrix (G)

G <- matrix(c(

  1.0,  0.4, -0.2,

  0.4,  0.8,  0.3,

 -0.2,  0.3,  0.6

), nrow = 3, byrow = TRUE)

rownames(G) <- colnames(G) <- c("Trait1", "Trait2", "Trait3")

G

library(tidyverse)

rA_data <- as.data.frame(as.table(cov2cor(G))) %>%
  rename(trait1 = Var1, trait2 = Var2, rA = Freq)

rA_data

# Panel B: rA heatmap
p2 <- ggplot(rA_data, aes(trait1, trait2, fill = rA)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(rA, 2)), size = 6) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(title = "B. Genetic correlations (rA)",
       x = NULL,  # remove x-axis title
       y = NULL) +  # remove y-axis title
  big_theme

p2

eig <- eigen(G)

eigen_data <- data.frame(
  PC = paste0("PC", 1:length(eig$values)),
  eigenvalue = eig$values
)

eigen_data

library(tidyverse)

rA_data <- as.data.frame(as.table(cov2cor(G))) %>%
  rename(trait1 = Var1, trait2 = Var2, rA = Freq)

rA_data

# Panel C: eigenvalues
p3 <- ggplot(eigen_data, aes(x = PC, y = eigenvalue)) +
  geom_col() +
  labs(title = "C. Eigenvalues of G",
       x = "Principal components", y = "Variance") +
  big_theme

p3

p1 + p2 + p3


ggsave("theoretical_plot.png",
       plot = (p1 + p2 + p3),
       width = 20, height = 8, dpi = 700)
```


Evolvability figure
```{r}
library(tidyverse)
library(patchwork)

evol_data <- tribble(
  ~Population, ~e_mean_mean, ~e_mean_l95, ~e_mean_u95, ~c_mean_mean, ~c_mean_l95, ~c_mean_u95, ~a_mean_mean, ~a_mean_l95, ~a_mean_u95,
  "AC", 0.487608,  0.391209,  0.589368,  0.4087777, 0.3246184, 0.4938654, 0.8423334, 0.7490436, 0.9264143,
  "BB", 0.5022489, 0.380663,  0.6146454, 0.4057414, 0.3193795, 0.4935332, 0.81479,   0.7187795, 0.9086239,
  "BO", 0.4648472, 0.3693606, 0.5734179, 0.3929468, 0.3088527, 0.4900554, 0.8490192, 0.7461625, 0.9287693,
  "HR", 0.4374148, 0.3463179, 0.5267968, 0.350562,  0.2855698, 0.4253106, 0.8134198, 0.7240783, 0.9072107
) %>%
  mutate(Population = factor(Population, levels = c("AC", "BB", "BO", "HR")))

make_panel <- function(data, mean_col, l95_col, u95_col, ylab, label) {
  ggplot(data, aes(x = Population, y = .data[[mean_col]])) +
    geom_errorbar(
      aes(ymin = .data[[l95_col]], ymax = .data[[u95_col]]),
      width = 0,
      linewidth = 0.5
    ) +
    geom_point(size = 3) +
    labs(x = "Population", y = ylab) +
    annotate("text", x = 0.9, y = 0.99, label = label, size = 6) +
    scale_y_continuous(limits = c(0, 1)) +
    theme_classic() +
    theme(
      axis.text = element_text(size = 20),
      axis.title = element_text(size = 20)
    )
}

p1 <- make_panel(
  evol_data,
  mean_col = "e_mean_mean",
  l95_col = "e_mean_l95",
  u95_col = "e_mean_u95",
  ylab = "Mean Evolvability",
  label = "A)"
)

p2 <- make_panel(
  evol_data,
  mean_col = "c_mean_mean",
  l95_col = "c_mean_l95",
  u95_col = "c_mean_u95",
  ylab = "Conditional Evolvability",
  label = "B)"
)

p3 <- make_panel(
  evol_data,
  mean_col = "a_mean_mean",
  l95_col = "a_mean_l95",
  u95_col = "a_mean_u95",
  ylab = "Autonomy",
  label = "C)"
)

p1 + p2 + p3 + plot_layout(nrow = 1)

ggsave("evolvability.jpg",
       plot = (p1 + p2 + p3),
       width = 15, height = 6, dpi = 700)
```

```{r}

# Create dataframe
selection_data <- tribble(
  ~Population, ~Trait, ~beta, ~lower, ~upper, ~pvalue,
  
  "AC", "Corolla diameter",  0.009, -0.175,  0.194, 0.919,
  "AC", "SLA",              -0.131, -0.314,  0.051, 0.158,
  "AC", "δ13C",              0.343,  0.152,  0.534, 0.001,
  "AC", "msm",              -0.235, -0.428, -0.041, 0.018,
  
  "BB", "Corolla diameter", -0.326, -1.559,  0.908, 0.594,
  "BB", "SLA",              -0.155, -1.839,  1.529, 0.852,
  "BB", "δ13C",              1.105,  0.167,  2.043, 0.022,
  "BB", "msm",              -0.780, -1.795,  0.235, 0.127,
  
  "BO", "Corolla diameter",  0.694, -0.338,  1.725, 0.183,
  "BO", "SLA",               0.549, -0.262,  1.360, 0.180,
  "BO", "δ13C",             -0.140, -1.121,  0.841, 0.776,
  "BO", "msm",               0.104, -0.804,  1.012, 0.820,
  
  "HR", "Corolla diameter",  0.087, -0.068,  0.242, 0.269,
  "HR", "SLA",               0.048, -0.147,  0.243, 0.628,
  "HR", "δ13C",              0.227,  0.055,  0.400, 0.010,
  "HR", "msm",              -0.240, -0.416, -0.064, 0.008
)

# Order traits
selection_data$Trait <- factor(
  selection_data$Trait,
  levels = c("Corolla diameter", "SLA", "δ13C", "msm")
)

# Forest plot
ggplot(selection_data,
       aes(x = beta,
           y = Population)) +
  
  geom_vline(xintercept = 0,
             linetype = "dashed",
             color = "gray40") +
  
  geom_errorbarh(aes(xmin = lower,
                     xmax = upper),
                 height = 0.2,
                 linewidth = 0.7) +
  
  geom_point(aes(shape = pvalue < 0.05),
             size = 4) +
  
  facet_wrap(~Trait,
             scales = "free_x",
             nrow = 1) +
  
  scale_shape_manual(
    values = c(1, 16),
    labels = c("NS", "p < 0.05"),
    name = NULL
  ) +
  
  labs(
    x = expression(Standardized~directional~selection~gradient~(beta)),
    y = "Population"
  ) +
  guides(

  shape = guide_legend(override.aes = list(size = 5))

) +

theme_classic() +

theme(

  axis.text = element_text(size = 12),

  axis.title = element_text(size = 14),

  strip.text = element_text(size = 13, face = "bold"),

  legend.position = "bottom",

  legend.text = element_text(size = 14)

)

ggsave(
  "selection_gradient.jpeg",
  width = 12,
  height = 5,
  dpi = 700
)

```


```{r}
options(rgl.useNULL = TRUE)
library(evolqg)


# For two G matrices
beta <- matrix(rnorm(5 * 1000), nrow = 5)

responses_G1 <- AC_G_mean %*% beta
responses_G2 <- BB_G_mean %*% beta
responses_G3 <- BO_G_mean %*% beta
responses_G4 <- HR_G_mean %*% beta

RandomSkewers(AC_G_mean, BB_G_mean)
RandomSkewers(AC_G_mean, BO_G_mean)
RandomSkewers(AC_G_mean, HR_G_mean)
RandomSkewers(BB_G_mean, HR_G_mean)
RandomSkewers(BB_G_mean, BO_G_mean)
RandomSkewers(BO_G_mean, HR_G_mean)


```
